---
title: "Machine Learning Pipeline for Imaging Mass Cytometry Data Processing and Automated
  Single-Cell Annotation in Spatial Proteomics"
author: "Yuliia Babkova"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

link to presentation

https://docs.google.com/presentation/d/1j092IUsKge5XvswE18HGZPmz4NGRUtOXNunP9SOr3i0/edit?usp=sharing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
fig_dir <- "C:/Users/user/Desktop/Panel2_full/Figures"

if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}
```


```{r shell}
# steinbock preprocess imc images --hpf 50
# steinbock segment deepcell --minmax
# steinbock measure intensities
# steinbock measure regionprops
# steinbock measure neighbors --type centroids --dmax 15
```


```{r load-libraries, message=FALSE}
library(SingleCellExperiment)

library(caret)
library(xgboost)
library(randomForest)

library(imcRtools)
library(cytomapper)
library(CATALYST)

library(scater)
library(scuttle)
library(scran)

library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(purrr)

library(ggplot2)
library(ggrepel)
library(viridis)
library(patchwork)
library(cowplot)
library(ggridges)

library(BiocParallel)
library(doParallel)
library(BiocNeighbors)

library(ComplexHeatmap)
library(circlize)

library(FNN)
library(igraph)

library(dittoSeq)
```

```{r read-spe}
base_path <- "C:/Users/user/Desktop/Panel2_full"

raw_exprs_cofactor <- 5

out_dir <- file.path(base_path, "ROut")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

spe <- read_steinbock(base_path, return_as = "spe")
```

# spillover

```{r define-readSCEfromTIFF}
readSCEfromTIFF <- function(x, image_df_path, panel_df_path) {
  files <- list.files(x, pattern = "\\.tiff$", full.names = TRUE)
  stopifnot(length(files) > 0)

  panel <- read.csv(panel_df_path)

  img_list <- lapply(files, EBImage::readImage)

  arr_list <- lapply(img_list, function(img) {
    arr <- as.array(img)
    dim(arr) <- c(dim(arr)[1] * dim(arr)[2], dim(arr)[3])
    t(arr)
  })

  counts_mat <- do.call(cbind, arr_list)

  n_pix <- vapply(arr_list, ncol, integer(1))
  sample_id <- rep(tools::file_path_sans_ext(basename(files)), times = n_pix)

  colnames(counts_mat) <- paste0(sample_id, "_", sequence(n_pix))
  rownames(counts_mat) <- paste0(panel$channel, "Di")

  sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts_mat)
  )

  colData(sce)$sample_id <- sample_id

  masses <- as.numeric(sub(".*?([0-9]+)$", "\\1", panel$channel))
  file_index <- rep(seq_along(files), times = n_pix)

  colData(sce)$sample_mass <- masses[file_index]

  sce
}

```


```{r}
readSCEfromTIFF <- function(x, image_df_path, panel_df_path) {
  files <- list.files(x, pattern = "\\.tiff$", full.names = TRUE)
  stopifnot(length(files) > 0)

  panel <- read.csv(panel_df_path)
  images <- read.csv(image_df_path)

  img_list <- lapply(files, EBImage::readImage)

  arr_list <- lapply(img_list, function(img) {
    arr <- as.array(img)
    dim(arr) <- c(dim(arr)[1] * dim(arr)[2], dim(arr)[3])
    t(arr)
  })

  counts_mat <- do.call(cbind, arr_list)

  n_pix <- vapply(arr_list, ncol, integer(1))
  sample_id <- rep(tools::file_path_sans_ext(basename(files)), times = n_pix)

  colnames(counts_mat) <- paste0(sample_id, "_", sequence(n_pix))
  rownames(counts_mat) <- paste0(panel$channel, "Di")
  rownames(counts_mat)[rownames(counts_mat) == "ArAr80Di"] <- "Ar80Di"

  sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts_mat)
  )

  colData(sce)$sample_id <- sample_id

  file_names <- basename(files)
  image_names <- images$image

  image_row <- match(file_names, image_names)

  mass_text <- paste(
    images$acquisition_description[image_row],
    images$source_file[image_row],
    images$recovery_file[image_row]
  )

  sample_mass_file <- as.numeric(sub(".*?([0-9]{2,3}).*", "\\1", mass_text))

  if (any(is.na(sample_mass_file))) {
    stop("Some sample_mass values could not be extracted from images.csv")
  }

  colData(sce)$sample_mass <- rep(sample_mass_file, times = n_pix)

  rowData(sce)$channel_name <- rownames(sce)
  rowData(sce)$marker_name <- rownames(sce)

  sce
}
```


```{r read-compensated-tiff}
comp_path <- list.files(base_path, pattern = "compensation$", full.names = TRUE)

img_path <- file.path(comp_path, "img")
image_csv <- file.path(comp_path, "images.csv")
panel_csv <- file.path(comp_path, "panel.csv")

sceTIFF <- readSCEfromTIFF(
  x = img_path,
  image_df_path = image_csv,
  panel_df_path = panel_csv
)

sceTIFF$Panel <- basename(base_path)
sceTIFF
```

# comp assay

```{r process-compensated-sce}
compensation_pixel_exprs_cofactor <- 5
compensated_exprs_cofactor <- 1

process_comp_sce <- function(x, minevents = 40) {
  assay(x, "exprs") <- asinh(counts(x) / compensation_pixel_exprs_cofactor)

  rownames(x)[rownames(x) == "ArAr80Di"] <- "Ar80Di"

  rowData(x)$channel_name <- rownames(x)
  rowData(x)$marker_name <- rownames(x)

  bc_key <- as.numeric(unique(x$sample_mass))
  bc_key <- bc_key[order(bc_key)]

  channel_mass <- as.numeric(gsub("[^0-9]", "", rowData(x)$channel_name))
  x <- x[channel_mass %in% bc_key, ]

  rowData(x)$channel_name <- rownames(x)
  rowData(x)$marker_name <- rownames(x)

  print(bc_key)
  print(rownames(x))

  x <- assignPrelim(x, bc_key = bc_key)
  x <- estCutoffs(x)
  x <- applyCutoffs(x)

  x <- filterPixels(
    x,
    minevents = minevents,
    correct_pixels = TRUE
  )

  x <- computeSpillmat(x)

  x
}

sceTIFF <- process_comp_sce(sceTIFF)
sceTIFF
```

```{r fix-spillover-matrix-names}
sm <- metadata(sceTIFF)$spillover_matrix

colnames(sm) <- rownames(sm)

metadata(sceTIFF)$spillover_matrix <- sm

all(rownames(sm) %in% colnames(sm))
setdiff(rownames(sm), colnames(sm))
setdiff(colnames(sm), rownames(sm))
```

```{r spillover-plot, fig.width=11, fig.height=11}

par(
  mar = c(18, 18, 4, 4),
  cex.axis = 0.7,
  xpd = TRUE
)

plotSpillmat(
  sceTIFF,
  isotope_list = isotope_list
)
```


```{r}
sm <- metadata(sceTIFF)$spillover_matrix

# The spillover matrix must have identical row and column names.
# Do not manually rename the first column to Ar80Di here.
colnames(sm) <- rownames(sm)

# Check that the matrix is valid before compensation
stopifnot(all(rownames(sm) == colnames(sm)))

metadata(sceTIFF)$spillover_matrix <- sm

# write.csv(sm, file.path(out_dir, "sm.csv"))

rowData(spe)$channel_name <- paste0(rowData(spe)$channel, "Di")

spe <- compCytof(
  spe,
  sm,
  transform = TRUE,
  cofactor = compensated_exprs_cofactor,
  isotope_list = isotope_list,
  overwrite = FALSE
)

assay(spe, "counts") <- assay(spe, "compcounts")
assay(spe, "exprs") <- assay(spe, "compexprs")

assay(spe, "compcounts") <- NULL
assay(spe, "compexprs") <- NULL

# already with comp assay
spe

saveRDS(spe, file.path(base_path, "ROut", "spe.rds"))

```

## already gated (labelled) images

USL - the group of colorectal cancer (CRC) tissue samples/patients that came from University Saint-Luc (Brussels)

```{r}
spe <- readRDS(file.path(base_path, "ROut", "spe.rds"))

gates <- list.files(
  file.path(base_path, "gates"),
  pattern = "\\.rds$",
  full.names = TRUE,
  recursive = TRUE
)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)

  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    cytomapper_CellLabel = as.character(x$cytomapper_CellLabel),
    gate_file = basename(path),
    row.names = NULL
  )
})

gate_labels$cytomapper_CellLabel <- sub(
  "_[0-9]+[a-zA-Z]?$",
  "",
  gate_labels$cytomapper_CellLabel
)

gate_labels$cytomapper_CellLabel <- sub(
  " \\([0-9]+\\)$",
  "",
  gate_labels$cytomapper_CellLabel
)

gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "Cd4"] <- "CD4"
gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "MDSC_"] <- "MDSC"
gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "Cd8"] <- "CD8"


gate_labels$cell_id <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)

sum(gate_labels$cell_id %in% colnames(spe))
nrow(gate_labels)

```
## Quality Control of Cell Annotation Mapping

```{r}
length(unique(gate_labels$sample_id))
length(unique(spe$sample_id))

setdiff(unique(gate_labels$sample_id), unique(spe$sample_id))
setdiff(unique(spe$sample_id), unique(gate_labels$sample_id))
```
```{r}
length(unique(matched_gate_labels$sample_id))
nrow(matched_gate_labels)
table(matched_gate_labels$cytomapper_CellLabel)
```


# img, masks

```{r}
matched_gate_labels <- gate_labels[
  gate_labels$cell_id %in% colnames(spe),
]

images_28 <- sort(unique(matched_gate_labels$sample_id))

images_28_pattern <- paste0(
  "^",
  images_28,
  "[.]tiff$",
  collapse = "|"
)

images_28_obj <- loadImages(
  file.path(base_path, "img"),
  pattern = images_28_pattern,
  BPPARAM = BiocParallel::SerialParam()
)

images_28_obj <- images_28_obj[images_28]
channelNames(images_28_obj) <- rownames(spe)

masks_28 <- loadImages(
  file.path(base_path, "masks"),
  pattern = images_28_pattern,
  as.is = TRUE,
  BPPARAM = BiocParallel::SerialParam()
)

masks_28 <- masks_28[names(images_28_obj)]
```

# Img QC

```{r create-spe-28}
spe_28 <- spe[, spe$sample_id %in% images_28]

spe_28

```


```{r heatmap, fig.height=10, fig.width=10}
image_mean <- aggregateAcrossCells(spe_28, 
                                   ids = spe_28$sample_id, 
                                   statistics="mean",
                                   use.assay.type = "counts")

assay(image_mean, "exprs") <- asinh(assay(image_mean, "counts"))

genes_to_plot <- rownames(image_mean)[
  rowSums(assay(image_mean, "exprs") != 0, na.rm = TRUE) > 0
]
if (length(genes_to_plot) == 0) {
  stop("No selected markers have non-zero image-level expression values.")
}

dittoHeatmap(
  image_mean, 
  genes = genes_to_plot,
  assay = "exprs", 
  cluster_cols = FALSE, 
  scale = "none",
  heatmap.colors = colorRampPalette(c(
    "white",
    "#B8CAE2", "#9EB7D8", "#4575B4", "#325584", "#264164", "#182A40"
  ))(100), 
  show_colnames = TRUE
)

```


```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)

image_mean <- aggregateAcrossCells(
  spe_28,
  ids = spe_28$sample_id,
  statistics = "mean",
  use.assay.type = "counts"
)

assay(image_mean, "exprs") <- asinh(assay(image_mean, "counts"))

genes_to_plot <- rownames(image_mean)[
  rowSums(assay(image_mean, "exprs") != 0, na.rm = TRUE) > 0
]

mat <- assay(image_mean, "exprs")[genes_to_plot, ]

colnames(mat) <- sub(".*IMC-02_", "img_", colnames(mat))

col_fun <- colorRamp2(
  c(0, 0.5, 1, 2, 4, max(mat, na.rm = TRUE)),
  c("white", "#EAF2FB", "#B7CDE8", "#6FA8DC", "#1F4E79", "#08306B")
)

image_signal <- colSums(mat, na.rm = TRUE)

top_anno <- HeatmapAnnotation(
  Signal = anno_barplot(
    image_signal,
    border = FALSE,
    gp = gpar(fill = "#2F5597"),
    height = unit(1.5, "cm")
  )
)

Heatmap(
  mat,
  name = "Mean exprs",
  col = col_fun,
  top_annotation = top_anno,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "right",
  row_names_gp = gpar(fontsize = 5),
  column_names_gp = gpar(fontsize = 5),
  column_names_rot = 40,
  rect_gp = gpar(col = "grey90", lwd = 0.25)
)
```


```{r}
colSums(assay(image_mean, "exprs"))[
  colnames(image_mean) == "10123883-USL-VAR-TIS-01-IMC-02_005"
]
```

```{r}
summary(
  assay(image_mean, "exprs")[,
    "10123883-USL-VAR-TIS-01-IMC-02_005"
  ]
)
```

delete 1 blank image from analysis

```{r}
spe_27 <- spe_28[
  ,
  spe_28$sample_id != "10123883-USL-VAR-TIS-01-IMC-02_005"
]

images_27 <- setdiff(
  names(images_28_obj),
  "10123883-USL-VAR-TIS-01-IMC-02_005"
)

images_27_obj <- images_28_obj[images_27]
masks_27 <- masks_28[images_27]

gate_labels <- gate_labels[
  gate_labels$sample_id != "10123883-USL-VAR-TIS-01-IMC-02_005",
]
```


## gating via cymoviewer

```{r, message=FALSE}
image_ids <- names(images_27_obj)
mask_ids <- names(masks_27)
spe_ids <- unique(as.character(spe_27$sample_id))

stopifnot(
  identical(image_ids, mask_ids),
  setequal(image_ids, spe_ids),
  !anyDuplicated(image_ids)
)

cur_mcols <- S4Vectors::DataFrame(
  image = image_ids,
  sample_id = image_ids,
  name = image_ids,
  row.names = image_ids
)

mcols(images_27_obj) <- cur_mcols
mcols(masks_27)      <- cur_mcols


identical(mcols(images_27_obj), mcols(masks_27))
identical(mcols(images_27_obj)$sample_id, names(masks_27))
setequal(
  mcols(images_27_obj)$sample_id,
  unique(as.character(spe_27$sample_id))
)
```

```{r}
if (interactive()) {
  cytomapperShiny(object = spe_27, mask = masks_27, image = images_27_obj,
                  cell_id = "ObjectNumber", img_id = "sample_id")
}
```


## consolidating labels

## cells assigned to the same class multiple times

```{r}
cur_tab <- unclass(table( #make a mat no table class
  gate_labels$cell_id,
  gate_labels$cytomapper_CellLabel
))

cur_tab_raw <- cur_tab

same_label_duplicate_ids <- rownames(cur_tab_raw)[rowSums(cur_tab_raw > 1) > 0]

if (length(same_label_duplicate_ids) > 0) {
  cur_tab_raw[same_label_duplicate_ids, , drop = FALSE]
} else {
  print("No same-label duplicated gates found.")
}
```

##Find duplicates

```{r}
cur_tab <- unclass(table(
  gate_labels$cell_id,
  gate_labels$cytomapper_CellLabel
))

cur_tab_raw <- cur_tab

same_label_duplicate_ids <- rownames(cur_tab_raw)[
  rowSums(cur_tab_raw > 1) > 0
]
```

```{r}
if (length(same_label_duplicate_ids) > 0) {
  cur_tab_raw[same_label_duplicate_ids, , drop = FALSE]
}
```

```{r}
# Convert duplicated same-class labels to presence/absence
cur_tab <- (cur_tab_raw > 0) * 1
```


## same label assigment

```{r}
cur_tab <- cur_tab_raw
cur_tab[cur_tab > 1] <- 1

# True multi-label doublets
multi_label_doublet_ids <- rownames(cur_tab)[rowSums(cur_tab) > 1]

if (length(multi_label_doublet_ids) > 0) {
  head(cur_tab[multi_label_doublet_ids, , drop = FALSE],5)
} else {
  message("No multi-label doublets found.")
}

```

# Consolidate labels

```{r}
#assign all cell_ids as doublets
cur_labels <- rep("doublets", nrow(cur_tab))
names(cur_labels) <- rownames(cur_tab) #cell IDs as names,

single_index <- rowSums(cur_tab) == 1 #finds cells that have exactly one label

# cells with exactly 1 label, this finds which column = 1
cur_labels[single_index] <- colnames(cur_tab)[
  apply(cur_tab[single_index, , drop = FALSE], 1, which.max)
]

#cells with exactly 2 labels, one = Tumor
double_tumor_index <- rowSums(cur_tab) == 2 & cur_tab[, "Tumor"] == 1

#cells = Tumor + one other label, "-" the Tumor, keeps the other label
if (sum(double_tumor_index) > 0) {
  no_tumor <- cur_tab[, colnames(cur_tab) != "Tumor", drop = FALSE]

  cur_labels[double_tumor_index] <- colnames(no_tumor)[
    apply(no_tumor[double_tumor_index, , drop = FALSE], 1, which.max)
  ]
} else {
  message("No Tumor + other double labels found.")
}

#unresolved non-Tumor doublets
removed_doublet_ids <- names(cur_labels)[cur_labels == "doublets"]

#"-" anything that still calls doublets
if (length(removed_doublet_ids) > 0) {
  cur_labels <- cur_labels[cur_labels != "doublets"]
} else {
  message("No non-tumor doublets to remove.")
}

consolidation_stats <- data.frame(
  category = c(
    "total_unique_gated_cell_ids",
    "same_label_duplicates_collapsed",
    "single_label_kept",
    "tumor_plus_other_renamed_to_other",
    "non_tumor_doublets_removed"
  ),
  n = c(
    nrow(cur_tab),
    length(same_label_duplicate_ids),
    sum(single_index),
    sum(double_tumor_index),
    length(removed_doublet_ids)
  ),
  row.names = NULL
)

consolidation_stats
```
```{r}
consolidation_stats

table(cur_labels)

length(unique(cur_labels))
```

```{r}
table(cur_labels)
```


```{r}
sce_labels <- rep("unlabelled", ncol(spe_27))
names(sce_labels) <- colnames(spe_27)

matched_ids <- intersect(names(cur_labels), colnames(spe_27))
unmatched_ids <- setdiff(names(cur_labels), colnames(spe_27))

message("Matched gated cells: ", length(matched_ids))
message("Unmatched gated cells: ", length(unmatched_ids))

#replaces "unlabelled" with the final labels for matching gated cells
sce_labels[matched_ids] <- cur_labels[matched_ids]

#label vector still has exactly one entry per spe_27 column, in the same order
stopifnot(length(sce_labels) == ncol(spe_27))
stopifnot(identical(names(sce_labels), colnames(spe_27)))

# final labels in colData(sce)
spe_27$cell_labels <- sce_labels

table(spe_27$cell_labels)

```

## visualization

```{r}
cell_types <- c(
  B = "red3",
  BnT = "indianred4",
  CD4 = "palevioletred1",
  CD8 = "darkorchid",
  Treg = "maroon4",
  NK = "yellow",
  Igg = "darkred",

  DC = "green4",
  Mac = "darkolivegreen",
  MacCD209 = "olivedrab1",
  MacCD204 = "olivedrab3",
  mregDC = "green",

  Neutrophil = "blue1",
  PMN_MDSC = "deepskyblue",
  MDSC = "aquamarine4",

  Vasculature = "brown2",
  vCAF = "tomato",
  Tumor = "sienna4",
  SMA = "goldenrod2",
  Fibro = "chocolate1",

  unlabelled = "gray",
  undefined = "black"
)

metadata(spe_27)$colour_vectors$cell_types <- cell_types
```

## QC


```{r}
## save labelled SPE objects

data_out_path <- file.path(base_path, "ROut")
dir.create(data_out_path, showWarnings = FALSE, recursive = TRUE)

spe_all <- spe_27

lab_sce <- spe_27[
  ,
  !is.na(spe_27$cell_labels) &
    spe_27$cell_labels != "unlabelled"
]

saveRDS(
  spe_all,
  file.path(data_out_path, "spe_all_lab.rds")
)

saveRDS(
  lab_sce,
  file.path(data_out_path, "lab_spe.rds")
)

spe_all
lab_sce
```

```{r}
lab_sce <- spe_27[, !is.na(spe_27$cell_labels) & spe_27$cell_labels != "unlabelled"]

# Aggregate mean expression per sample and cell label
agr_sce <- scater::aggregateAcrossCells(
  lab_sce,
  ids = colData(lab_sce)[, c("sample_id", "cell_labels")],
  statistics = "mean"
)

assay(agr_sce, "exprs") <- asinh(counts(agr_sce))

# Clean column names
colnames(agr_sce) <- paste0(agr_sce$sample_id, "_", agr_sce$cell_labels)

# Markers used for visualization
cur_markers <- rownames(spe_27)[
  !grepl("DNA|Histone", rownames(spe_27), ignore.case = TRUE)
]

# Keep colors only for labels present in aggregated object
cell_type_colors <- metadata(spe_27)$colour_vectors$cell_types
cell_type_colors <- cell_type_colors[
  names(cell_type_colors) %in% unique(agr_sce$cell_labels)
]
```


```{r define-color-vector, fig.height=12, fig.width=10}
dittoHeatmap(
  agr_sce[cur_markers, ],
  assay = "exprs",
  annot.by = c("sample_id", "cell_labels"),
  cluster_rows = FALSE,
  scale = "none",
  heatmap.colors = viridis::viridis(100),
  order.by = "cell_labels",
  annotation_colors = list(
    cell_labels = cell_type_colors
  )
)
```

# train classifier

## classifier rules

```{r}
lab_sce <- spe_27[, !is.na(spe_27$cell_labels) & spe_27$cell_labels != "unlabelled"]
unlab_sce <- spe_27[, is.na(spe_27$cell_labels) | spe_27$cell_labels == "unlabelled"]

# - Igg < 1
exprs_mat <- assay(lab_sce, "exprs")
low_igg <- lab_sce$cell_labels == "Igg" & exprs_mat["IgG", ] < 1
lab_sce <- lab_sce[, !low_igg]

# - NKs withCD7 expression < 0.5
exprs_mat <- assay(lab_sce, "exprs")
low_cd7_nk <- lab_sce$cell_labels == "NK" & exprs_mat["CD7", ] < 0.5
lab_sce <- lab_sce[, !low_cd7_nk]

# Relabel mregDC cells with high epithelial/tumor marker expression to Tumor
exprs_mat <- assay(lab_sce, "exprs")
mregdc_to_tumor <- lab_sce$cell_labels == "mregDC" &
  (
    exprs_mat["panCK", ] > 0.5 |
      exprs_mat["Ecad", ] > 0.8
  )

lab_sce$cell_labels[mregdc_to_tumor] <- "Tumor"

table(lab_sce$cell_labels)
```
## save lab_sce

```{r}
dat <- as.data.frame(unclass(table(lab_sce$sample_id,lab_sce$cell_labels)))
dat
```


```{r}
table(
  label_status = ifelse(
    spe_27$cell_labels == "unlabelled",
    "unlabelled",
    "labelled"
  )
)
```

## find a good test/train split

## split data into test & train sets

```{r}
# Stratified train/test row split for multi-label count data.
#
# Each row is a sample with non-negative counts of each cell type in the
# columns. The split is at the row level, but it targets *cell-level* totals:
# achieved_test_frac[l] := test_count[l] / total[l] should be close to
# `default_test_frac` for common labels and `rare_test_frac` for the labels
# in `rare_labels` (default: "MDSC"). Additional labels whose total count is
# below `rare_threshold` are treated as rare too, if that argument is given.
#
# Objective (minimized):
#   sum_l  w_l * (test_count_l / total_l - target_frac_l)^2
# with w_l = `rare_weight` for rare labels, 1 otherwise.
#
# Algorithm: random init (biased toward test for rows containing rare labels),
# then steepest-descent single-row flips evaluated for all rows in one
# vectorized matrix multiply per step. Repeated from `n_restarts` initial
# seeds; the best basin wins. Converges in milliseconds for a few hundred
# rows; scales as O(restarts * flips * n * L).

split_test_rows <- function(counts,
                            default_test_frac = 0.20,
                            rare_test_frac    = 0.40,
                            rare_labels       = c("MDSC"),
                            rare_threshold    = NULL,
                            rare_weight       = 1,
                            n_restarts        = 100,
                            seed              = 42) {
  set.seed(seed)

  if (is.data.frame(counts)) {
    if (!is.numeric(counts[[1]])) {
      rn     <- as.character(counts[[1]])
      counts <- as.matrix(counts[, -1, drop = FALSE])
      rownames(counts) <- rn
    } else {
      counts <- as.matrix(counts)
    }
  }
  storage.mode(counts) <- "double"

  labels <- colnames(counts)
  totals <- colSums(counts)
  n      <- nrow(counts)

  if (!is.null(rare_threshold))
    rare_labels <- union(rare_labels, labels[totals < rare_threshold])
  rare_labels <- intersect(rare_labels, labels)
  is_rare     <- labels %in% rare_labels

  target_frac        <- ifelse(is_rare, rare_test_frac, default_test_frac)
  names(target_frac) <- labels

  # Zero out empty columns so they don't contribute a constant target_frac^2
  # penalty that no flip can fix.
  active     <- totals > 0
  w          <- ifelse(is_rare, rare_weight, 1) * active
  inv_totals <- ifelse(active, 1 / pmax(totals, 1), 0)

  # rate[i,l] = counts[i,l] / total[l]; precomputed once.
  rate      <- counts * rep(inv_totals, each = n)
  rate_sq_w <- as.numeric(rate^2 %*% w)   # length n, constant across iterations

  # Bias rows containing rare-label cells toward starting in test.
  has_rare  <- if (any(is_rare)) rowSums(counts[, is_rare, drop = FALSE]) > 0
               else rep(FALSE, n)
  init_prob <- ifelse(has_rare, rare_test_frac, default_test_frac)

  best_score   <- Inf
  best_in_test <- logical(n)

  for (restart in seq_len(n_restarts)) {
    in_test  <- runif(n) < init_prob
    test_sum <- colSums(counts[in_test, , drop = FALSE])

    repeat {
      cur_err   <- test_sum * inv_totals - target_frac
      cur_score <- sum(w * cur_err * cur_err)

      sgn        <- 1 - 2 * in_test                            # +1 = will add to test, -1 = will leave test
      linear     <- as.numeric(rate %*% (w * cur_err))
      flip_delta <- 2 * sgn * linear + rate_sq_w               # change in objective for each candidate flip

      best_i <- which.min(flip_delta)
      if (flip_delta[best_i] >= -1e-12) break                  # local minimum reached

      test_sum        <- test_sum + sgn[best_i] * counts[best_i, ]
      in_test[best_i] <- !in_test[best_i]
    }

    if (cur_score < best_score) {
      best_score   <- cur_score
      best_in_test <- in_test
    }
  }

  test_sum <- colSums(counts[best_in_test, , drop = FALSE])
  achieved <- ifelse(active, test_sum * inv_totals, NA_real_)

  summary_df <- data.frame(
    label              = labels,
    total              = totals,
    test_count         = test_sum,
    train_count        = totals - test_sum,
    target_test_frac   = target_frac,
    achieved_test_frac = round(achieved, 3),
    is_rare            = is_rare,
    row.names          = NULL
  )

  list(
    test_idx       = which(best_in_test),
    train_idx      = which(!best_in_test),
    test_rownames  = rownames(counts)[best_in_test],
    train_rownames = rownames(counts)[!best_in_test],
    summary        = summary_df,
    score          = best_score
  )
}
```


## looking for a seed with the best distribution between train/test

```{r}
res <- split_test_rows(
  dat,
  default_test_frac = 0.30,
  rare_test_frac = 0.30,
  rare_labels = c("B", "BnT", "Igg", "mregDC", "NK"),
  rare_weight = 10,
  n_restarts = 3000,
  seed = 42
)

res$summary
```

```{r train-test-split-summary}
test_sce <- lab_sce[, lab_sce$sample_id %in% res$test_rownames]
train_sce <- lab_sce[, lab_sce$sample_id %in% res$train_rownames]

all_labels <- sort(unique(c(
  as.character(train_sce$cell_labels),
  as.character(test_sce$cell_labels)
)))

split_stats <- data.frame(
  label = all_labels,
  train = as.integer(table(factor(train_sce$cell_labels, levels = all_labels))),
  test = as.integer(table(factor(test_sce$cell_labels, levels = all_labels)))
) |>
  dplyr::mutate(test_frac = round(test / (train + test), 3))

split_stats
mean(split_stats$test_frac, na.rm = TRUE)

```
```{r}
summary(split_stats$test_frac)

split_stats |>
  dplyr::arrange(test_frac)
```


```{r}
test_sce <- lab_sce[, lab_sce$sample_id %in% res$test_rownames]
train_sce <- lab_sce[, lab_sce$sample_id %in% res$train_rownames]
```

## train/test split summary

```{r}
all_labels <- sort(unique(c(
  as.character(train_sce$cell_labels),
  as.character(test_sce$cell_labels)
)))

all_labels <- all_labels[
  !is.na(all_labels) &
    all_labels != "unlabelled"
]

train_tab <- table(
  factor(as.character(train_sce$cell_labels), levels = all_labels)
)

test_tab <- table(
  factor(as.character(test_sce$cell_labels), levels = all_labels)
)

split_stats <- data.frame(
  cell_label = all_labels,
  train_n = as.integer(train_tab),
  test_n = as.integer(test_tab),
  row.names = NULL
)

split_stats
```

## rare cell types distribition between samples

```{r}
rare_celltypes <- c("BnT", "Igg", "MDSC", "mregDC", "vCAF")

rare_balance <- bind_rows(
  as.data.frame(colData(train_sce)) |> 
    mutate(split = "train"),
  as.data.frame(colData(test_sce)) |> 
    mutate(split = "test")
) |> 
  dplyr::filter(cell_labels %in% rare_celltypes) |> 
  group_by(split, cell_labels) |> 
  summarise(
    total_cells = n(),
    n_samples = n_distinct(sample_id),
    .groups = "drop"
  ) |> 
  arrange(split, cell_labels)

rare_balance
```
## Train-Test Split Quality Assessment

```{r}
mean(res$summary$achieved_test_frac)
summary(res$summary$achieved_test_frac)
```


```{r}
res$summary |>
  dplyr::select(
    label,
    train_count,
    test_count,
    achieved_test_frac
  ) |>
  dplyr::arrange(desc(test_count))
```
```{r}
res$summary |>
  dplyr::filter(train_count < 50 | test_count < 20)
```



## prepare data for classifier

1. split the labelled data into training + test data at the image level by `sample_id`. 
All cells from the same image are assigned training or test set

The split was generated with 5 grouped folds using `groupKFold`, 
and `Fold5` was used as the held-out test set. 

Within the training set another 5-fold grouped cross-validation for XGBoost parameter tuning was applied

```{r prepare-data, message = FALSE, eval=TRUE}

# Create unique cell identifiers so each cell can be traced back to its image
# and object number after training, prediction, or error analysis.
train_sce$cell_id <- paste0(train_sce$sample_id, "_", train_sce$ObjectNumber)
test_sce$cell_id <- paste0(test_sce$sample_id, "_", test_sce$ObjectNumber)

colnames(train_sce) <- train_sce$cell_id
colnames(test_sce) <- test_sce$cell_id

marker_keep <- !grepl("DNA|Histone", rownames(train_sce))

# XGBoost expects cells as rows and features as columns.
# The assay matrix has markers as rows and cells as columns, so we transpose it.
X <- t(assay(train_sce, "exprs")[marker_keep, ])
X_test_all <- t(assay(test_sce, "exprs")[marker_keep, ])

stopifnot(identical(colnames(X), colnames(X_test_all)))
stopifnot(!anyDuplicated(colnames(X)))

# XGBoost requires numeric class labels starting at 0.
# The factor levels from the training set define the fixed mapping between
# numeric class IDs and biological cell labels.
y_true <- factor(train_sce$cell_labels)
label_mapping <- levels(y_true)

# Convert training labels from factors to 0-based numeric labels.
y <- as.integer(y_true) - 1

# Number of classes required by the multiclass XGBoost objective.
num_classes <- length(label_mapping)

stopifnot(!anyNA(y))
stopifnot(min(y) == 0)
stopifnot(max(y) == num_classes - 1)

# Keep only labelled test cells for evaluation.
# Unlabelled cells are excluded from Xy_test because XGBoost cannot accept NA labels.
test_eval_keep <- !is.na(test_sce$cell_labels) &
  test_sce$cell_labels != "unlabelled" &
  test_sce$cell_labels %in% label_mapping

X_test_eval <- t(assay(test_sce, "exprs")[marker_keep, test_eval_keep])

y_test_true <- factor(
  test_sce$cell_labels[test_eval_keep],
  levels = label_mapping
)

y_test <- as.integer(y_test_true) - 1

stopifnot(!anyNA(y_test))
stopifnot(min(y_test) >= 0)
stopifnot(max(y_test) <= num_classes - 1)

# Inverse-frequency class weights so rare cell types contribute more
# to the loss function during training.
w_class <- 1 / (table(y_true) / length(y_true))

# Down-weight undefined and IgG classes so they do not dominate the classifier
# despite being noisy or less biologically informative labels.
if ("undefined" %in% names(w_class)) {
  w_class["undefined"] <- w_class["undefined"] / 10
}

if ("Igg" %in% names(w_class)) {
  w_class["Igg"] <- w_class["Igg"] / 30
}

# Convert class-level weights into one weight per training cell.
# Rescale weights to mean 1 for numerical stability.
w <- as.numeric(w_class[y_true])
w <- w / mean(w)

# Training data: features, labels, and cell weights.
Xy <- xgboost::xgb.DMatrix(
  data = X,
  label = y,
  weight = w
)

# Labelled held-out test data, for evaluation only.
Xy_test <- xgboost::xgb.DMatrix(
  data = X_test_eval,
  label = y_test
)

# All held-out/test cells, including unlabelled cells, for prediction/classification.
Xy_test_all <- xgboost::xgb.DMatrix(
  data = X_test_all
)

test_eval_cell_ids <- colnames(test_sce)[test_eval_keep]
test_all_cell_ids <- colnames(test_sce)

table(y_test_true)
```

## check the rules

Here it is clearly seen the upward rules for the specific markers work

# TEMP / QC plots
```{r}
coldat <- as.data.frame(t(assay(train_sce, "exprs")))
coldat$label <- train_sce$cell_labels

ggplot(coldat |> dplyr::filter(label == "Igg"),
       aes(x = IgG, y = DNA1, col = label)) +
  geom_point() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  theme_bw()

ggplot(coldat |> dplyr::filter(label == "mregDC"),
       aes(x = Ecad, y = panCK, col = label)) +
  geom_point() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0.8, linetype = "dashed") +
  theme_bw()

```

## params for classofier, cross validation

All heavy calculations are here

```{r train-model-cv, message = FALSE, eval = TRUE}

# Make sure cells from the same image are not split across CV train/test folds.
# groupKFold returns training indices, while xgb.cv expects validation indices,
# so we convert each fold to the held-out indices.
cv_train_index <- caret::groupKFold(
  group = factor(train_sce$sample_id),
  k = 5
)

all_indices <- seq_len(ncol(train_sce))

fold_list <- lapply(
  cv_train_index,
  function(train_idx) setdiff(all_indices, train_idx)
)

# XGBoost builds an ensemble of decision trees sequentially
# Each new tree tries to correct the errors made by the previous trees

param_grid <- expand.grid( #creates all combinations of the parameter values 3 * 2 * 1 * 1 * 2 = 12 models
  
  # Maximum depth of each tree.
  # Smaller values make simpler trees and reduce overfitting
  # Larger values allow more complex interactions between markers
  max_depth = c(3, 4, 5),

  # Learning rate controls how much each new tree contributes to the final model
  # < eta usually needs more trees but can generalize better
  eta = c(0.05, 0.1),

  # Fraction of cells sampled for each tree
  # Subsampling reduces overfitting by making trees less correlated
  subsample = c(0.8), #for each tree, XGBoost randomly uses 80% of the training cells and leaves out the other 20%

  # fraction of markers/features sampled for each tree, reduces overfitting and makes the model more robust.
  colsample_bytree = c(0.8), #40 * 0.8 = 32 markers

  # Minimum amount of weighted data required in a child node.
  # Higher values make the model more conservative, especially for rare classes
  min_child_weight = c(1, 5)
)

# Max N of boosting rounds, 1 round adds one tree per class for multiclass classification
# Early stopping below can stop before this limit if validation loss stops improving
nround <- 300

# Store the full CV models and a compact summary of their performance.
models <- list()
cv_summary <- list()

# Loop over all parameter combinations
for (i in seq_len(nrow(param_grid))) {
  
  # Parameters for the current XGBoost model
  params <- list(
    booster = "gbtree",

    # Multiclass classification with predicted probabilities for each cell type
    objective = "multi:softprob",

    # Optimize and monitor multiclass log-loss.
    #  <  mlogloss means predicted probabilities are closer to true labels.
    eval_metric = "mlogloss", #CD8 probability = 0.95 - low loss, good

    # Number of cell-type classes.
    num_class = num_classes,

    max_depth = param_grid$max_depth[i],
    eta = param_grid$eta[i],
    subsample = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i],
    min_child_weight = param_grid$min_child_weight[i],

    # Histogram-based tree construction, faster for larger datasets.
    tree_method = "hist",
    nthread = max(1, parallel::detectCores() - 1)
  )
  
  message(
    "Running: max_depth=", params$max_depth,
    "; eta=", params$eta,
    "; subsample=", params$subsample,
    "; colsample_bytree=", params$colsample_bytree,
    "; min_child_weight=", params$min_child_weight
  )
  
  # Cross-validation on image-level folds.
  # Each fold holds out complete images, so cells from the same image are not
  # split between training and validation.
  cv_results <- xgboost::xgb.cv(
    params = params,
    data = Xy,
    folds = fold_list, #@Round 1: train Fold2 + Fold3 + Fold4 + Fold5, test Fold1, etc.
    nrounds = nround, #300 ceiling, 107 - best

    # Stop if validation log-loss does not improve for 20 rounds.
    # This helps prevent overfitting and saves computation time.
    early_stopping_rounds = 20, #can go more after the best (108), if not will keep 108

    verbose = F, #if T - there is a message about the ML process

    # During grid search we only need test loss, not cell-level predictions
    prediction = F
  )
  
  models[[i]] <- cv_results
  
  # Select the boosting round with the lowest validation log-loss.
  best_iteration <- which.min(cv_results$evaluation_log$test_mlogloss_mean)
  best_mlogloss <- cv_results$evaluation_log$test_mlogloss_mean[best_iteration]

  # Store parameter values and their best CV score.
  cv_summary[[i]] <- data.frame(
    param_grid[i, ],
    best_iteration = best_iteration,
    best_mlogloss = best_mlogloss
  )
}

# Combine all model summaries into one table for selecting the best parameters.
cv_summary <- bind_rows(cv_summary)
```


```{r show-cv-summary}
cv_summary |> 
  dplyr::arrange(best_mlogloss)
```


## cross-validation on the best param

```{r select-best-model, message = FALSE, eval = TRUE}
# parameter combination with the lowest cross-validation log-loss
best_params <- cv_summary |> 
  arrange(best_mlogloss) |> 
  dplyr::slice(1)

best_params

# > best_params
#   max_depth eta subsample colsample_bytree
# 1         4 0.1       0.8              0.8
#   min_child_weight best_iteration best_mlogloss
# 1                5            108      0.292158

# Use the best number of boosting rounds found during CV.
# This is the round with the lowest validation log-loss before early stopping.
best_nrounds <- as.integer(best_params$best_iteration)

# Define the final set of XGBoost parameters selected by grid search.
params_best <- list(
  booster = "gbtree",

  # Multiclass classification with class probabilities as output.
  objective = "multi:softprob",

  # Evaluation metric used during tuning.
  eval_metric = "mlogloss",

  # Number of cell-label classes.
  num_class = num_classes,

  # Best tree and regularization parameters from CV.
  max_depth = as.integer(best_params$max_depth),
  eta = best_params$eta,
  subsample = best_params$subsample,
  colsample_bytree = best_params$colsample_bytree,
  min_child_weight = best_params$min_child_weight,

  # Faster tree construction for larger datasets.
  tree_method = "hist",
  nthread = max(1, parallel::detectCores() - 1)
)

# Rerun cross-validation only for the selected best model.
# Here prediction = TRUE stores out-of-fold predictions for each training cell.
# These predictions are used to estimate per-cell-type performance.
best_cv <- xgboost::xgb.cv(
  params = params_best,
  data = Xy,
  folds = fold_list,
  nrounds = best_nrounds,
  verbose = FALSE,
  prediction = TRUE
)

```


## Check CV accuracy per cell type for the best model

```{r cv-oof-predictions, fig.height=8, fig.width=12}
# Extract out-of-fold predicted class probabilities from the best CV model.
# Rows are cells, columns are cell-label classes.
cv_prob <- best_cv$cv_predict

if (is.list(cv_prob)) {
  cv_prob <- do.call(rbind, cv_prob)
}

if (is.null(dim(cv_prob))) {
  cv_prob <- matrix(
    cv_prob,
    ncol = num_classes,
    byrow = TRUE
  )
}

cv_prob <- as.matrix(cv_prob)
storage.mode(cv_prob) <- "double"

# Check that the prediction matrix has one row per training cell
# and one column per cell-label class.
stopifnot(nrow(cv_prob) == length(y))
stopifnot(ncol(cv_prob) == num_classes)

# Convert predicted probabilities to predicted cell labels.
# max.col() finds the class with the highest probability for each cell.
# label_mapping converts numeric class positions back to cell-type names.
cv_pred <- factor(
  label_mapping[max.col(cv_prob)],
  levels = label_mapping
)

# Compare predicted labels against the true training labels.
# Because predictions are out-of-fold, each cell was predicted by a model
# that was not trained on that cell/image.
cm_cv <- caret::confusionMatrix(
  data = cv_pred,
  reference = y_true
)

# cm_cv$table
#              Reference
# Prediction        B   BnT   CD4   CD8    DC Fibro   Igg   Mac MacCD204 MacCD209
#   B             587     2     0     0     2     0     1     0        0        0
#   BnT             0   230     1     0     0     0     0     0        0        0
#   CD4             5    12  1796    21     2     0     3     0        0        2
#   CD8             0     1     9  2464     3     0     0     0        2        1
#   DC              1     0     0     0   346     0     0     0       11        2
#   Fibro           0     0     1     1     3   738    55     0        0        2
#   Igg             0     0     0     0     0    21    66     0        0        0
#   Mac             0     0     0     0     0     0     0   509       66       27
#   MacCD204        0     0     1     0    11     3     1    36     2711        1
#   MacCD209        0     1     2     0     1     1    13    50        1      328
#   MDSC            0     0     0     0     0     0     0     0        0        0
#   mregDC          0     0     1     0     0     1     0     1        2        2
#   Neutrophil      0     0     0     0     0     0     0     0        0        0
#   NK              0     0     1     0     1     4     0     0        0        0
#   PMN_MDSC        0     0     1     0     0     0     0     0        0        0
#   SMA             0     0     0     0     0    12     0     0        0        0
#   Treg            0     1     9     1     0     0     1     0        1        1
#   Tumor           0     0     1     0     0    80     0     0        0        0
#   undefined       0     0    20     1     0     2     1     0        0        0
#   Vasculature     0     0     0     1     0     0     0     0        0        0
#   vCAF            0     0     0     0     0     0     0     0        0        0
```

check where fibroblasts come from

```{r}
train_sce$predicted <- cv_pred
table(train_sce$predicted,train_sce$cell_labels)
table(train_sce$cell_labels)
table(test_sce$cell_labels)
cur_sce <- train_sce[,train_sce$cell_labels == "Tumor" & train_sce$predicted == "Fibro"]
table(cur_sce$sample_id)
table(cur_sce$predicted,cur_sce$cell_labels)
```


```{r cv-accuracy-per-celltype}

cat(
  "Accuracy =", round(cm_cv$overall["Accuracy"], 4), "\n",
  "Kappa =", round(cm_cv$overall["Kappa"], 4), "\n",
  "AccuracyLower =", round(cm_cv$overall["AccuracyLower"], 4), "\n",
  "AccuracyUpper =", round(cm_cv$overall["AccuracyUpper"], 4), "\n",
  "AccuracyNull =", round(cm_cv$overall["AccuracyNull"], 4), "\n",
  "AccuracyPValue =", signif(cm_cv$overall["AccuracyPValue"], 3), "\n",
  "McnemarPValue =", cm_cv$overall["McnemarPValue"], "\n\n",
  sep = ""
)

# 97.97% of out-of-fold predictions are correct.
# Kappa corrects for agreement expected by chance/class imbalance; excellent agreement.
# AccuracyNull is the accuracy obtained by always predicting the most common class.

# AccuracyPValue = 0
# The model performs significantly better than a null/majority-class classifier.
# McnemarPValue = NaN (not informative in this setting).
```

```{r cv-byclass-performance, fig.height=8, fig.width=12}
cm_cv$byClass

# > cm_cv$byClass
#                    Sensitivity Specificity Pos Pred Value
# Class: B             0.9898820   0.9998836      0.9882155
# Class: BnT           0.9311741   0.9999669      0.9913793
# Class: CD4           0.9744981   0.9989474      0.9666308
# Class: CD8           0.9899558   0.9996567      0.9919485
# Class: DC            0.9376694   0.9995859      0.9326146

# Sensitivity (true positive rate): of all true cells of that class,
# how many were correctly identified.

# Specificity: of all cells that do not belong to that class,
# how many were correctly excluded.

# Positive Predictive Value (precision): of all cells predicted as that class,
# how many truly belong to that class.

# B, CD4 and CD8 show particularly strong classification performance.
```


## FPR vs TPR on the best model param

```{r accuracy-per-cell-type, fig.height=8, fig.width=12}
cv_perf <- data.frame(cm_cv$byClass) |> 
  mutate(
    class = sub("Class: ", "", rownames(cm_cv$byClass)),
    fpr = 1 - Specificity
  )

ggplot(cv_perf) +
  geom_point(
    aes(
      x = fpr,
      y = Sensitivity,
      size = Detection.Rate,
      fill = class
    ),
    shape = 21,
    color = "black",
    alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    aes(
      x = fpr,
      y = Sensitivity,
      label = class
    ),
    size = 3,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  scale_fill_manual(
    values = metadata(spe_27)$colour_vectors$cell_types,
    guide = "none"
  ) +
  scale_size(
    range = c(1, 15),
    guide = guide_legend(title.position = "top")
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_text(hjust = 0.5)
  ) +
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")

```




## check best model 
extract evaluation dataframes

```{r extract-evaluation-dataframes}
#models <- readRDS(file.path(base_path, "ROut_rds_all", "models.rds"))

eval_data <- purrr::imap_dfr(models, function(model, model_id) {
  
  df <- model$evaluation_log
  
  best_iteration <- which.min(df$test_mlogloss_mean)
  best_mlogloss <- df$test_mlogloss_mean[best_iteration]
  
  df |> 
    mutate(
      model_id = model_id,
      max_depth = model$params$max_depth,
      eta = model$params$eta,
      subsample = model$params$subsample,
      colsample_bytree = model$params$colsample_bytree,
      min_child_weight = model$params$min_child_weight,
      best_iteration = best_iteration,
      best_mlogloss = best_mlogloss,
      is_best_iteration = iter == best_iteration
    ) |> 
    pivot_longer(
      cols = c(
        train_mlogloss_mean,
        train_mlogloss_std,
        test_mlogloss_mean,
        test_mlogloss_std
      ),
      names_to = c("Type", ".value"),
      names_pattern = "(train|test)_mlogloss_(mean|std)"
    ) |> 
    dplyr::rename(
      mlogloss_mean = mean,
      mlogloss_std = std
    ) |> 
    mutate(
      Type = stringr::str_to_title(Type),
      max_depth = as.factor(max_depth),
      eta = as.factor(eta),
      min_child_weight = as.factor(min_child_weight),
      params = paste(
        "depth", max_depth,
        "eta", eta,
        "sub", subsample,
        "col", colsample_bytree,
        "mcw", min_child_weight,
        sep = "_"
      )
    )
})

eval_data
```

```{r best-model-summary}
best_model_summary <- eval_data |> 
  dplyr::filter(Type == "Test", is_best_iteration) |> 
  arrange(best_mlogloss) |> 
  select(
    model_id,
    max_depth,
    eta,
    subsample,
    colsample_bytree,
    min_child_weight,
    best_iteration,
    best_mlogloss,
    params
  )

best_model_summary

```



## visualize model evaluation

```{r model-evaluation}
plot_heatmap_data <- eval_data |> 
  dplyr::group_by(
    Type,
    eta,
    subsample,
    colsample_bytree,
    min_child_weight,
    max_depth
  ) |> 
  dplyr::summarise(
    mlogloss = min(mlogloss_mean),
    .groups = "drop"
  ) |> 
  dplyr::mutate(
    eta = as.factor(eta),
    subsample = as.factor(subsample),
    colsample_bytree = as.factor(colsample_bytree),
    min_child_weight = as.factor(min_child_weight),
    max_depth = as.factor(max_depth)
  )
```

```{r model-evaluation-2}
make_heatmap <- function(plot_data, cur_type, show_param_annotation = TRUE) {
  
  df <- plot_data |> 
    dplyr::filter(Type == cur_type) |> 
    arrange(
      eta,
      subsample,
      colsample_bytree,
      min_child_weight,
      max_depth
    ) |> 
    pivot_wider(
      names_from = max_depth,
      values_from = mlogloss
    )
  
  depth_cols <- sort(unique(as.character(plot_data$max_depth)))
  
  mat <- df |> 
    select(dplyr::all_of(depth_cols)) |> 
    as.matrix()
  
  cur_col_fun <- if (cur_type == "Train") {
  circlize::colorRamp2(
    c(
      min(mat, na.rm = TRUE),
      median(mat, na.rm = TRUE),
      max(mat, na.rm = TRUE)
    ),
    c("#2166ac", "white", "#b2182b")
  )
} else {
  circlize::colorRamp2(
    c(
      min(mat, na.rm = TRUE),
      median(mat, na.rm = TRUE),
      max(mat, na.rm = TRUE)
    ),
    c("darkgreen", "white", "red")
  )
}
  
  anno_colors <- list(
  "learning rate" = c(
    "0.05" = "#1b9e77",
    "0.1"  = "#d95f02"
  ),
  "cell subsampl" = c(
    "0.8" = "#7570b3"
  ),
  "marker subsam" = c(
    "0.8" = "#66a61e"
  ),
  "min child w" = c(
    "1" = "#e7298a",
    "5" = "#e6ab02"
  )
)
  
  row_anno <- if (show_param_annotation) {
    ComplexHeatmap::rowAnnotation(
      "learning rate" = df$eta,
      "cell subsampl" = df$subsample,
      "marker subsam" = df$colsample_bytree,
      "min child w" = df$min_child_weight,
      col = anno_colors
    )
  } else {
    NULL
  }
  
  ComplexHeatmap::Heatmap(
    mat,
    name = paste0(cur_type, "\nmlogloss"),
    column_title = paste0(cur_type, " multiclass log-loss"),
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    col = cur_col_fun,
    left_annotation = row_anno,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 10),
    show_row_names = FALSE,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid::grid.text(
        sprintf("%.3f", mat[i, j]),
        x,
        y,
        gp = grid::gpar(fontsize = 9)
      )
    }
  )
}

hm_train <- make_heatmap(plot_heatmap_data, "Train", show_param_annotation = TRUE)
hm_test  <- make_heatmap(plot_heatmap_data, "Test", show_param_annotation = FALSE)

```



```{r model-evaluation-heatmap, fig.height=7.5, fig.width=12}
hm_train + hm_test
```


The best cross-validation performance was obtained with 
eta = 0.1, subsample = 0.8, colsample_bytree = 0.8, 
min_child_weight = 5, and max_depth = 3. 

The test mlogloss indicates that max_depth = 3 gives the best balance between model complexity and generalization, 
while deeper trees tend to reduce train loss but increase test loss, suggesting overfitting. 

The higher min_child_weight value performs better, 
indicating that a more conservative tree-splitting strategy is preferable for these data


```{r best-model-summary-table, fig.height=10, fig.width=12}
best_model_summary <- eval_data |> 
  dplyr::filter(Type == "Test", is_best_iteration) |> 
  arrange(best_mlogloss) |> 
  dplyr::slice(1)

best_model_summary
```


```{r best-model-evaluation-curve, fig.height=10, fig.width=12}
best_param_name <- best_model_summary$params
best_iter <- best_model_summary$best_iteration
best_loss <- best_model_summary$best_mlogloss

```



```{r model-evaluation-curves, fig.height=10, fig.width=14}
ggplot(
  eval_data,
  aes(
    x = iter,
    y = mlogloss_mean,
    color = params,
    group = params
  )
) +
  geom_line(alpha = 0.75, linewidth = 0.6) +
  geom_vline(
    data = best_model_summary,
    aes(xintercept = best_iteration),
    inherit.aes = FALSE,
    linetype = "dashed",
    color = "black"
  ) +
  geom_point(
    data = best_model_summary,
    aes(
      x = best_iteration,
      y = best_mlogloss
    ),
    inherit.aes = FALSE,
    size = 3,
    color = "black"
  ) +
  facet_grid(rows = vars(Type), cols = vars(eta), scales = "free_y") +
  scale_color_brewer(palette = "Set3") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    legend.title = element_blank()
  ) +
  labs(
    title = "XGBoost Multiclass Log-Loss Across Boosting Iterations",
    subtitle = paste0(
      "Best CV model: ",
      best_param_name,
      "; iteration = ", best_iter,
      "; test mlogloss = ", round(best_loss, 3)
    ),
    x = "Boosting iteration",
    y = "Mean multiclass log-loss"
  )
```


Columns separate the learning rate values: eta = 0.05 on the left and eta = 0.1 on the right. 
In both training and test panels, log-loss decreases as more boosting iterations are added. 
Training loss continues to approach zero, especially for more flexible models, while test loss decreases initially and then plateaus. 
This gap between train and test loss reflects expected overfitting as boosting continues

The black dashed vertical line marks the selected best iteration (105)

The selected XGBoost model showed strong grouped cross-validation performance, with an overall accuracy of 95.1% and Cohen's kappa of 0.896
learly above the no-information rate of 73.6%. 
Most major cell types were classified with high sensitivity and precision, including B, BnT, CD4, CD8, DC, MacCD204, SMA, Treg, Tumor, undefined, and Vasculature

The main weaker classes were Fibro, Igg, Neutrophil, and PMN_MDSC. 
Fibro and PMN_MDSC showed high sensitivity but low precision, indicating that the model over-predicts these labels. 
Igg and Neutrophil showed low sensitivity, meaning that a substantial fraction of true cells from these classes are missed. 
These classes should therefore be interpreted with caution in downstream predictions.


# final classifier

```{r train-final-xgboost-model}
# Train the final XGBoost model on all training cells using the best CV parameters.
# CV is only used to select parameters; this model is the one used for held-out
# test evaluation and later prediction on unlabelled cells.
final_model <- xgboost::xgb.train(
  params = params_best,
  data = Xy,
  nrounds = best_nrounds,
  verbose = TRUE
)

```


save final model via saveRDS

```{r train-final-model, message = FALSE, eval = TRUE}
# saveRDS(
#   final_model,
#   file = file.path(base_path, "classifier_xgboost.rds")
# )
```



This uses:

max_depth = 3
eta = 0.1
subsample = 0.8
colsample_bytree = 0.8
min_child_weight = 5
nrounds = 108

```{r xgboost-variable-importance, fig.height=8, fig.width=8}
# final_model <- readRDS(
#   file.path(base_path, "classifier_xgboost.rds")
# )

# Extract marker importance from the trained XGBoost model.
# Gain reflects how much each marker contributed to reducing the loss.
#xgb.importance(model = final_model,feature_names =c("SMA"))

var_imp <- xgboost::xgb.importance(
  feature_names = colnames(X),
  model = final_model
)

var_imp |> 
  slice_max(Gain, n = 20)

var_imp_top <- var_imp |> 
  dplyr::slice_max(Gain, n = 20) |> 
  dplyr::mutate(
    Feature = forcats::fct_reorder(Feature, Gain)
  )

```



```{r variable-importance, fig.height=8, fig.width=8}
ggplot(var_imp_top, aes(x = Gain, y = Feature)) +
  geom_col(fill = "steelblue") +
  theme_classic() +
  labs(
    title = "Top XGBoost Marker Importances",
    x = "Gain",
    y = "Marker"
  )
```



Variable importance was assessed using XGBoost gain, which measures how much each marker contributes to reducing the model loss across tree splits. 
The most informative markers were CD3, CD204, CD209, CD20, Lamp3, CD31, FOXP3, CD66b, and CD146. 
These markers correspond well to the expected biology of the annotated cell types, including T cells, B cells, macrophage/DC populations, vasculature, neutrophil-like cells, and stromal compartments.

This is a useful sanity check: the model is not relying on strange markers; it is using biologically meaningful features.

# prepare test data

```{r prepare-test-data}
test_sce$cell_id <- paste0(
  test_sce$sample_id,
  "_",
  test_sce$ObjectNumber
)

colnames(test_sce) <- test_sce$cell_id

# Labelled cells only for evaluation
test_eval_keep <- !is.na(test_sce$cell_labels) &
  test_sce$cell_labels != "unlabelled" &
  test_sce$cell_labels %in% label_mapping

# Use the same marker set as training
X_test <- t(assay(test_sce, "exprs")[marker_keep, test_eval_keep])

# Check that train and test feature columns match
stopifnot(identical(colnames(X), colnames(X_test)))

# Convert test labels using the training label mapping
y_test_true <- factor(
  test_sce$cell_labels[test_eval_keep],
  levels = label_mapping
)

y_test <- as.integer(y_test_true) - 1

stopifnot(!anyNA(y_test))

# Store labelled test data in XGBoost format
Xy_test <- xgboost::xgb.DMatrix(
  data = X_test,
  label = y_test
)

Xy_test

```

## visualize predicted results

```{r test-performance-check, fig.height=8, fig.width=12}

raw_test_pred <- predict(final_model, newdata = Xy_test)

# Try both possible reshaping directions
test_prob_byrow <- matrix(
  raw_test_pred,
  ncol = num_classes,
  byrow = TRUE
)

test_prob_bycol <- matrix(
  raw_test_pred,
  ncol = num_classes,
  byrow = FALSE
)

# Check which one has probabilities summing to 1 per cell
mean(abs(rowSums(test_prob_byrow) - 1))

mean(abs(rowSums(test_prob_bycol) - 1))
```

```{r test-performance-final, fig.height=8, fig.width=12}
if (
  mean(abs(rowSums(test_prob_byrow) - 1)) <
  mean(abs(rowSums(test_prob_bycol) - 1))
) {
  test_prob <- test_prob_byrow
} else {
  test_prob <- test_prob_bycol
}

stopifnot(nrow(test_prob) == sum(test_eval_keep))
stopifnot(nrow(test_prob) == nrow(X_test))
stopifnot(ncol(test_prob) == num_classes)
stopifnot(mean(abs(rowSums(test_prob) - 1)) < 1e-6)

rownames(test_prob) <- colnames(test_sce)[test_eval_keep]
colnames(test_prob) <- label_mapping

```

```{r}
test_pred <- factor(
  label_mapping[max.col(test_prob, ties.method = "first")],
  levels = label_mapping
)

test_idx <- match(
  rownames(test_prob),
  colnames(test_sce)
)

stopifnot(!anyNA(test_idx))

test_true <- factor(
  test_sce$cell_labels[test_idx],
  levels = label_mapping
)

stopifnot(
  length(test_pred) == length(test_true),
  !anyNA(test_true)
)

cm_test <- caret::confusionMatrix(
  data = test_pred,
  reference = test_true
)
```


```{r}
cm_test$table
```


```{r}
cm_test$overall
```



```{r test-performance, fig.height=8, fig.width=12}
# Per-cell-type performance plot.
test_perf <- data.frame(cm_test$byClass) |> 
  dplyr::mutate(
    class = sub("Class: ", "", rownames(cm_test$byClass)),
    fpr = 1 - Specificity
  )

ggplot(test_perf) +
  geom_point(
    aes(
      x = fpr,
      y = Sensitivity,
      size = Detection.Rate,
      fill = class
    ),
    shape = 21,
    color = "black",
    alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    aes(
      x = fpr,
      y = Sensitivity,
      label = class
    ),
    size = 3,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  scale_fill_manual(
    values = metadata(spe_27)$colour_vectors$cell_types,
    guide = "none"
  ) +
  scale_size(
    range = c(1, 15),
    guide = guide_legend(title.position = "top")
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_text(hjust = 0.5)
  ) +
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")
```

## Class-specific protein markers driving cell-type predictions

```{r}
shap_raw <- predict(
  final_model,
  newdata = xgboost::xgb.DMatrix(X_test),
  predcontrib = TRUE
)

dim(shap_raw)
```


```{r class-specific-shap-setup}

feature_names <- colnames(X_test)

# Remove BIAS column from SHAP array
shap_values <- shap_raw[, , 1:length(feature_names)]

stopifnot(dim(shap_values)[1] == nrow(X_test))
stopifnot(dim(shap_values)[2] == length(label_mapping))
stopifnot(dim(shap_values)[3] == length(feature_names))

# Function to extract top SHAP markers for one class
get_class_shap <- function(class_name, top_n = 15) {
  
  class_id <- which(label_mapping == class_name)
  
  if (length(class_id) != 1) {
    stop("Class not found in label_mapping: ", class_name)
  }
  
  shap_class <- shap_values[, class_id, , drop = FALSE]
  shap_class <- shap_class[, 1, ]
  
  out <- data.frame(
    class = class_name,
    marker = feature_names,
    mean_abs_shap = apply(abs(shap_class), 2, mean),
    mean_shap = apply(shap_class, 2, mean)
  )
  
  out <- out |>
    dplyr::arrange(dplyr::desc(mean_abs_shap)) |>
    dplyr::slice_head(n = top_n)
  
  out
}

tumor_shap <- get_class_shap("Tumor", top_n = 15)
fibro_shap <- get_class_shap("Fibro", top_n = 15)
sma_shap   <- get_class_shap("SMA", top_n = 15)
vcaf_shap  <- get_class_shap("vCAF", top_n = 15)

tumor_shap
fibro_shap
sma_shap
vcaf_shap
```



```{r class-specific-shap-true-cells}

get_class_shap_true_cells <- function(class_name, top_n = 10) {
  
  class_id <- which(label_mapping == class_name)
  
  if (length(class_id) != 1) {
    stop("Class not found in label_mapping: ", class_name)
  }
  
  keep_cells <- test_true == class_name
  
  if (sum(keep_cells) == 0) {
    stop("No test cells found for class: ", class_name)
  }
  
  shap_class <- shap_values[keep_cells, class_id, , drop = FALSE]
  shap_class <- shap_class[, 1, ]
  
  out <- data.frame(
    class = class_name,
    marker = feature_names,
    mean_shap = apply(shap_class, 2, mean),
    mean_abs_shap = apply(abs(shap_class), 2, mean)
  )
  
  positive <- out |>
    dplyr::filter(mean_shap > 0) |>
    dplyr::arrange(dplyr::desc(mean_shap)) |>
    dplyr::slice_head(n = top_n) |>
    dplyr::mutate(direction = "positive")
  
  negative <- out |>
    dplyr::filter(mean_shap < 0) |>
    dplyr::arrange(mean_shap) |>
    dplyr::slice_head(n = top_n) |>
    dplyr::mutate(direction = "negative")
  
  dplyr::bind_rows(positive, negative)
}

shap_true_directional <- dplyr::bind_rows(
  get_class_shap_true_cells("Tumor", top_n = 10),
  get_class_shap_true_cells("Fibro", top_n = 10),
  get_class_shap_true_cells("SMA", top_n = 10),
  get_class_shap_true_cells("vCAF", top_n = 10)
)

shap_true_directional

```


```{r class-specific-shap-plot, fig.height=8, fig.width=12}
shap_plot_df <- shap_true_directional |>
  dplyr::filter(direction == "positive") |>
  dplyr::group_by(class) |>
  dplyr::slice_max(mean_shap, n = 8) |>
  dplyr::ungroup()

ggplot(
  shap_plot_df,
  aes(
    x = reorder(marker, mean_shap),
    y = mean_shap
  )
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~ class, scales = "free_y") +
  theme_bw(base_size = 13) +
  xlab("Protein marker") +
  ylab("Mean positive SHAP value") +
  ggtitle("Class-specific protein markers driving XGBoost predictions")
```

## Interpretable protein-state map of XGBoost-predicted cell populations

```{r protein-state-shap-map-publication, fig.height=7, fig.width=13}
library(ggplot2)
library(ggrepel)

state_markers <- shap_true_directional |>
  dplyr::filter(direction == "positive") |>
  dplyr::group_by(class) |>
  dplyr::slice_max(mean_shap, n = 6) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    cell_state = dplyr::case_when(
      class == "Tumor" ~ "Epithelial tumor state",
      class == "Fibro" ~ "Podoplanin+ stromal state",
      class == "SMA" ~ "Contractile myofibroblast state",
      class == "vCAF" ~ "Vascular-associated CAF state",
      TRUE ~ class
    ),
    marker_group = dplyr::case_when(
      marker %in% c("Ecad", "panCK", "EGFR") ~ "Epithelial / tumor",
      marker %in% c("SMA", "Podoplanin", "CD10", "CD73") ~ "Stromal / CAF",
      marker %in% c("CD31", "CD146") ~ "Vascular / perivascular",
      marker %in% c("CD45", "CD3", "CD20", "CD204", "CD209", "CD11c", "CD66b") ~ "Immune-associated",
      TRUE ~ "Other"
    ),
    label_marker = ifelse(mean_shap > 0.25, marker, NA)
  )

ggplot(
  state_markers,
  aes(
    x = marker,
    y = cell_state,
    size = mean_abs_shap,
    fill = mean_shap
  )
) +
  geom_point(shape = 21, color = "black", alpha = 0.9) +
  ggrepel::geom_text_repel(
    aes(label = label_marker),
    size = 3.5,
    na.rm = TRUE,
    max.overlaps = Inf
  ) +
  facet_grid(. ~ marker_group, scales = "free_x", space = "free_x") +
  scale_size(range = c(3, 13)) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    title = "Protein-state map of XGBoost-predicted cell populations",
    subtitle = "Class-specific SHAP values identify protein programs driving cell-state prediction",
    x = "Protein marker",
    y = "Interpreted cell state",
    size = "Mean |SHAP|",
    fill = "Mean SHAP"
  )
```
## XGBoost-derived protein-state architecture

```{r shap-protein-state-architecture, fig.height=6, fig.width=10}
library(ggplot2)

# Build SHAP matrix: rows = cell states, columns = protein markers
shap_state_matrix <- shap_true_directional |>
  dplyr::group_by(class, marker) |>
  dplyr::summarise(
    mean_shap = mean(mean_shap),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = marker,
    values_from = mean_shap,
    values_fill = 0
  )

shap_state_long <- shap_state_matrix |>
  tidyr::pivot_longer(
    cols = -class,
    names_to = "marker",
    values_to = "mean_shap"
  ) |>
  dplyr::mutate(
    class = factor(
      class,
      levels = c("Tumor", "Fibro", "SMA", "vCAF")
    )
  )

ggplot(
  shap_state_long,
  aes(
    x = marker,
    y = class,
    fill = mean_shap
  )
) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient2(
    low = "navy",
    mid = "white",
    high = "firebrick",
    midpoint = 0
  ) +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "XGBoost-derived protein-state architecture",
    subtitle = "Positive and negative SHAP values reveal protein programs defining cell states",
    x = "Protein marker",
    y = "Cell state",
    fill = "Mean SHAP"
  )

```

## Phenotypic trajectory of stromal-to-tumor protein states

```{r}
library(scater)
library(dplyr)

set.seed(42)

test_sce <- scater::runUMAP(
  test_sce,
  exprs_values = "exprs",
  subset_row = feature_names,
  name = "UMAP"
)

umap_df <- data.frame(
  UMAP1 = reducedDim(test_sce, "UMAP")[, 1],
  UMAP2 = reducedDim(test_sce, "UMAP")[, 2],
  true_label = test_true,
  pred_label = test_pred,
  is_correct = test_true == test_pred
)

head(umap_df)
```

```{r xgboost_predicted_protein_defined_umap, fig.height=8.5, fig.width=12.5}

library(ggplot2)
library(dplyr)

focus_classes <- c(
  "Tumor", "Fibro", "SMA", "vCAF",
  "Vasculature", "MacCD204", "CD4", "CD8"
)

state_order <- c(
  "Tumor", "Fibro", "SMA", "vCAF",
  "Vasculature", "MacCD204", "CD4", "CD8",
  "Other predicted classes"
)

plot_df <- umap_df |>
  mutate(
    state_raw = ifelse(
      pred_label %in% focus_classes,
      as.character(pred_label),
      "Other predicted classes"
    ),
    state_raw = factor(state_raw, levels = state_order)
  )

state_counts <- table(plot_df$state_raw)

plot_df <- plot_df |>
  mutate(
    state = paste0(
      state_raw,
      " (n=",
      state_counts[as.character(state_raw)],
      ")"
    ),
    state = factor(
      state,
      levels = paste0(
        state_order,
        " (n=",
        state_counts[state_order],
        ")"
      )
    )
  )

state_colors <- setNames(
  c(
    "#E64B35",
    "#00BFC4",
    "#B65FCF",
    "#FF1493",
    "#1F78B4",
    "#009E73",
    "#253494",
    "#FFD700",
    "#C9C9C9"
  ),
  levels(plot_df$state)
)

get_dense_center <- function(df, k = 3) {

  if (nrow(df) < 5) {
    return(data.frame(
      point_x = median(df$UMAP1),
      point_y = median(df$UMAP2)
    ))
  }

  set.seed(42)

  km <- kmeans(
    df[, c("UMAP1", "UMAP2")],
    centers = min(k, nrow(df)),
    nstart = 50
  )

  main_cluster <- as.integer(
    names(which.max(table(km$cluster)))
  )

  df_main <- df[km$cluster == main_cluster, ]

  data.frame(
    point_x = median(df_main$UMAP1),
    point_y = median(df_main$UMAP2)
  )
}

label_df <- plot_df |>
  filter(state_raw != "Other predicted classes") |>
  group_by(state_raw) |>
  group_modify(~ get_dense_center(.x)) |>
  ungroup() |>
  mutate(
    label_x = case_when(
      state_raw == "Tumor" ~ point_x + 0.2,
      state_raw == "MacCD204" ~ point_x - 1.2,
      state_raw == "Fibro" ~ point_x - 0.9,
      state_raw == "SMA" ~ point_x - 0.4,
      state_raw == "vCAF" ~ point_x - 0.8,
      state_raw == "Vasculature" ~ point_x - 1.0,
      state_raw == "CD4" ~ point_x + 0.8,
      state_raw == "CD8" ~ point_x + 1.0,
      TRUE ~ point_x + 0.4
    ),
    label_y = case_when(
      state_raw == "Tumor" ~ point_y + 1.0,
      state_raw == "MacCD204" ~ point_y + 0.7,
      state_raw == "Fibro" ~ point_y - 0.1,
      state_raw == "SMA" ~ point_y - 0.5,
      state_raw == "vCAF" ~ point_y - 0.9,
      state_raw == "Vasculature" ~ point_y - 1.1,
      state_raw == "CD4" ~ point_y + 0.6,
      state_raw == "CD8" ~ point_y - 0.8,
      TRUE ~ point_y + 0.5
    )
  )

ggplot() +

  geom_point(
    data = plot_df |> filter(state_raw == "Other predicted classes"),
    aes(UMAP1, UMAP2, color = state),
    size = 0.40,
    alpha = 0.15,
    stroke = 0
  ) +

  geom_point(
    data = plot_df |> filter(state_raw == "Tumor"),
    aes(UMAP1, UMAP2, color = state),
    size = 0.50,
    alpha = 0.55,
    stroke = 0
  ) +

  geom_point(
    data = plot_df |> filter(
      !state_raw %in% c(
        "Tumor",
        "Other predicted classes"
      )
    ),
    aes(UMAP1, UMAP2, color = state),
    size = 1.55,
    alpha = 0.97,
    stroke = 0
  ) +

  geom_segment(
    data = label_df,
    aes(
      x = label_x,
      y = label_y,
      xend = point_x,
      yend = point_y
    ),
    linewidth = 0.30,
    color = "grey35"
  ) +

  geom_text(
    data = label_df,
    aes(
      x = label_x,
      y = label_y,
      label = state_raw
    ),
    size = 4.2,
    fontface = "bold",
    color = "black"
  ) +

  scale_color_manual(
    values = state_colors,
    drop = FALSE
  ) +

  guides(
    color = guide_legend(
      override.aes = list(
        size = 4.8,
        alpha = 1
      )
    )
  ) +

  coord_equal() +

  theme_bw(base_size = 13) +

  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    plot.subtitle = element_text(
      size = 12
    ),
    legend.position = "right",
    legend.title = element_text(
      face = "bold",
      size = 12
    ),
    legend.text = element_text(
      size = 9.3
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      linewidth = 0.22,
      color = "grey91"
    )
  ) +

  labs(
    title = "UMAP of XGBoost-predicted protein-defined cell states",
    subtitle = "Predicted protein-defined cell states; major states highlighted",
    color = "Cell state",
    x = "UMAP1",
    y = "UMAP2"
  )

```


## XGBoost classification errors reveal biologically related cell states

```{r xgboost-confusion-matrix-publication, fig.height=8, fig.width=10}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)

# Confusion matrix from held-out test set
cm_mat <- as.matrix(cm_test$table)

# Convert counts to column percentages:
# each column = true/reference class
cm_prop <- prop.table(cm_mat, margin = 2) * 100

cm_df <- as.data.frame(cm_prop) |>
  dplyr::rename(
    predicted = Prediction,
    reference = Reference,
    percent = Freq
  ) |>
  dplyr::mutate(
    is_correct = predicted == reference,
    label = ifelse(percent >= 1, sprintf("%.1f", percent), "")
  )

ggplot(
  cm_df,
  aes(
    x = reference,
    y = predicted,
    fill = percent
  )
) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_text(
    aes(label = label),
    size = 2.6,
    color = "black"
  ) +
  scale_fill_gradient(
    low = "white",
    high = "#2166AC"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right"
  ) +
  labs(
    title = "XGBoost confusion matrix on held-out test set",
    subtitle = "Column-normalized percentages show how true cell states were classified",
    x = "True cell state",
    y = "Predicted cell state",
    fill = "% of true class"
  )

```



```{r xgboost-error-map-biological-confusions, fig.height=6, fig.width=9}
error_df <- cm_df |>
  dplyr::filter(predicted != reference, percent > 0) |>
  dplyr::arrange(dplyr::desc(percent)) |>
  dplyr::slice_head(n = 20) |>
  dplyr::mutate(
    confusion = paste0(reference, " → ", predicted),
    confusion = forcats::fct_reorder(confusion, percent)
  )

ggplot(
  error_df,
  aes(
    x = percent,
    y = confusion,
    fill = percent
  )
) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_fill_gradient(
    low = "grey85",
    high = "#B2182B"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "none"
  ) +
  labs(
    title = "Dominant XGBoost classification errors",
    subtitle = "Misclassifications highlight rare or phenotypically related cell states",
    x = "% of true class misclassified",
    y = "True → predicted cell state"
  )
```

```{r check-rare-class-errors}
# Absolute number of true cells per class in the test set
test_class_counts <- colSums(cm_test$table)

sort(test_class_counts)

```


```{r check-igg-bnt-errors}
# Absolute confusion counts for small classes
cm_test$table[, c("Igg", "BnT")]

# Percent confusion for the same classes
round(prop.table(cm_test$table, margin = 2)[, c("Igg", "BnT")] * 100, 1)
```

## Spatial neighborhood analysis of XGBoost-predicted cell states

```{r}
## Build spatial coordinates object

spe_coords <- as.data.frame(spatialCoords(spe))

spe_coords$cell_id <- colnames(spe)

coords <- data.frame(
  cell_id = colnames(test_sce),
  sample_id = as.character(test_sce$sample_id),
  ObjectNumber = as.character(test_sce$ObjectNumber),
  pred_label = as.character(test_pred),
  true_label = as.character(test_true),
  stringsAsFactors = FALSE
)

coords <- coords |>
  dplyr::left_join(
    spe_coords |> dplyr::select(cell_id, Pos_X, Pos_Y),
    by = "cell_id"
  ) |>
  dplyr::rename(
    x = Pos_X,
    y = Pos_Y
  )

head(coords)
dim(coords)
sum(is.na(coords$x))
sum(is.na(coords$y))
str(coords)
```


```{r}
sum(is.na(coords$x))
# 0

sum(is.na(coords$y))
# 0
```

```{r}
spatial_pred_df <- coords

head(spatial_pred_df)
dim(spatial_pred_df)
```


```{r}
class(spatial_pred_df)
dim(spatial_pred_df)
head(spatial_pred_df)
```

```{r}
colnames(spatial_pred_df)
```


```{r}
summary(spatial_pred_df$x)
summary(spatial_pred_df$y)
```

```{r}
library(ggplot2)

ggplot(spatial_pred_df,
       aes(x = x, y = y, color = pred_label)) +
  geom_point(size = 0.3, alpha = 0.7) +
  theme_bw()
```


```{r}
focus_classes <- c(
  "Tumor","Fibro","SMA","vCAF",
  "Vasculature","MacCD204","CD4","CD8"
)

plot_df <- spatial_pred_df

plot_df$group <- ifelse(
  plot_df$pred_label %in% focus_classes,
  plot_df$pred_label,
  "Other"
)

ggplot(plot_df,
       aes(x = x, y = y, color = group)) +
  geom_point(size = 0.25, alpha = 0.8) +
  coord_fixed() +
  theme_bw(base_size = 14) +
  labs(
    title = "Spatial organization of XGBoost-predicted cell states",
    x = "X coordinate",
    y = "Y coordinate",
    color = "Cell state"
  )
```

```{r}
library(ggplot2)
library(dplyr)

focus_classes <- c(
  "Tumor", "Fibro", "SMA", "vCAF",
  "Vasculature", "MacCD204", "CD4", "CD8"
)

top_samples <- spatial_pred_df |>
  count(sample_id, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(sample_id)

spatial_plot_df <- spatial_pred_df |>
  filter(sample_id %in% top_samples) |>
  mutate(
    cell_state = ifelse(
      pred_label %in% focus_classes,
      pred_label,
      "Other"
    )
  )

ggplot(spatial_plot_df, aes(x = x, y = y, color = cell_state)) +
  geom_point(size = 0.35, alpha = 0.85) +
  coord_fixed() +
  facet_wrap(~ sample_id, ncol = 3) +

  guides(
    color = guide_legend(
      override.aes = list(
        size = 3,
        alpha = 1
      )
    )
  ) +

  theme_bw(base_size = 11) +

  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 8),
    legend.text = element_text(size = 7),

    strip.text = element_text(
      size = 5,
      face = "bold"
    ),

    panel.grid = element_blank()
  ) +

  labs(
    title = "Spatial distribution of XGBoost-predicted cell states",
    subtitle = "Representative tissue images shown separately",
    x = "X coordinate",
    y = "Y coordinate",
    color = "Cell state"
  )
```


## Random Forest validation of cell-state annotation 
## Independent Random Forest validation of protein-defined cell states

```{r random-forest-prepare-data}
set.seed(42)

rf_train_df <- as.data.frame(X)
rf_train_df$cell_label <- y_true

rf_test_df <- as.data.frame(X_test)
rf_test_df$cell_label <- y_test_true

table(rf_train_df$cell_label)
table(rf_test_df$cell_label)

```

```{r train-random-forest-model}
set.seed(42)

rf_model <- randomForest(
  cell_label ~ .,
  data = rf_train_df,
  ntree = 500,
  mtry = floor(sqrt(ncol(X))),
  importance = TRUE
)

rf_model
```


```{r random-forest-test-performance}
rf_test_pred <- predict(
  rf_model,
  newdata = rf_test_df,
  type = "response"
)

rf_test_pred <- factor(rf_test_pred, levels = label_mapping)
rf_test_true <- factor(rf_test_df$cell_label, levels = label_mapping)

cm_rf_test <- caret::confusionMatrix(
  data = rf_test_pred,
  reference = rf_test_true
)

cm_rf_test$table
cm_rf_test$overall
```


```{r random-forest-by-class-performance}
cm_rf_test$byClass
```


```{r random-forest-marker-importance, fig.height=8, fig.width=8}
rf_importance <- importance(rf_model)

rf_importance_df <- data.frame(
  marker = rownames(rf_importance),
  MeanDecreaseGini = rf_importance[, "MeanDecreaseGini"],
  row.names = NULL
) |>
  dplyr::arrange(desc(MeanDecreaseGini)) |>
  dplyr::slice_head(n = 20)

ggplot(
  rf_importance_df,
  aes(
    x = MeanDecreaseGini,
    y = reorder(marker, MeanDecreaseGini)
  )
) +
  geom_col(fill = "steelblue") +
  theme_classic(base_size = 13) +
  labs(
    title = "Top Random Forest marker importances",
    x = "Mean decrease Gini",
    y = "Marker"
  )
```

## Biological interpretation of classification errors

```{r}
## Biological stromal continuum: Fibro - vCAF - SMA

library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

marker_expr <- t(as.matrix(assay(spe, "exprs")))

marker_df <- as.data.frame(marker_expr) |>
  rownames_to_column("cell_id")

meta_df <- as.data.frame(colData(spe)) |>
  rownames_to_column("cell_id")

label_df <- as.data.frame(colData(lab_sce)) |>
  rownames_to_column("cell_id") |>
  select(cell_id, cell_labels)

bio_df <- meta_df |>
  left_join(marker_df, by = "cell_id") |>
  left_join(label_df, by = "cell_id") |>
  mutate(pred_label = cell_labels) |>
  filter(!is.na(pred_label))

stromal_markers <- c("SMA", "Podoplanin", "CD146")

bio_df |>
  filter(pred_label %in% c("Fibro", "vCAF", "SMA")) |>
  select(pred_label, all_of(stromal_markers)) |>
  pivot_longer(
    cols = all_of(stromal_markers),
    names_to = "marker",
    values_to = "expression"
  ) |>
  ggplot(aes(x = pred_label, y = expression, fill = pred_label)) +
  geom_violin(trim = FALSE, alpha = 0.75) +
  geom_boxplot(width = 0.12, outlier.size = 0.4, alpha = 0.85) +
  facet_wrap(~ marker, scales = "free_y") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Fibro–vCAF–SMA continuum",
    subtitle = "Stromal populations show related marker profiles, explaining model confusion",
    x = "Predicted cell state",
    y = "Marker expression"
  )
```
```{r}
## Biological validation of immune-related classification overlaps

library(dplyr)
library(tidyr)
library(ggplot2)

## 1. PMN_MDSC versus Neutrophil
## These populations are expected to share granulocytic/myeloid markers.

myeloid_overlap_markers <- c(
  "CD66b",
  "CD11b",
  "Arg1",
  "CD15",
  "CXCL8"
)

bio_df |>
  filter(pred_label %in% c("PMN_MDSC", "Neutrophil")) |>
  select(pred_label, all_of(myeloid_overlap_markers)) |>
  pivot_longer(
    cols = all_of(myeloid_overlap_markers),
    names_to = "marker",
    values_to = "expression"
  ) |>
  ggplot(
    aes(
      x = pred_label,
      y = expression,
      fill = pred_label
    )
  ) +
  geom_violin(
    trim = FALSE,
    alpha = 0.75
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.size = 0.4,
    alpha = 0.85
  ) +
  facet_wrap(
    ~ marker,
    scales = "free_y"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "PMN-MDSC and neutrophil marker overlap",
    subtitle = "Granulocytic/myeloid marker profiles explain partial model confusion",
    x = "Predicted cell state",
    y = "Marker expression"
  )
```
```{r}
## 2. NK versus CD8 T cells
## These populations may overlap through cytotoxic/lymphoid marker expression.

nk_cd8_markers <- c(
  "CD3",
  "CD8a",
  "NKG2A",
  "CD56",
  "EOMES",
  "Tbet"
)

bio_df |>
  filter(pred_label %in% c("NK", "CD8")) |>
  select(pred_label, all_of(nk_cd8_markers)) |>
  pivot_longer(
    cols = all_of(nk_cd8_markers),
    names_to = "marker",
    values_to = "expression"
  ) |>
  ggplot(
    aes(
      x = pred_label,
      y = expression,
      fill = pred_label
    )
  ) +
  geom_violin(
    trim = FALSE,
    alpha = 0.75
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.size = 0.4,
    alpha = 0.85
  ) +
  facet_wrap(
    ~ marker,
    scales = "free_y"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "NK and CD8 T-cell marker overlap",
    subtitle = "Shared cytotoxic/lymphoid markers explain partial model confusion",
    x = "Predicted cell state",
    y = "Marker expression"
  )
```


```{r}
## Summary table: mean marker expression by selected related cell states

related_classes <- c(
  "Fibro", "vCAF", "SMA",
  "PMN_MDSC", "Neutrophil",
  "NK", "CD8"
)

related_markers <- c(
  "SMA", "Podoplanin", "CD146",
  "CD66b", "CD11b", "Arg1", "CD15", "CXCL8",
  "CD3", "CD8a", "NKG2A", "CD56", "EOMES", "Tbet"
)

bio_marker_summary <- bio_df |>
  filter(pred_label %in% related_classes) |>
  group_by(pred_label) |>
  summarise(
    across(
      all_of(related_markers),
      ~ mean(.x, na.rm = TRUE)
    ),
    n_cells = dplyr::n(),
    .groups = "drop"
  )

bio_marker_summary
```
## Disease-relevant cell state–marker network

```{r}
## Beautiful biological disease-link network
## cell states -> markers -> biological processes -> diseases

library(dplyr)
library(tidyr)
library(tibble)
library(igraph)
library(ggraph)
library(ggplot2)
library(scales)
library(ggrepel)

marker_expr <- t(as.matrix(assay(spe, "exprs")))

marker_df <- as.data.frame(marker_expr) |>
  rownames_to_column("cell_id")

meta_df <- as.data.frame(colData(spe)) |>
  rownames_to_column("cell_id")

label_df <- as.data.frame(colData(lab_sce)) |>
  rownames_to_column("cell_id") |>
  select(cell_id, cell_labels)

bio_df <- meta_df |>
  left_join(marker_df, by = "cell_id") |>
  left_join(label_df, by = "cell_id") |>
  mutate(pred_label = cell_labels) |>
  filter(!is.na(pred_label))

bio_links <- tibble::tribble(
  ~cell_state, ~marker, ~process, ~disease, ~module,

  "Tumor", "panCK", "Epithelial tumor identity", "Carcinoma", "Tumor",
  "Tumor", "Ecad",  "Epithelial adhesion", "Carcinoma", "Tumor",
  "Tumor", "EGFR",  "Growth factor signaling", "EGFR-driven cancer", "Tumor",
  "Tumor", "Ki67",  "Cell proliferation", "Aggressive tumor growth", "Tumor",
  "Tumor", "p53",   "DNA damage / tumor stress", "Tumor progression", "Tumor",

  "Fibro", "SMA",        "Fibroblast activation", "Cancer-associated fibroblasts", "Stroma",
  "Fibro", "Podoplanin", "Stromal activation", "Invasion / metastasis", "Stroma",
  "Fibro", "MMP9",       "Matrix degradation", "Invasion / metastasis", "Stroma",

  "SMA", "SMA",        "Activated stroma", "Fibrosis-like remodeling", "Stroma",
  "SMA", "CD146",      "Vascular stromal niche", "Tumor angiogenesis", "Stroma",
  "vCAF", "CD146",     "Vascular CAF program", "Tumor angiogenesis", "Stroma",
  "vCAF", "Podoplanin", "CAF remodeling", "Invasion / metastasis", "Stroma",

  "Mac", "CD11b",  "Myeloid infiltration", "Chronic inflammation", "Myeloid",
  "Mac", "CD204",  "Suppressive macrophages", "Tumor immune evasion", "Myeloid",
  "Mac", "CD11c",  "Antigen-presenting myeloid cells", "Inflammatory tumor niche", "Myeloid",

  "MacCD204", "CD204", "M2-like macrophages", "Tumor immune evasion", "Myeloid",
  "MacCD204", "Arg1",  "Arginine metabolism / suppression", "Immunosuppression", "Myeloid",
  "MacCD204", "CD11b", "Myeloid suppressive niche", "Immunosuppression", "Myeloid",

  "PMN_MDSC", "CD66b", "Granulocytic MDSC", "Immunosuppression", "Myeloid",
  "PMN_MDSC", "Arg1",  "T-cell suppression", "Tumor immune evasion", "Myeloid",
  "PMN_MDSC", "CXCL8", "Neutrophil recruitment / inflammation", "Inflammatory cancer", "Myeloid",

  "Neutrophil", "CD15",  "Neutrophil program", "Inflammatory cancer", "Myeloid",
  "Neutrophil", "CD66b", "Granulocytic inflammation", "Inflammatory cancer", "Myeloid",
  "Neutrophil", "CXCL8", "CXCL8-driven inflammation", "Tumor angiogenesis", "Myeloid",

  "CD4", "CD3",  "T-cell immunity", "Anti-tumor immune response", "Lymphoid",
  "CD4", "CD45", "Immune cell infiltration", "Inflammatory tumor niche", "Lymphoid",

  "CD8", "CD3",  "Cytotoxic T-cell immunity", "Anti-tumor immune response", "Lymphoid",
  "CD8", "CD8a", "Cytotoxic killing", "Anti-tumor immune response", "Lymphoid",
  "CD8", "PD1",  "T-cell exhaustion", "Immune checkpoint exhaustion", "Lymphoid",

  "Treg", "FOXP3", "Regulatory T-cell program", "Tumor immune evasion", "Lymphoid",
  "Treg", "GITR",  "Treg activation", "Immunosuppression", "Lymphoid",
  "Treg", "CD3",   "Regulatory lymphoid niche", "Tumor immune evasion", "Lymphoid",

  "NK", "CD56",  "NK-cell cytotoxicity", "Anti-tumor immune response", "Lymphoid",
  "NK", "NKG2A", "NK inhibitory signaling", "Immune checkpoint exhaustion", "Lymphoid",
  "NK", "EOMES", "NK differentiation", "Anti-tumor immune response", "Lymphoid",
  "NK", "Tbet",  "Cytotoxic lymphoid program", "Anti-tumor immune response", "Lymphoid"
)

available_markers <- intersect(
  unique(bio_links$marker),
  colnames(bio_df)
)

marker_means <- bio_df |>
  group_by(pred_label) |>
  summarise(
    across(
      all_of(available_markers),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = all_of(available_markers),
    names_to = "marker",
    values_to = "mean_expression"
  ) |>
  rename(cell_state = pred_label)

network_df <- bio_links |>
  filter(marker %in% available_markers) |>
  left_join(marker_means, by = c("cell_state", "marker")) |>
  filter(!is.na(mean_expression)) |>
  filter(mean_expression > 0) |>
  mutate(
    marker_node  = paste0("Marker: ", marker),
    process_node = paste0("Process: ", process),
    disease_node = paste0("Disease: ", disease),
    module_node  = paste0("Module: ", module)
  )

edges_module_cell <- network_df |>
  distinct(module_node, cell_state, module) |>
  transmute(
    from = module_node,
    to = cell_state,
    weight = 1.5,
    edge_type = "module_lace",
    module = module
  )

edges_cell_marker <- network_df |>
  transmute(
    from = cell_state,
    to = marker_node,
    weight = rescale(mean_expression, to = c(0.5, 2.2)),
    edge_type = "expression",
    module = module
  )

edges_marker_process <- network_df |>
  distinct(marker_node, process_node, module) |>
  transmute(
    from = marker_node,
    to = process_node,
    weight = 0.65,
    edge_type = "biology",
    module = module
  )

edges_process_disease <- network_df |>
  distinct(process_node, disease_node, module) |>
  transmute(
    from = process_node,
    to = disease_node,
    weight = 0.9,
    edge_type = "disease_link",
    module = module
  )

edges_disease_lace <- network_df |>
  distinct(disease_node, module) |>
  group_by(module) |>
  summarise(nodes = list(unique(disease_node)), .groups = "drop") |>
  rowwise() |>
  mutate(edge_tbl = list(
    if (length(nodes) >= 2) {
      as.data.frame(t(combn(nodes, 2)))
    } else {
      data.frame(V1 = character(), V2 = character())
    }
  )) |>
  ungroup() |>
  unnest(edge_tbl) |>
  transmute(
    from = V1,
    to = V2,
    weight = 0.18,
    edge_type = "disease_lace",
    module = module
  )

edges <- bind_rows(
  edges_module_cell,
  edges_cell_marker,
  edges_marker_process,
  edges_process_disease,
  edges_disease_lace
) |>
  distinct(from, to, module, edge_type, .keep_all = TRUE)

module_lookup <- bio_links |>
  select(cell_state, marker, process, disease, module) |>
  pivot_longer(
    cols = c(cell_state, marker, process, disease),
    names_to = "kind",
    values_to = "label"
  ) |>
  distinct(label, module)

nodes <- tibble(name = unique(c(edges$from, edges$to))) |>
  mutate(
    label = case_when(
      grepl("^Marker: ", name)  ~ sub("^Marker: ", "", name),
      grepl("^Process: ", name) ~ sub("^Process: ", "", name),
      grepl("^Disease: ", name) ~ sub("^Disease: ", "", name),
      grepl("^Module: ", name)  ~ sub("^Module: ", "", name),
      TRUE ~ name
    ),
    node_type = case_when(
      grepl("^Module: ", name)  ~ "Biological module",
      grepl("^Marker: ", name)  ~ "Marker",
      grepl("^Process: ", name) ~ "Biological process",
      grepl("^Disease: ", name) ~ "Disease / pathology",
      TRUE ~ "Cell state"
    )
  ) |>
  left_join(module_lookup, by = "label") |>
  mutate(
    module = case_when(
      node_type == "Biological module" ~ label,
      is.na(module) ~ "Other",
      TRUE ~ module
    ),
    node_size = case_when(
      node_type == "Biological module" ~ 18,
      node_type == "Disease / pathology" ~ 10,
      node_type == "Cell state" ~ 8,
      node_type == "Biological process" ~ 5.5,
      node_type == "Marker" ~ 3.2,
      TRUE ~ 3
    ),
    label_size = case_when(
      node_type == "Biological module" ~ 4.2,
      node_type == "Disease / pathology" ~ 3.0,
      node_type == "Cell state" ~ 2.8,
      node_type == "Biological process" ~ 2.2,
      node_type == "Marker" ~ 1.8,
      TRUE ~ 2
    ),
    show_label = case_when(
      node_type %in% c("Biological module", "Disease / pathology", "Cell state") ~ TRUE,
      node_type == "Biological process" ~ TRUE,
      node_type == "Marker" ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  distinct(name, .keep_all = TRUE)

graph_obj <- graph_from_data_frame(
  d = edges,
  vertices = nodes,
  directed = FALSE
)

set.seed(777)

layout_df <- create_layout(
  graph_obj,
  layout = "fr",
  niter = 12000,
  weights = E(graph_obj)$weight
)

module_colors <- c(
  "Tumor" = "#D84A9A",
  "Stroma" = "#00A6D6",
  "Myeloid" = "#00A878",
  "Lymphoid" = "#4F63B6",
  "Other" = "#BDBDBD"
)

p <- ggraph(layout_df) +
  geom_edge_arc(
    aes(
      width = weight,
      color = module,
      alpha = edge_type
    ),
    strength = 0.18,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_node_point(
    aes(
      size = node_size,
      fill = module
    ),
    shape = 21,
    color = "white",
    stroke = 0.55,
    alpha = 0.96
  ) +
  geom_node_text(
    data = layout_df |> filter(show_label),
    aes(
      label = label,
      size = label_size
    ),
    repel = TRUE,
    color = "grey10",
    max.overlaps = Inf,
    point.padding = 0.12,
    box.padding = 0.22,
    segment.alpha = 0.18,
    segment.size = 0.15
  ) +
  scale_fill_manual(values = module_colors) +
  scale_edge_color_manual(values = module_colors) +
  scale_size_identity() +
  scale_edge_width(range = c(0.05, 1.6)) +
  scale_edge_alpha_manual(
    values = c(
      "module_lace" = 0.08,
      "expression" = 0.32,
      "biology" = 0.20,
      "disease_link" = 0.26,
      "disease_lace" = 0.08
    )
  ) +
  coord_equal(clip = "off") +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 21, hjust = 0.5),
    plot.subtitle = element_text(size = 11.5, hjust = 0.5),
    plot.margin = ggplot2::margin(20, 30, 20, 30)
  ) +
  labs(
    title = "Cell state–marker–disease association network",
    subtitle = "Tumor, stromal, myeloid and lymphoid programs linked to disease-like biological pathology",
    fill = "Biological module"
  )

print(p)

if (!exists("fig_dir")) {
  fig_dir <- getwd()
}

ggsave(
  filename = file.path(fig_dir, "beautiful_cell_state_marker_disease_network.png"),
  plot = p,
  width = 12,
  height = 10,
  dpi = 320,
  bg = "white"
)
```
```{r}
## Disease association heatmap
## cell states vs disease-like pathology

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(scales)

available_markers <- intersect(
  unique(bio_links$marker),
  colnames(bio_df)
)

marker_means <- bio_df |>
  group_by(pred_label) |>
  summarise(
    across(
      all_of(available_markers),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = all_of(available_markers),
    names_to = "marker",
    values_to = "mean_expression"
  ) |>
  rename(cell_state = pred_label)

disease_activity <- bio_links |>
  filter(marker %in% available_markers) |>
  left_join(marker_means, by = c("cell_state", "marker")) |>
  filter(!is.na(mean_expression)) |>
  group_by(cell_state, disease, module) |>
  summarise(
    disease_score = mean(mean_expression, na.rm = TRUE),
    n_markers = n_distinct(marker),
    markers = paste(unique(marker), collapse = ", "),
    .groups = "drop"
  ) |>
  mutate(
    disease_score_scaled = rescale(disease_score, to = c(0, 1))
  )

p_heatmap <- ggplot(
  disease_activity,
  aes(
    x = cell_state,
    y = disease,
    fill = disease_score_scaled
  )
) +
  geom_tile(
    color = "white",
    linewidth = 0.6,
    alpha = 0.95
  ) +
  geom_point(
    aes(size = n_markers),
    shape = 21,
    fill = "white",
    color = "grey25",
    stroke = 0.4,
    alpha = 0.8
  ) +
  facet_grid(
    module ~ .,
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_gradient(
    low = "#F7FBFF",
    high = "#08306B",
    name = "Disease\nassociation\nscore"
  ) +
  scale_size_continuous(
    range = c(2, 6),
    name = "Number of\nmarkers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text.y = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  ) +
  labs(
    title = "Disease-like biological pathology across cell states",
    subtitle = "Heatmap shows marker-based disease association score for each cell state",
    x = "Cell state",
    y = "Disease / pathology"
  )

print(p_heatmap)

ggsave(
  filename = file.path(fig_dir, "disease_pathology_heatmap_cell_states.png"),
  plot = p_heatmap,
  width = 11,
  height = 8,
  dpi = 320,
  bg = "white"
)
```
```{r}
hallmark_links <- tibble::tribble(
  ~marker,      ~hallmark,

  # Tumor
  "Ki67",       "Proliferation",
  "EGFR",       "Growth signaling",
  "p53",        "DNA damage response",
  "panCK",      "Epithelial identity",
  "Ecad",       "Cell adhesion",

  # Stroma
  "SMA",        "ECM remodeling",
  "Podoplanin", "ECM remodeling",
  "MMP9",       "Invasion / EMT",
  "CD146",      "Angiogenesis",

  # Myeloid
  "CD11b",      "Myeloid infiltration",
  "CD204",      "Immunosuppression",
  "Arg1",       "Immunosuppression",
  "CXCL8",      "Inflammation",
  "CD66b",      "Inflammation",
  "CD15",       "Inflammation",
  "CD11c",      "Antigen presentation",

  # Lymphoid
  "CD3",        "T-cell immunity",
  "CD8a",       "Cytotoxicity",
  "PD1",        "Immune exhaustion",
  "FOXP3",      "Treg suppression",
  "GITR",       "Treg suppression",
  "NKG2A",      "Immune exhaustion",
  "CD56",       "NK cytotoxicity",
  "EOMES",      "NK differentiation",
  "Tbet",       "NK differentiation"
)
```


```{r}
hallmark_df <- marker_means |>
  inner_join(hallmark_links, by = "marker") |>
  group_by(cell_state, hallmark) |>
  summarise(
    score = mean(mean_expression, na.rm = TRUE),
    n_markers = n(),
    .groups = "drop"
  )
```


```{r}
hallmark_df <- hallmark_df |>
  group_by(hallmark) |>
  mutate(
    zscore = as.numeric(scale(score))
  ) |>
  ungroup()
```

```{r}
hallmark_order <- c(
  "Proliferation",
  "Growth signaling",
  "DNA damage response",
  "Epithelial identity",
  "Cell adhesion",
  "ECM remodeling",
  "Invasion / EMT",
  "Angiogenesis",
  "Myeloid infiltration",
  "Inflammation",
  "Immunosuppression",
  "Antigen presentation",
  "T-cell immunity",
  "Cytotoxicity",
  "Immune exhaustion",
  "Treg suppression",
  "NK cytotoxicity",
  "NK differentiation"
)

hallmark_df$hallmark <- factor(
  hallmark_df$hallmark,
  levels = rev(hallmark_order)
)
```

```{r}
marker_order <- marker_means |>
  group_by(marker) |>
  summarise(
    mean_signal = mean(mean_expression, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(mean_signal) |>
  pull(marker)

marker_df_plot <- marker_means |>
  group_by(marker) |>
  mutate(
    zscore = as.numeric(scale(mean_expression))
  ) |>
  ungroup()

marker_df_plot$marker <- factor(
  marker_df_plot$marker,
  levels = marker_order
)
```

```{r}
library(ggplot2)

p <- ggplot(
  marker_df_plot,
  aes(
    x = cell_state,
    y = marker,
    fill = zscore
  )
) +
  geom_tile(
    color = "white",
    linewidth = 0.6
  ) +
  geom_text(
    aes(label = round(mean_expression, 1)),
    size = 2.3
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Expression\n(Z-score)"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 40,
      hjust = 1
    ),
    plot.title = element_text(
      face = "bold",
      size = 13
    )
  ) +
  labs(
    title = "Marker expression across cell states",
    subtitle = "Computed directly from experimental data",
    x = "Cell state",
    y = "Marker"
  )

print(p)
```

```{r}
library(randomForest)

set.seed(42)

rf_model <- randomForest(
  x = as.data.frame(X),
  y = as.factor(y),
  ntree = 500,
  mtry = floor(sqrt(ncol(X))),
  importance = TRUE
)

saveRDS(
  rf_model,
  file = file.path(out_dir, "RandomForest_model.rds")
)

rf_model

```
```{r}
library(dplyr)
library(ggplot2)

rf_imp_df <- randomForest::importance(rf_model) |>
  as.data.frame() |>
  tibble::rownames_to_column("Marker") |>
  arrange(desc(MeanDecreaseGini)) |>
  slice_head(n = 20)

rf_imp_df <- rf_imp_df |>
  mutate(
    Module = case_when(
      Marker %in% c("Ecad","panCK","Ki67") ~ "Tumor",
      Marker %in% c("CD3","CD8a","CD45","CD7") ~ "Lymphoid",
      Marker %in% c("CD204","CD11c","CD11b","Arg1","CD66b") ~ "Myeloid",
      Marker %in% c("SMA","Podoplanin","CD146","CD31") ~ "Stroma",
      TRUE ~ "Other"
    )
  )

p_rf <- ggplot(
  rf_imp_df,
  aes(
    x = reorder(Marker, MeanDecreaseGini),
    y = MeanDecreaseGini,
    fill = Module
  )
) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      Tumor = "#E53935",
      Lymphoid = "#1E88E5",
      Myeloid = "#43A047",
      Stroma = "#FB8C00",
      Other = "#9E9E9E"
    )
  ) +
  labs(
    title = "Random Forest marker importance",
    subtitle = paste0(
      "Accuracy = ",
      round(cm_test$overall["Accuracy"],3),
      "   |   Kappa = ",
      round(cm_test$overall["Kappa"],3)
    ),
    x = NULL,
    y = "Mean Decrease Gini"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    legend.position = "right",
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold")
  )

print(p_rf)
```
```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

# ---------------------------
# Random Forest importance
# ---------------------------

rf_df <- randomForest::importance(rf_model) |>
  as.data.frame() |>
  rownames_to_column("Marker") |>
  select(Marker, MeanDecreaseGini) |>
  mutate(
    RF = MeanDecreaseGini / sum(MeanDecreaseGini)
  ) |>
  select(Marker, RF)

# ---------------------------
# XGBoost importance
# ---------------------------

xgb_df <- var_imp |>
  as.data.frame() |>
  transmute(
    Marker = Feature,
    XGBoost = Gain / sum(Gain)
  )

# ---------------------------
# Merge
# ---------------------------

imp_compare <- full_join(
  rf_df,
  xgb_df,
  by = "Marker"
) |>
  mutate(
    RF = ifelse(is.na(RF), 0, RF),
    XGBoost = ifelse(is.na(XGBoost), 0, XGBoost)
  )


top_markers <- imp_compare |>
  mutate(mean_imp = (RF + XGBoost) / 2) |>
  arrange(desc(mean_imp)) |>
  slice_head(n = 20) |>
  pull(Marker)

plot_df <- imp_compare |>
  filter(Marker %in% top_markers) |>
  pivot_longer(
    cols = c(RF, XGBoost),
    names_to = "Model",
    values_to = "Importance"
  )

# ---------------------------
# Plot
# ---------------------------

p_compare <- ggplot(
  plot_df,
  aes(
    x = Importance,
    y = reorder(Marker, Importance),
    fill = Model
  )
) +
  geom_col(
    position = "dodge",
    width = 0.8
  ) +
 scale_fill_manual(
  values = c(
    RF = "#D55E00",      
    XGBoost = "#0072B2"  
  )

  ) +
  labs(
    title = "Marker importance across two models",
    subtitle = paste0(
      "Random Forest (Accuracy = ",
      round(cm_test$overall["Accuracy"],3),
      ", Kappa = ",
      round(cm_test$overall["Kappa"],3),
      ")"
    ),
    x = "Normalized importance",
    y = "Marker"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    legend.title = element_blank()
  )

print(p_compare)
```





















