---
title: "XGBoost rare-cell tuning — results report"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
base_path <- "/Users/svitlanamoiseyenko/Repos/genomic/xgboost-imc-panel2-cell-classifier/XGBoost_classifier/Panel_2_10"
obj  <- readRDS(file.path(base_path, "ROut", "report_cms.rds"))
res  <- obj$results
cm_store    <- obj$cm_store
split_store <- obj$split_store
xgb_grid    <- obj$xgb_grid

critical <- c("MDSC", "BnT", "Igg", "mregDC", "PMN_MDSC", "Neutrophil")
res$is_model <- grepl("^M", res$model_id)
# stable display order: by model index then ablations
res <- res[order(!res$is_model, suppressWarnings(as.integer(sub("[MA]", "", res$model_id)))), ]
r3 <- function(x) round(as.numeric(x), 4)
```

# Summary

10 XGBoost parameter combinations (M1–M10) on a **fixed tier-aware image-level split**, plus 4
cleaning/weight **ablations** (A1–A4). Each is evaluated on the **held-out test images**. The dataset
is Tumor-dominated (`AccuracyNull ≈ 0.84`), so the report ranks on **critical-class recall**
(MDSC, BnT, Igg, mregDC, PMN_MDSC, Neutrophil), not raw accuracy. Confusion matrices below were
recomputed by retraining each model at its tuned `nrounds_best`; headline metrics come from the
tuning run (`tuning_results.csv`).

# 1. How to read the split params

The held-out test set is built at the **image level** (all cells of an image go to one side), tuned by tier:

- **`tier_test_frac = c(critical = 0.30, medium = 0.25, non_critical = 0.20)`** — share of each tier's
  cells forced into the test set. Critical types get **30 %** tested vs **20 %** for abundant types, so
  rare-class recall is measured on enough cells.
- **`tier_weight = c(critical = 40, medium = 10, non_critical = 1)`** — how hard the optimiser tries to
  hit that fraction. A critical-class mismatch is penalised **40×** harder than a common one.
- **`min_test_count = c(critical = 8, medium = 5, non_critical = 0)`** — a hard **floor**: ≥8 critical
  cells per type must land in the test set, otherwise its sensitivity is unestimable.

# 2. Achieved split (shared across M1–M10)

```{r split-table}
knitr::kable(split_store[["M1"]],
             caption = "Tier-aware split: per-class counts and achieved test fraction")
```

```{r split-totals}
cat(sprintf("Train: %d cells | Test: %d cells | %d test images.\n",
            res$n_train_cells[res$model_id == "M1"][1],
            res$n_test_cells[res$model_id == "M1"][1],
            res$n_test_images[res$model_id == "M1"][1]))
```

# 3. The 10-combination XGBoost grid

```{r grid-table}
knitr::kable(xgb_grid, caption = "Grid varied only XGBoost params; split + cleaning held constant")
```

# 4. Scenario results (sorted by critical-class recall)

```{r scenario-table}
tab <- res |>
  dplyr::arrange(dplyr::desc(Sensitivity_critical_mean_test)) |>
  dplyr::transmute(
    scenario_label,
    Accuracy_test  = r3(Accuracy_test),
    Kappa_test     = r3(Kappa_test),
    Sens_crit_test = r3(Sensitivity_critical_mean_test),
    F1_crit_test   = r3(F1_critical_mean_test),
    nrounds_best)
knitr::kable(tab, caption = "Held-out test metrics per scenario (best = top)")
```

# 5. Per-class critical sensitivity (recall) matrix

```{r sens-matrix}
sens_cols <- paste0("Sens_", critical, "_test")
m <- res[, c("scenario_label", sens_cols)]
colnames(m) <- c("scenario", critical)
m[critical] <- lapply(m[critical], r3)
m <- m[order(-rowMeans(sapply(m[critical], as.numeric), na.rm = TRUE)), ]
knitr::kable(m, row.names = FALSE,
             caption = "Per-class test recall for the 6 critical cell types (higher = better)")
```

# 6. Per-iteration detail

```{r iterations, results='asis'}
analyse_row <- function(r) {
  sv <- as.numeric(r[paste0("Sens_", critical, "_test")]); names(sv) <- critical
  missed <- names(sv)[!is.na(sv) & sv < 0.5]
  msgs <- c()
  if (length(missed)) msgs <- c(msgs, paste0("missed (recall<0.5): ", paste(missed, collapse = ", ")))
  if (r$Accuracy_test - r$AccuracyNull_test < 0.05)
    msgs <- c(msgs, "accuracy barely above the majority-Tumor null — leaning on the dominant class")
  if (r$Precision_critical_mean_test < r$Sensitivity_critical_mean_test - 0.1)
    msgs <- c(msgs, "critical classes over-predicted (precision < recall)")
  if (!length(msgs)) msgs <- "no critical class below recall 0.5; balanced precision/recall"
  paste(msgs, collapse = "; ")
}

ord <- order(-res$Sensitivity_critical_mean_test)
for (i in ord) {
  r <- res[i, ]
  cat(sprintf("\n\n## %s — %s\n\n", r$model_id, r$scenario_label))

  cat("**XGBoost params**\n\n")
  print(knitr::kable(data.frame(
    max_depth = r$max_depth, eta = r$eta, min_child_weight = r$min_child_weight,
    gamma = r$gamma, lambda = r$lambda, max_delta_step = r$max_delta_step,
    subsample = r$subsample, colsample_bytree = r$colsample_bytree,
    nrounds_best = r$nrounds_best, weight_scheme = r$weight_scheme, row.names = NULL)))

  cat("\n\n**Test metrics**\n\n")
  print(knitr::kable(data.frame(
    Accuracy = r3(r$Accuracy_test), Kappa = r3(r$Kappa_test),
    Lower = r3(r$AccuracyLower_test), Upper = r3(r$AccuracyUpper_test),
    Null = r3(r$AccuracyNull_test), Sens_crit = r3(r$Sensitivity_critical_mean_test),
    F1_crit = r3(r$F1_critical_mean_test), row.names = NULL)))

  cm <- cm_store[[r$model_id]]
  if (!is.null(cm)) {
    cat("\n\n**Confusion matrix** (rows = Prediction, cols = Reference)\n\n")
    print(knitr::kable(as.data.frame.matrix(cm)))
  }
  cat("\n\n**Analysis:** ", analyse_row(r), "\n\n")
}
```

# 7. Combined summary — models vs ablations

```{r combined}
res$group <- ifelse(res$is_model, "grid model", "ablation")
comb <- res |>
  dplyr::arrange(dplyr::desc(Sensitivity_critical_mean_test)) |>
  dplyr::transmute(group, scenario_label,
    Acc = r3(Accuracy_test), Kappa = r3(Kappa_test),
    Sens_crit = r3(Sensitivity_critical_mean_test), F1_crit = r3(F1_critical_mean_test),
    n_train = n_train_cells, n_test = n_test_cells, n_test_img = n_test_images)
knitr::kable(comb, caption = "All scenarios; train/test counts vary only where cleaning drops cells")
```

The **grid models** share one split (15 428 train / 11 608 test, 11 test images). Ablations A2/A3
change the cell set (drop undefined / area QC) so their split counts differ slightly. The capped-weight
ablation A1 keeps the same split as the grid.

# 8. Conclusion — best model

```{r conclusion, results='asis'}
acc_floor <- 0.93; kappa_floor <- 0.85
elig <- res[res$Accuracy_test >= acc_floor & res$Kappa_test >= kappa_floor, ]
elig <- elig[order(-elig$Sensitivity_critical_mean_test, -elig$F1_critical_mean_test,
                   -elig$Kappa_test), ]
best <- elig[1, ]
base <- res[res$model_id == "M1", ][1, ]

cat(sprintf(
"**Guardrail:** Accuracy_test ≥ %.2f and Kappa_test ≥ %.2f → %d/%d scenarios eligible (%s).\n\n",
acc_floor, kappa_floor, nrow(elig), nrow(res), paste(elig$model_id, collapse = ", ")))

cat(sprintf(
"**Selected model: %s — %s**\n\n- Held-out test: Accuracy = %.3f, Kappa = %.3f (null = %.3f)\n- Critical-class recall = %.3f (baseline M1 = %.3f), critical F1 = %.3f\n- Params: max_depth=%s, eta=%s, min_child_weight=%s, gamma=%s, lambda=%s, max_delta_step=%s, colsample=%s, nrounds=%s\n\n",
best$model_id, best$scenario_label,
best$Accuracy_test, best$Kappa_test, best$AccuracyNull_test,
best$Sensitivity_critical_mean_test, base$Sensitivity_critical_mean_test,
best$F1_critical_mean_test,
best$max_depth, best$eta, best$min_child_weight, best$gamma, best$lambda,
best$max_delta_step, best$colsample_bytree, best$nrounds_best))
```

```{r delta-vs-baseline}
sens_cols <- paste0("Sens_", critical, "_test")
delta <- data.frame(class = critical,
                    baseline_M1 = r3(base[sens_cols]),
                    best        = r3(best[sens_cols]))
delta$delta <- round(delta$best - delta$baseline_M1, 3)
knitr::kable(delta[order(delta$delta), ], row.names = FALSE,
             caption = "Per-critical-class recall: best model vs baseline M1")
```

**Takeaways**

- Ranking by critical recall (not accuracy) is essential: `AccuracyNull ≈ 0.84` means a Tumor-only
  classifier already "looks" ~84 % accurate, so accuracy hides rare-cell failures.
- Among guardrail-passing models, the winner recovers the most rare cells while staying near the
  accuracy/Kappa frontier.
- Any critical class with test recall < 0.5 (see §5) should be flagged "interpret with caution" —
  it needs more gated examples or better features rather than more tuning.
