---
title: "LightGBM vs XGBoost — Panel 2 IMC cell classifier"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
    self_contained: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
suppressMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(knitr); library(scales)
})
R <- readRDS("lgb_comparison.rds")
model_cols <- c(XGBoost = "#2166ac", `Random Forest` = "#9970ab",
                LightGBM = "#e67e22", RandomForest = "#9970ab")
```

## Summary

Three classifiers — **XGBoost** (the production model), **Random Forest**, and **LightGBM** —
were trained on the **same image-level train/test split** of the manually gated Panel 2 IMC
cells and evaluated on the held-out test set. LightGBM is added by the `kos.Rmd` fork; it shares
the exact feature matrix, label mapping, and inverse-frequency class weights used by XGBoost.

- **Train:** `r format(R$n_train, big.mark=",")` cells from `r R$n_train_images` images
- **Test:** `r format(R$n_test, big.mark=",")` labelled cells from `r R$n_test_images` images
- **Classes:** `r length(R$label_mapping)` cell types · **Features:** `r nrow(R$varimp)` markers (DNA/Histone excluded)

```{r headline}
R$metrics |>
  mutate(Accuracy = round(Accuracy, 4), Kappa = round(Kappa, 4),
         `Test mlogloss` = round(Test_mlogloss, 4)) |>
  select(Model, Accuracy, Kappa, `Test mlogloss`) |>
  arrange(desc(Accuracy)) |>
  kable(caption = "Held-out test performance (higher Accuracy/Kappa, lower mlogloss = better)")
```

**Verdict:** LightGBM edges out XGBoost on every metric (accuracy
`r round(R$metrics$Accuracy[R$metrics$Model=="LightGBM"],4)` vs
`r round(R$metrics$Accuracy[R$metrics$Model=="XGBoost"],4)`), but the gap is tiny
(< 0.1 pp). Both gradient-boosting models clearly beat Random Forest.

## Train / test split

The split is at the **image level** (every cell of an image goes to one side) using the
project's `split_test_rows()` optimizer, called with:

```{r split-params}
data.frame(
  parameter = c("default_test_frac", "rare_test_frac", "rare_labels",
                "rare_weight", "n_restarts", "seed"),
  value = c("0.20", "0.20", "B, BnT, MDSC, Igg, mregDC, NK", "20", "1000", "42")
) |> kable(caption = "split_test_rows() configuration (identical for all three models)")
```

```{r split-fig, fig.width=9, fig.height=4.5}
ss <- R$split_stats |>
  pivot_longer(c(train_n, test_n), names_to = "split", values_to = "n") |>
  mutate(split = recode(split, train_n = "train", test_n = "test"),
         cell_label = factor(cell_label, levels = R$split_stats$cell_label[order(-rowSums(R$split_stats[,-1]))]))
ggplot(ss, aes(cell_label, n, fill = split)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(train = "#4575b4", test = "#d73027")) +
  scale_y_log10(labels = comma) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "cells (log10)", fill = NULL,
       title = "Train / test cell counts per class")
```

```{r split-frac}
R$split_summary |>
  transmute(label, total, test_count, train_count,
            target_test_frac, achieved_test_frac, rare = is_rare) |>
  arrange(desc(total)) |>
  kable(caption = "Achieved test fraction per class vs target")
```

## How LightGBM was used

LightGBM is wired identically to XGBoost — same `X`, `label_mapping`, and the inverse-frequency
class weights (`undefined`/10, `Igg`/30) applied **per cell** via `lgb.Dataset(weight=...)`. The
test set is passed as the validation set (`reference = dtrain`) and drives **early stopping**.

```{r lgb-params}
lp <- R$lgb_params
data.frame(
  parameter = names(lp),
  value = sapply(lp, function(x) as.character(x))
) |> kable(caption = "LightGBM parameters (kos.Rmd)")
```

- **Objective:** `multiclass` (softmax over `r length(R$label_mapping)` classes), metric multiclass log-loss.
- **Boosting:** `nrounds = 500` with `early_stopping_rounds = 20` →
  **best iteration = `r R$lgb_best_iter`** (vs XGBoost's fixed 108 rounds at `max_depth = 3`).
- **Tree shape:** leaf-wise growth, `num_leaves = 31`, `max_depth = 4`, `learning_rate = 0.1`,
  feature/row subsampling 0.8, `lambda_l2 = 1.0`.
- **Note:** early stopping monitors the test set itself, so the reported LightGBM mlogloss is
  mildly optimistic relative to XGBoost (whose rounds were fixed by grouped CV).

## Final metrics & comparison

```{r acc-fig, fig.width=8, fig.height=4}
m <- R$metrics |> mutate(Model = factor(Model, levels = Model[order(Accuracy)]))
ggplot(m, aes(Accuracy, Model, fill = Model)) +
  geom_col(width = .65) +
  geom_text(aes(label = sprintf("%.4f", Accuracy)), hjust = 1.1, color = "white", fontface = "bold") +
  scale_fill_manual(values = model_cols, guide = "none") +
  coord_cartesian(xlim = c(0.97, 1)) +
  theme_bw(base_size = 12) +
  labs(x = "Test accuracy", y = NULL, title = "Overall accuracy")
```

```{r kappa-logloss-fig, fig.width=9, fig.height=3.6}
mm <- R$metrics |>
  select(Model, Kappa, Test_mlogloss) |>
  pivot_longer(-Model, names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, Test_mlogloss = "Test mlogloss"))
ggplot(mm, aes(value, Model, fill = Model)) +
  geom_col(width = .65) +
  geom_text(aes(label = sprintf("%.3f", value)), hjust = -0.1, size = 3.2) +
  scale_fill_manual(values = model_cols, guide = "none") +
  facet_wrap(~metric, scales = "free_x") +
  theme_bw(base_size = 11) +
  labs(x = NULL, y = NULL, title = "Cohen's kappa and test log-loss") +
  theme(plot.margin = margin(5, 18, 5, 5))
```

### Per-class sensitivity

Where the models differ is the **rare classes**. Below, recall (sensitivity) per cell type.

```{r byclass-fig, fig.width=10, fig.height=6}
bc <- R$byclass |>
  pivot_longer(-class, names_to = "Model", values_to = "Sensitivity") |>
  mutate(Sensitivity = ifelse(is.na(Sensitivity), 0, Sensitivity))
ord <- R$byclass |> mutate(m = rowMeans(cbind(XGBoost, RandomForest, LightGBM), na.rm = TRUE)) |>
  arrange(m) |> pull(class)
bc <- bc |> mutate(class = factor(class, levels = ord),
                   Model = recode(Model, RandomForest = "Random Forest"))
ggplot(bc, aes(Sensitivity, class, fill = Model)) +
  geom_col(position = position_dodge(width = .75), width = .7) +
  scale_fill_manual(values = model_cols) +
  theme_bw(base_size = 11) +
  labs(x = "Sensitivity (recall)", y = NULL, fill = NULL,
       title = "Per-class recall — XGBoost vs Random Forest vs LightGBM")
```

```{r byclass-tab}
R$byclass |>
  mutate(across(c(XGBoost, RandomForest, LightGBM), ~ round(.x, 3))) |>
  rename(`Random Forest` = RandomForest) |>
  arrange(class) |>
  kable(caption = "Per-class sensitivity")
```

### FPR vs TPR (XGBoost vs LightGBM)

```{r roc-fig, fig.width=8.5, fig.height=6}
mk_perf <- function(cm, model) {
  data.frame(cm$byClass) |>
    transmute(class = sub("Class: ", "", rownames(cm$byClass)),
              TPR = Sensitivity, FPR = 1 - Specificity, model = model)
}
perf <- bind_rows(mk_perf(R$cm_test, "XGBoost"), mk_perf(R$cm_lgb, "LightGBM")) |>
  filter(!is.na(TPR))
ggplot(perf, aes(FPR, TPR, color = model)) +
  geom_point(size = 2.4, alpha = .85) +
  ggrepel::geom_text_repel(aes(label = class), size = 2.7, max.overlaps = Inf,
                           show.legend = FALSE, segment.alpha = .4) +
  scale_color_manual(values = model_cols) +
  theme_bw(base_size = 11) +
  labs(x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)", color = NULL,
       title = "Per-class operating points")
```

### LightGBM confusion matrix

```{r cm-fig, fig.width=8.5, fig.height=7.5}
tab <- as.data.frame(R$cm_lgb$table)
ggplot(tab, aes(Reference, Prediction, fill = Freq)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = ifelse(Freq > 0, Freq, "")), size = 2.6) +
  scale_fill_gradient(low = "white", high = "#e67e22", trans = "log1p", guide = "none") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "LightGBM confusion matrix (held-out test)")
```

### Three-model consensus

```{r consensus}
cs <- R$consensus_stats
data.frame(
  metric = c("Full agreement (all 3 models)", "Majority agreement (>=2/3)",
             "Majority-vote accuracy", "High-entropy cells (top 10%)",
             "Full-agreement among high-entropy cells",
             "Full-agreement among low-entropy cells"),
  value = c(sprintf("%.2f%%", cs$full_agreement_pct),
            sprintf("%.2f%%", cs$majority_agree_pct),
            sprintf("%.4f", cs$majority_accuracy),
            cs$n_high_entropy,
            sprintf("%.1f%%", cs$high_entropy_full_agree_pct),
            sprintf("%.1f%%", cs$low_entropy_full_agree_pct))
) |> kable(caption = "XGBoost + RF + LightGBM consensus on the test set")
```

The three models fully agree on **`r sprintf("%.1f%%", cs$full_agreement_pct)`** of test cells.
Disagreement concentrates in high-entropy (uncertain) cells, confirming entropy as a useful
flag for cells needing review.

## Marker importance

```{r imp-fig, fig.width=9, fig.height=6.5}
vi <- R$varimp |>
  slice_max(mean_importance, n = 20) |>
  select(Feature, XGBoost, LightGBM, RandomForest) |>
  pivot_longer(-Feature, names_to = "Model", values_to = "Importance") |>
  mutate(Model = recode(Model, RandomForest = "Random Forest"),
         Feature = factor(Feature, levels = rev(R$varimp$Feature[1:20])))
ggplot(vi, aes(Importance, Feature, fill = Model)) +
  geom_col(position = position_dodge(width = .75), width = .7) +
  scale_fill_manual(values = model_cols) +
  theme_bw(base_size = 11) +
  labs(x = "Normalized importance", y = NULL, fill = NULL,
       title = "Top-20 markers by mean importance across models")
```

The top markers (CD3, CD204, CD209, CD20, Lamp3, CD31, FOXP3, …) agree with the expected
biology and are largely shared across all three models, so the LightGBM gain comes from better
fitting, not from a different feature basis.

## Conclusion

LightGBM is a **drop-in alternative that marginally outperforms XGBoost**
(accuracy `r round(R$metrics$Accuracy[R$metrics$Model=="LightGBM"],4)` /
kappa `r round(R$metrics$Kappa[R$metrics$Model=="LightGBM"],3)` vs
`r round(R$metrics$Accuracy[R$metrics$Model=="XGBoost"],4)` /
`r round(R$metrics$Kappa[R$metrics$Model=="XGBoost"],3)`), converging in only
`r R$lgb_best_iter` rounds, while Random Forest lags on rare classes. Given the < 0.1 pp gap and
that LightGBM's edge partly reflects early-stopping on the test set, the two boosters are
effectively equivalent here; their **`r sprintf("%.1f%%", cs$full_agreement_pct)` consensus**
makes a majority-vote ensemble the safest downstream choice.
