---
title: "08_systematic_tuning — results report"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Systematic improvement of the rare-cell XGBoost classifier

This script does **not** retrain the production model. It is an **experiment harness** that runs a
grid of *scenarios* (`split_cfg` × `clean_cfg` × `xgb_params`), evaluates each one with grouped
cross-validation **and** the held-out test images, and writes **one CSV row per experiment** with
the split parameters (6.1), the XGBoost parameters (6.2) and the evaluation metrics (6.3). The final
section (6.4) selects the best model by **critical-class recall behind an accuracy/Kappa guardrail**.

Goal: the dataset is dominated by Tumor (`AccuracyNull ≈ 0.74`), so plain accuracy hides failures on
the rare/critical populations (MDSC, BnT, Igg, mregDC, PMN_MDSC, Neutrophil). Everything here is
designed to surface and optimise those.

## Prerequisites

Run `01_read_data.Rmd` first **in the same R session**, at least through the cleaning chunk that
builds `lab_sce` and `dat`. This harness re-uses `lab_sce` (the gated, biologically-cleaned cells).
If you knit `08` in a fresh session, save the object once from `01`:

```r
# add to the end of 01 (optional), then it loads automatically below:
# saveRDS(lab_sce, file.path(out_dir, "lab_sce.rds"))
```

```{r load-libraries}
library(SingleCellExperiment)
library(caret)
library(xgboost)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
```

```{r config-paths}
base_path <- "/Users/svitlanamoiseyenko/Repos/genomic/xgboost-imc-panel2-cell-classifier/XGBoost_classifier/Panel_2_10"
out_dir   <- file.path(base_path, "ROut")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
```

```{r load-lab-sce}
# Re-use lab_sce from 01 (same session). Fall back to a saved copy if knitting fresh.
if (!exists("lab_sce")) {
  lab_rds <- file.path(out_dir, "lab_sce.rds")
  if (file.exists(lab_rds)) {
    lab_sce <- readRDS(lab_rds)
  } else {
    stop("`lab_sce` not found. Run 01_read_data.Rmd through the cleaning chunk first, ",
         "or save it with saveRDS(lab_sce, file.path(out_dir, 'lab_sce.rds')).")
  }
}

# Markers used as classifier features (DNA/Histone dropped, as in 01).
marker_keep_global <- !grepl("DNA|Histone", rownames(lab_sce), ignore.case = TRUE)

# Keep only confidently labelled cells.
lab_sce_base <- lab_sce[, !is.na(lab_sce$cell_labels) & lab_sce$cell_labels != "unlabelled"]
table(lab_sce_base$cell_labels)
```

---

# Part A — 3-tier criticality system

A single source of truth for cell-type priority. It drives both the train/test split (how much test
data each class is guaranteed) and, optionally, the class weights.

```{r criticality-tiers}
cell_priority <- list(
  critical     = c("MDSC", "BnT", "Igg", "mregDC", "PMN_MDSC", "Neutrophil"),
  medium       = c("B", "NK", "DC", "MacCD209", "Treg", "vCAF"),
  non_critical = c("Tumor", "CD4", "CD8", "Mac", "MacCD204", "Fibro",
                   "SMA", "Vasculature", "undefined")
)

priority_of <- function(lab) {
  dplyr::case_when(
    lab %in% cell_priority$critical ~ "critical",
    lab %in% cell_priority$medium   ~ "medium",
    TRUE                            ~ "non_critical"
  )
}

critical_classes <- cell_priority$critical
```

## Tier-aware image-level split

Generalises `split_test_rows()` from `01`. Differences:

* per-label test-fraction **target** and **weight** come from the label's tier (instead of a binary
  rare/common flag);
* a **minimum-test-count hinge penalty** pushes the optimiser to keep at least N cells of each
  critical class in the test set — the whole point is to be able to *estimate* rare-class recall.

The objective minimised over image subsets `S` (test side):

```
sum_l  w_l * (test_count_l / total_l - target_l)^2            # hit the per-tier test fraction
+ penalty_scale * sum_l  w_l * max(0, min_test_count_l - test_count_l)^2   # guarantee a floor in test
```

Optimisation is the same fast steepest-descent single-row-flip loop; with ~27 images each candidate
flip's full objective (penalty included) is evaluated in one vectorised matrix op per step.

```{r split-test-rows-tiered}
split_test_rows_tiered <- function(counts,
                                    cell_priority,
                                    tier_test_frac = c(critical = 0.30, medium = 0.25, non_critical = 0.20),
                                    tier_weight    = c(critical = 40,   medium = 10,   non_critical = 1),
                                    min_test_count = c(critical = 8,    medium = 5,    non_critical = 0),
                                    penalty_scale  = 1,
                                    n_restarts     = 1000,
                                    seed           = 42) {
  set.seed(seed)

  if (is.data.frame(counts)) counts <- as.matrix(counts)
  storage.mode(counts) <- "double"

  labels <- colnames(counts)
  totals <- colSums(counts)
  n      <- nrow(counts)

  tier        <- priority_of(labels)
  target_frac <- as.numeric(tier_test_frac[tier]); names(target_frac) <- labels
  min_tc      <- as.numeric(min_test_count[tier]); names(min_tc)      <- labels

  active     <- totals > 0
  w          <- as.numeric(tier_weight[tier]) * active
  inv_totals <- ifelse(active, 1 / pmax(totals, 1), 0)
  min_tc     <- min_tc * active

  score_of <- function(test_sum) {
    err <- test_sum * inv_totals - target_frac
    pen <- pmax(0, min_tc - test_sum)
    sum(w * err * err) + penalty_scale * sum(w * pen * pen)
  }

  # Bias rows that contain higher-priority cells toward starting in the test set.
  row_prob <- apply(counts, 1, function(r) {
    pres <- r > 0
    if (any(pres)) max(target_frac[pres]) else min(tier_test_frac)
  })

  best_score   <- Inf
  best_in_test <- logical(n)

  for (restart in seq_len(n_restarts)) {
    in_test   <- runif(n) < row_prob
    test_sum  <- colSums(counts[in_test, , drop = FALSE])
    cur_score <- score_of(test_sum)

    repeat {
      sgn  <- 1 - 2 * in_test                                   # +1 add to test, -1 remove from test
      # candidate test totals for flipping each single row: row i -> test_sum + sgn_i * counts[i, ]
      cand <- sweep(sgn * counts, 2, test_sum, FUN = "+")
      err  <- sweep(sweep(cand, 2, inv_totals, "*"), 2, target_frac, "-")
      pen  <- pmax(sweep(-cand, 2, min_tc, "+"), 0)            # matrix first: keep dim for %*%
      cand_score <- as.numeric((err * err) %*% w) +
                    penalty_scale * as.numeric((pen * pen) %*% w)

      best_i <- which.min(cand_score)
      if (cand_score[best_i] >= cur_score - 1e-12) break        # local minimum

      test_sum        <- test_sum + sgn[best_i] * counts[best_i, ]
      in_test[best_i] <- !in_test[best_i]
      cur_score       <- cand_score[best_i]
    }

    if (cur_score < best_score) {
      best_score   <- cur_score
      best_in_test <- in_test
    }
  }

  test_sum    <- colSums(counts[best_in_test, , drop = FALSE])
  train_sum   <- totals - test_sum
  achieved    <- ifelse(active, test_sum * inv_totals, NA_real_)

  summary_df <- data.frame(
    label              = labels,
    tier               = tier,
    total              = totals,
    test_count         = test_sum,
    train_count        = train_sum,
    target_test_frac   = target_frac,
    achieved_test_frac = round(achieved, 3),
    min_test_count     = min_tc,
    meets_min          = test_sum >= min_tc,
    row.names          = NULL
  )

  list(
    test_idx       = which(best_in_test),
    train_idx      = which(!best_in_test),
    test_rownames  = rownames(counts)[best_in_test],
    train_rownames = rownames(counts)[!best_in_test],
    summary        = summary_df,
    score          = best_score
  )
}
```

---

# Part B — Data cleaning (toggleable rules)

Existing `01` rules (low-IgG Igg dropped, low-CD7 NK dropped, mregDC→Tumor) are already baked into
`lab_sce`. The functions below add **new** cleaning that can be switched on/off per scenario so we
can measure its effect rather than assume it.

```{r cleaning-helpers}
# Robust area QC using steinbock regionprops `area` (segmentation-artifact removal).
area_vector <- function(sce) {
  cd <- colData(sce)
  nm <- grep("^area$", colnames(cd), ignore.case = TRUE, value = TRUE)
  if (length(nm) == 0) return(NULL)
  as.numeric(cd[[nm[1]]])
}

# Per-class marker-profile outliers: cells whose profile barely correlates with their
# class centroid (candidate mislabels). Used either for reporting or removal.
profile_outlier_cor <- function(sce, marker_keep, min_class_n = 30) {
  mat  <- as.matrix(assay(sce, "exprs")[marker_keep, , drop = FALSE])
  labs <- as.character(sce$cell_labels)
  cors <- rep(NA_real_, ncol(sce))
  for (cl in unique(labs)) {
    idx <- which(labs == cl)
    if (length(idx) < min_class_n) next
    centroid <- apply(mat[, idx, drop = FALSE], 1, median)
    cors[idx] <- apply(mat[, idx, drop = FALSE], 2, function(v) {
      suppressWarnings(stats::cor(v, centroid, method = "spearman"))
    })
  }
  cors
}

apply_cleaning <- function(sce, cfg, marker_keep) {
  if (isTRUE(cfg$drop_undefined)) {
    sce <- sce[, sce$cell_labels != "undefined"]
  }
  if (isTRUE(cfg$area_qc)) {
    area <- area_vector(sce)
    if (!is.null(area)) {
      la   <- log(pmax(area, 1))
      keep <- la >= (median(la) - 3 * mad(la)) & la <= (median(la) + 3 * mad(la))
      sce  <- sce[, keep]
    }
  }
  if (isTRUE(cfg$outlier_remove)) {
    cors <- profile_outlier_cor(sce, marker_keep)
    keep <- is.na(cors) | cors >= (cfg$outlier_min_cor %||% 0)
    sce  <- sce[, keep]
  }
  sce
}
```

## Outlier report (report-only, no removal)

Run this once to see candidate mislabels before deciding whether to enable `outlier_remove`.

```{r outlier-report}
cors0 <- profile_outlier_cor(lab_sce_base, marker_keep_global)
outlier_report <- data.frame(
  cell_label   = as.character(lab_sce_base$cell_labels),
  sample_id    = as.character(lab_sce_base$sample_id),
  centroid_cor = round(cors0, 3)
) |>
  dplyr::filter(!is.na(centroid_cor)) |>
  dplyr::group_by(cell_label) |>
  dplyr::summarise(
    n                = dplyr::n(),
    median_cor       = round(median(centroid_cor), 3),
    frac_below_0.2   = round(mean(centroid_cor < 0.2), 3),
    .groups = "drop"
  ) |>
  dplyr::arrange(median_cor)

outlier_report
```

---

# Part C/D — Helpers: feature matrix, folds, metrics, one experiment

```{r make-xy}
# Build XGBoost matrices + capped inverse-frequency weights. Factors out 01:966-1068.
make_xy <- function(train_sce, test_sce, marker_keep,
                    weight_cap       = c(0.2, 5),
                    weight_overrides = NULL) {

  train_sce$cell_id <- paste0(train_sce$sample_id, "_", train_sce$ObjectNumber)
  test_sce$cell_id  <- paste0(test_sce$sample_id,  "_", test_sce$ObjectNumber)
  colnames(train_sce) <- train_sce$cell_id
  colnames(test_sce)  <- test_sce$cell_id

  X       <- t(assay(train_sce, "exprs")[marker_keep, ])
  y_true  <- factor(train_sce$cell_labels)
  label_mapping <- levels(y_true)
  y       <- as.integer(y_true) - 1
  num_classes <- length(label_mapping)

  test_keep <- !is.na(test_sce$cell_labels) &
    test_sce$cell_labels != "unlabelled" &
    test_sce$cell_labels %in% label_mapping
  X_test      <- t(assay(test_sce, "exprs")[marker_keep, test_keep])
  y_test_true <- factor(test_sce$cell_labels[test_keep], levels = label_mapping)
  y_test      <- as.integer(y_test_true) - 1

  stopifnot(identical(colnames(X), colnames(X_test)))

  # Capped inverse-frequency class weights (no magic /10, /30 unless overridden).
  freq    <- as.numeric(table(y_true) / length(y_true))
  w_class <- 1 / freq
  names(w_class) <- levels(y_true)
  if (!is.null(weight_overrides)) {
    for (nm in names(weight_overrides)) {
      if (nm %in% names(w_class)) w_class[nm] <- w_class[nm] * weight_overrides[[nm]]
    }
  }
  w_class <- w_class / mean(w_class)
  w_class <- pmin(pmax(w_class, weight_cap[1]), weight_cap[2])
  w       <- as.numeric(w_class[as.character(y_true)])
  w       <- w / mean(w)

  list(
    Xy            = xgboost::xgb.DMatrix(data = X, label = y, weight = w),
    Xy_test       = xgboost::xgb.DMatrix(data = X_test, label = y_test),
    X = X, X_test = X_test,
    y_true = y_true, y_test_true = y_test_true,
    label_mapping = label_mapping, num_classes = num_classes
  )
}

# Image-grouped CV folds (validation indices), as in 01:1112-1122.
make_folds <- function(train_sce, k = 5) {
  tr_idx  <- caret::groupKFold(factor(train_sce$sample_id), k = k)
  all_idx <- seq_len(ncol(train_sce))
  lapply(tr_idx, function(idx) setdiff(all_idx, idx))
}

# Reshape an xgb.cv / predict() softprob output into a cells × classes matrix.
to_prob_matrix <- function(p, n, num_classes) {
  if (is.list(p)) p <- do.call(rbind, p)
  if (is.null(dim(p))) {
    byrow_m <- matrix(p, ncol = num_classes, byrow = TRUE)
    bycol_m <- matrix(p, ncol = num_classes, byrow = FALSE)
    p <- if (mean(abs(rowSums(byrow_m) - 1)) <= mean(abs(rowSums(bycol_m) - 1)))
      byrow_m else bycol_m
  }
  as.matrix(p)
}
```

```{r metrics-helper}
# Pull the exact cm$overall fields plus critical-class metrics. suffix = "cv" or "test".
extract_metrics <- function(pred, truth, label_mapping, critical, suffix) {
  cm <- caret::confusionMatrix(
    data      = factor(as.character(pred),  levels = label_mapping),
    reference = factor(as.character(truth), levels = label_mapping)
  )
  ov <- cm$overall
  bc <- as.data.frame(cm$byClass)
  rownames(bc) <- sub("Class: ", "", rownames(bc))
  crit <- intersect(critical, rownames(bc))

  out <- list(
    Accuracy       = unname(ov["Accuracy"]),
    Kappa          = unname(ov["Kappa"]),
    AccuracyLower  = unname(ov["AccuracyLower"]),
    AccuracyUpper  = unname(ov["AccuracyUpper"]),
    AccuracyNull   = unname(ov["AccuracyNull"]),
    AccuracyPValue = unname(ov["AccuracyPValue"]),
    McnemarPValue  = unname(ov["McnemarPValue"]),
    BalancedAccuracy_macro    = mean(bc[["Balanced Accuracy"]], na.rm = TRUE),
    Sensitivity_critical_mean = mean(bc[crit, "Sensitivity"],     na.rm = TRUE),
    Precision_critical_mean   = mean(bc[crit, "Pos Pred Value"],  na.rm = TRUE),
    F1_critical_mean          = mean(bc[crit, "F1"],              na.rm = TRUE)
  )
  for (cl in critical) {
    out[[paste0("Sens_", cl)]] <- if (cl %in% rownames(bc)) bc[cl, "Sensitivity"] else NA_real_
  }
  stats::setNames(out, paste0(names(out), "_", suffix))
}
```

```{r run-experiment}
# One scenario end-to-end -> single-row data.frame of params + metrics.
run_experiment <- function(scenario, lab_sce_base, marker_keep, critical) {

  # 1. clean
  lab_clean <- apply_cleaning(lab_sce_base, scenario$clean_cfg, marker_keep)

  # 2. tier-aware image-level split
  dat <- as.data.frame(unclass(table(lab_clean$sample_id, lab_clean$cell_labels)))
  sp  <- do.call(split_test_rows_tiered,
                 c(list(counts = dat, cell_priority = cell_priority), scenario$split_cfg))

  train_sce <- lab_clean[, lab_clean$sample_id %in% sp$train_rownames]
  test_sce  <- lab_clean[, lab_clean$sample_id %in% sp$test_rownames]

  # 3. feature matrices + weights
  d <- make_xy(train_sce, test_sce, marker_keep,
               weight_cap       = scenario$clean_cfg$weight_cap %||% c(0.2, 5),
               weight_overrides = scenario$clean_cfg$weight_overrides)

  # 4. CV folds + params
  folds  <- make_folds(train_sce)
  params <- c(
    list(booster = "gbtree", objective = "multi:softprob", eval_metric = "mlogloss",
         num_class = d$num_classes, tree_method = "hist",
         nthread = max(1, parallel::detectCores() - 1)),
    scenario$xgb
  )

  # 5. grouped CV with out-of-fold predictions
  cv <- xgboost::xgb.cv(
    params = params, data = d$Xy, folds = folds,
    nrounds = scenario$nrounds %||% 500L,
    early_stopping_rounds = 20, verbose = FALSE, prediction = TRUE
  )
  best_it <- which.min(cv$evaluation_log$test_mlogloss_mean)

  cv_prob <- to_prob_matrix(cv$pred %||% cv$cv_predict, length(d$y_true), d$num_classes)
  cv_pred <- d$label_mapping[max.col(cv_prob, ties.method = "first")]

  # 6. final model on all training cells, evaluate on held-out test images
  final_model <- xgboost::xgb.train(params = params, data = d$Xy,
                                    nrounds = best_it, verbose = FALSE)
  test_prob <- to_prob_matrix(predict(final_model, d$Xy_test),
                              length(d$y_test_true), d$num_classes)
  test_pred <- d$label_mapping[max.col(test_prob, ties.method = "first")]

  # 7. metrics (cv + test) + capture confusion matrix / split summary for the report
  m_cv   <- extract_metrics(cv_pred,   d$y_true,      d$label_mapping, critical, "cv")
  m_test <- extract_metrics(test_pred, d$y_test_true, d$label_mapping, critical, "test")

  if (exists("cm_store")) {
    cm_obj <- caret::confusionMatrix(
      factor(test_pred, levels = d$label_mapping), d$y_test_true)
    cm_store[[scenario$model_id]]    <<- cm_obj$table   # rows = Prediction, cols = Reference
    split_store[[scenario$model_id]] <<- sp$summary
  }

  # 8. assemble the row: 6.1 split params, 6.2 model params, 6.3 metrics
  scfg <- scenario$split_cfg
  split_params <- list(
    split_id                   = scenario$split_id,
    default_test_frac          = unname((scfg$tier_test_frac %||% c(non_critical = 0.20))["non_critical"]),
    tier_test_frac_critical    = unname((scfg$tier_test_frac %||% c(critical = 0.30))["critical"]),
    tier_test_frac_medium      = unname((scfg$tier_test_frac %||% c(medium = 0.25))["medium"]),
    tier_test_frac_noncrit     = unname((scfg$tier_test_frac %||% c(non_critical = 0.20))["non_critical"]),
    tier_weight_critical       = unname((scfg$tier_weight %||% c(critical = 40))["critical"]),
    tier_weight_medium         = unname((scfg$tier_weight %||% c(medium = 10))["medium"]),
    tier_weight_noncrit        = unname((scfg$tier_weight %||% c(non_critical = 1))["non_critical"]),
    min_test_count_critical    = unname((scfg$min_test_count %||% c(critical = 8))["critical"]),
    n_restarts                 = scfg$n_restarts %||% 1000,
    seed                       = scfg$seed %||% 42,
    n_train_cells              = ncol(train_sce),
    n_test_cells               = ncol(test_sce),
    n_test_images              = length(sp$test_rownames),
    min_test_count_observed    = min(sp$summary$test_count)
  )

  model_params <- list(
    model_id         = scenario$model_id,
    max_depth        = params$max_depth,
    eta              = params$eta,
    subsample        = params$subsample,
    colsample_bytree = params$colsample_bytree,
    min_child_weight = params$min_child_weight,
    gamma            = params$gamma %||% 0,
    lambda           = params$lambda %||% 1,
    alpha            = params$alpha %||% 0,
    max_delta_step   = params$max_delta_step %||% 0,
    nrounds_best     = best_it,
    drop_undefined   = isTRUE(scenario$clean_cfg$drop_undefined),
    weight_scheme    = scenario$clean_cfg$weight_scheme %||% "capped_invfreq"
  )

  as.data.frame(c(scenario_label = scenario$label,
                  split_params, model_params, m_cv, m_test),
                stringsAsFactors = FALSE, check.names = FALSE)
}
```

---

# Part D — Scenario grid (10 model combinations)

Baseline `clean_cfg` mirrors `01` (so row #1 reproduces the current setup), and the shared
tier-aware `split_cfg` guarantees a test floor for critical classes. The 10 rows vary **only the
XGBoost parameters** so the model effect is isolated. Ablation scenarios (cleaning / split changes)
are provided separately below.

```{r scenario-grid}
# Shared split + cleaning across the 10 model rows.
split_cfg_main <- list(
  tier_test_frac = c(critical = 0.30, medium = 0.25, non_critical = 0.20),
  tier_weight    = c(critical = 40,   medium = 10,   non_critical = 1),
  min_test_count = c(critical = 8,    medium = 5,    non_critical = 0),
  n_restarts     = 1000,
  seed           = 42
)

# Baseline cleaning = same as 01 (legacy weights, no new rules) for an honest baseline.
clean_cfg_baseline <- list(
  drop_undefined   = FALSE,
  area_qc          = FALSE,
  outlier_remove   = FALSE,
  weight_cap       = c(0, Inf),
  weight_overrides = list(undefined = 1 / 10, Igg = 1 / 30),
  weight_scheme    = "legacy_invfreq"
)

# 10-combination XGBoost grid (Part C of the plan).
xgb_grid <- tibble::tribble(
  ~max_depth, ~eta,  ~min_child_weight, ~gamma, ~lambda, ~max_delta_step, ~subsample, ~colsample_bytree, ~note,
  3,          0.10,  5,                 0,      1,       0,               0.8,        0.8,               "baseline (current best)",
  3,          0.10,  5,                 0,      1,       1,               0.8,        0.8,               "+max_delta_step",
  3,          0.05,  5,                 1,      1,       1,               0.8,        0.8,               "slower + gamma",
  4,          0.05,  5,                 1,      5,       1,               0.8,        0.8,               "+L2",
  4,          0.05,  1,                 0,      1,       1,               0.7,        0.6,               "more sampling",
  3,          0.03,  5,                 1,      5,       2,               0.8,        0.8,               "strong reg",
  4,          0.10,  1,                 0,      1,       0,               0.8,        0.8,               "flexible",
  5,          0.05,  5,                 1,      5,       1,               0.8,        0.8,               "depth probe",
  3,          0.05,  10,                1,      1,       1,               0.8,        0.6,               "conservative",
  4,          0.05,  5,                 1,      5,       1,               0.7,        0.8,               "balanced reg"
)

scenarios <- lapply(seq_len(nrow(xgb_grid)), function(i) {
  g <- xgb_grid[i, ]
  list(
    label     = paste0("model_", i, " (", g$note, ")"),
    split_id  = "split_main",
    model_id  = paste0("M", i),
    split_cfg = split_cfg_main,
    clean_cfg = clean_cfg_baseline,
    nrounds   = 500L,
    xgb = list(
      max_depth        = g$max_depth,
      eta              = g$eta,
      min_child_weight = g$min_child_weight,
      gamma            = g$gamma,
      lambda           = g$lambda,
      alpha            = 0,
      max_delta_step   = g$max_delta_step,
      subsample        = g$subsample,
      colsample_bytree = g$colsample_bytree
    )
  )
})

length(scenarios)
```

## Run the grid (heavy compute)

```{r run-grid, eval = TRUE}
cm_store    <- list()   # per-model test confusion matrices (filled inside run_experiment)
split_store <- list()   # per-model split summaries

results <- purrr::map_dfr(scenarios, function(sc) {
  message("Running ", sc$label)
  run_experiment(sc, lab_sce_base, marker_keep_global, critical_classes)
})

results_path <- file.path(out_dir, "tuning_results.csv")
write.csv(results, results_path, row.names = FALSE)
message("Wrote ", nrow(results), " rows to ", results_path)

results |>
  dplyr::select(scenario_label, Accuracy_test, Kappa_test,
                Sensitivity_critical_mean_test, F1_critical_mean_test) |>
  dplyr::arrange(dplyr::desc(Sensitivity_critical_mean_test))
```

## Optional ablation scenarios (cleaning / split / weights)

Run these to measure the impact of the new cleaning rules and weight scheme. They append rows to the
same CSV so everything is comparable. Uses the best XGBoost params found above.

```{r ablation-scenarios, eval = FALSE}
best_xgb <- scenarios[[ which.max(results$Sensitivity_critical_mean_test) ]]$xgb

ablations <- list(
  list(label = "ablation_capped_weights", split_id = "split_main", model_id = "A1",
       split_cfg = split_cfg_main, nrounds = 500L, xgb = best_xgb,
       clean_cfg = list(drop_undefined = FALSE, area_qc = FALSE, outlier_remove = FALSE,
                        weight_cap = c(0.2, 5), weight_overrides = NULL,
                        weight_scheme = "capped_invfreq")),
  list(label = "ablation_drop_undefined", split_id = "split_main", model_id = "A2",
       split_cfg = split_cfg_main, nrounds = 500L, xgb = best_xgb,
       clean_cfg = list(drop_undefined = TRUE, area_qc = FALSE, outlier_remove = FALSE,
                        weight_cap = c(0.2, 5), weight_overrides = NULL,
                        weight_scheme = "capped_invfreq")),
  list(label = "ablation_area_qc", split_id = "split_main", model_id = "A3",
       split_cfg = split_cfg_main, nrounds = 500L, xgb = best_xgb,
       clean_cfg = list(drop_undefined = FALSE, area_qc = TRUE, outlier_remove = FALSE,
                        weight_cap = c(0.2, 5), weight_overrides = NULL,
                        weight_scheme = "capped_invfreq")),
  list(label = "ablation_outlier_remove", split_id = "split_main", model_id = "A4",
       split_cfg = split_cfg_main, nrounds = 500L, xgb = best_xgb,
       clean_cfg = list(drop_undefined = FALSE, area_qc = FALSE, outlier_remove = TRUE,
                        outlier_min_cor = 0.1, weight_cap = c(0.2, 5),
                        weight_overrides = NULL, weight_scheme = "capped_invfreq"))
)

results_abl <- purrr::map_dfr(ablations, run_experiment,
                              lab_sce_base = lab_sce_base,
                              marker_keep  = marker_keep_global,
                              critical     = critical_classes)

results <- dplyr::bind_rows(results, results_abl)
write.csv(results, results_path, row.names = FALSE)
results_abl |>
  dplyr::select(scenario_label, Accuracy_test, Kappa_test,
                Sensitivity_critical_mean_test, F1_critical_mean_test)
```

---

# Part F — Result analysis & model selection (6.4)

Rule: **critical-class recall first, behind an accuracy/Kappa guardrail**. We never trade away
overall quality — only models that stay near the current performance are eligible, and among those
we pick the one that recovers the most rare cells.

```{r select-best}
acc_floor   <- 0.93   # guardrail: held-out test accuracy
kappa_floor <- 0.85   # guardrail: held-out test Kappa

eligible <- results |>
  dplyr::filter(Accuracy_test >= acc_floor, Kappa_test >= kappa_floor)

if (nrow(eligible) == 0) {
  warning("No model passed the guardrail; relax acc_floor/kappa_floor or inspect results.")
  eligible <- results
}

ranked <- eligible |>
  dplyr::arrange(
    dplyr::desc(Sensitivity_critical_mean_test),
    dplyr::desc(F1_critical_mean_test),
    dplyr::desc(Kappa_test)
  )

best_row <- ranked |> dplyr::slice(1)
best_row |>
  dplyr::select(scenario_label, model_id, Accuracy_test, Kappa_test,
                Sensitivity_critical_mean_test, F1_critical_mean_test, nrounds_best)
```

## Delta vs the baseline (model #1)

Which critical classes actually improved over the current setup.

```{r delta-vs-baseline}
baseline_row <- results |> dplyr::filter(model_id == "M1") |> dplyr::slice(1)

sens_cols <- grep("^Sens_.*_test$", names(results), value = TRUE)
delta <- data.frame(
  class        = sub("^Sens_(.*)_test$", "\\1", sens_cols),
  baseline     = round(as.numeric(baseline_row[sens_cols]), 3),
  best         = round(as.numeric(best_row[sens_cols]), 3)
)
delta$delta <- round(delta$best - delta$baseline, 3)
delta |> dplyr::arrange(delta)
```

## Per-cell-type TPR vs FPR for the winning model

Re-fits the selected scenario once to draw the diagnostic plot (reuses the `01` style).

```{r winner-plot, fig.height = 8, fig.width = 12}
best_scenario <- c(scenarios, if (exists("ablations")) ablations)[[
  which(vapply(c(scenarios, if (exists("ablations")) ablations),
               function(s) s$label, character(1)) == best_row$scenario_label)
]]

lab_clean <- apply_cleaning(lab_sce_base, best_scenario$clean_cfg, marker_keep_global)
dat       <- as.data.frame(unclass(table(lab_clean$sample_id, lab_clean$cell_labels)))
sp        <- do.call(split_test_rows_tiered,
                     c(list(counts = dat, cell_priority = cell_priority), best_scenario$split_cfg))
train_sce <- lab_clean[, lab_clean$sample_id %in% sp$train_rownames]
test_sce  <- lab_clean[, lab_clean$sample_id %in% sp$test_rownames]
d         <- make_xy(train_sce, test_sce, marker_keep_global,
                     weight_cap       = best_scenario$clean_cfg$weight_cap %||% c(0.2, 5),
                     weight_overrides = best_scenario$clean_cfg$weight_overrides)
params    <- c(list(booster = "gbtree", objective = "multi:softprob", eval_metric = "mlogloss",
                    num_class = d$num_classes, tree_method = "hist",
                    nthread = max(1, parallel::detectCores() - 1)), best_scenario$xgb)
fm        <- xgboost::xgb.train(params = params, data = d$Xy,
                                nrounds = best_row$nrounds_best, verbose = FALSE)
test_prob <- to_prob_matrix(predict(fm, d$Xy_test), length(d$y_test_true), d$num_classes)
test_pred <- factor(d$label_mapping[max.col(test_prob, ties.method = "first")],
                    levels = d$label_mapping)
cm_test   <- caret::confusionMatrix(test_pred, d$y_test_true)

perf <- data.frame(cm_test$byClass) |>
  dplyr::mutate(class = sub("Class: ", "", rownames(cm_test$byClass)),
                fpr   = 1 - Specificity,
                tier  = priority_of(class))

ggplot(perf) +
  geom_point(aes(x = fpr, y = Sensitivity, size = Detection.Rate, fill = tier),
             shape = 21, color = "black", alpha = 0.9) +
  ggrepel::geom_text_repel(aes(x = fpr, y = Sensitivity, label = class),
                           size = 3, max.overlaps = Inf, min.segment.length = 0) +
  scale_fill_manual(values = c(critical = "red3", medium = "orange2",
                               non_critical = "grey60")) +
  scale_size(range = c(1, 15)) +
  theme_bw() +
  labs(title = paste0("Held-out test performance: ", best_row$scenario_label),
       y = "Sensitivity (TPR)", x = "1 - Specificity (FPR)")
```

---

# REPORT

## How to read the split params

The held-out test set is built at the **image level** (all cells of an image go to one side), tuned by tier:

- **`tier_test_frac`** = `c(critical = 0.30, medium = 0.25, non_critical = 0.20)` — share of each tier's
  cells forced into the test set. Critical types (MDSC, BnT, Igg, mregDC, PMN_MDSC, Neutrophil) get **30%**
  tested vs **20%** for abundant types, so rare-class recall is measured on enough cells.
- **`tier_weight`** = `c(critical = 40, medium = 10, non_critical = 1)` — how hard the optimiser tries to
  hit that fraction. Critical mismatch is penalised **40×** harder than common.
- **`min_test_count`** = `c(critical = 8, medium = 5, non_critical = 0)` — a hard **floor**: ≥8 critical
  cells per type must land in test, otherwise its sensitivity is unestimable.

## Summary — all results (sorted by critical-class recall)

```{r report-summary}
summary_tab <- results |>
  dplyr::arrange(dplyr::desc(Sensitivity_critical_mean_test)) |>
  dplyr::transmute(
    scenario_label,
    Accuracy_test  = round(Accuracy_test, 4),
    Kappa_test     = round(Kappa_test, 4),
    Sens_crit_test = round(Sensitivity_critical_mean_test, 4),
    F1_crit_test   = round(F1_critical_mean_test, 4),
    nrounds_best
  )
knitr::kable(summary_tab, caption = "Test metrics per scenario (best = top)")
```

Achieved per-class test/train counts (shared split, identical across the 10 model rows):

```{r report-split-table}
knitr::kable(split_store[["M1"]], caption = "Tier-aware split: per-class counts and achieved test fraction")
```

## Per-iteration detail

```{r report-iterations, results='asis'}
analyse_row <- function(r) {
  sens_cols <- grep("^Sens_.*_test$", names(r), value = TRUE)
  sv  <- as.numeric(r[sens_cols]); names(sv) <- sub("^Sens_(.*)_test$", "\\1", sens_cols)
  missed <- names(sv)[!is.na(sv) & sv < 0.5]
  msgs <- c()
  if (length(missed)) msgs <- c(msgs, paste0("missed (recall<0.5): ", paste(missed, collapse = ", ")))
  if (!is.na(r$Accuracy_test) && !is.na(r$AccuracyNull_test) &&
      r$Accuracy_test - r$AccuracyNull_test < 0.05)
    msgs <- c(msgs, "accuracy barely above majority-Tumor null — leaning on the dominant class")
  if (!is.na(r$Precision_critical_mean_test) && !is.na(r$Sensitivity_critical_mean_test) &&
      r$Precision_critical_mean_test < r$Sensitivity_critical_mean_test - 0.1)
    msgs <- c(msgs, "critical classes over-predicted (precision < recall)")
  if (!length(msgs)) msgs <- "no critical class below recall 0.5; balanced precision/recall"
  paste(msgs, collapse = "; ")
}

ord <- order(-results$Sensitivity_critical_mean_test)
for (i in ord) {
  r <- results[i, ]
  cat(sprintf("\n\n### Iteration %s — %s\n\n", r$model_id, r$scenario_label))

  cat("**Split (test fraction targets / weights / floor)**\n\n")
  print(knitr::kable(data.frame(
    tier      = c("critical", "medium", "non_critical"),
    test_frac = c(r$tier_test_frac_critical, r$tier_test_frac_medium, r$tier_test_frac_noncrit),
    weight    = c(r$tier_weight_critical, r$tier_weight_medium, r$tier_weight_noncrit),
    min_test  = c(r$min_test_count_critical, NA, NA)
  )))
  cat(sprintf("\n\nTrain %d cells / Test %d cells across %d test images.\n\n",
              r$n_train_cells, r$n_test_cells, r$n_test_images))

  cat("\n**XGBoost params**\n\n")
  print(knitr::kable(r |> dplyr::select(max_depth, eta, min_child_weight, gamma, lambda,
                                        max_delta_step, subsample, colsample_bytree, nrounds_best)))

  cat("\n**Test metrics**\n\n")
  print(knitr::kable(r |> dplyr::transmute(
    Accuracy = round(Accuracy_test, 4), Kappa = round(Kappa_test, 4),
    Lower = round(AccuracyLower_test, 4), Upper = round(AccuracyUpper_test, 4),
    Null = round(AccuracyNull_test, 4),
    Sens_crit = round(Sensitivity_critical_mean_test, 4),
    F1_crit = round(F1_critical_mean_test, 4))))

  cm <- cm_store[[r$model_id]]
  if (!is.null(cm)) {
    cat("\n**Confusion matrix** (rows = Prediction, cols = Reference)\n\n")
    print(knitr::kable(as.data.frame.matrix(cm)))
  }

  cat("\n\n**Analysis:** ", analyse_row(r), "\n\n")
}
```

## Conclusions

```{r conclusions}
cat(sprintf(
"Selected model: %s
  Held-out test : Accuracy = %.3f  Kappa = %.3f  (null = %.3f)
  Critical recall (mean over %s): %.3f   |  baseline M1: %.3f
  Guardrail     : Accuracy >= %.2f and Kappa >= %.2f  -> %d/%d models eligible

Interpretation:
  - Ranking by critical-class recall (not raw accuracy) targets the rare-cell mission,
    because AccuracyNull (~%.2f) means a Tumor-only classifier already 'looks' good.
  - Inspect the delta table: classes with delta <= 0 still need more gates or features.
  - Any critical class with test sensitivity < 0.5 should be flagged 'interpret with caution'.
",
  best_row$scenario_label,
  best_row$Accuracy_test, best_row$Kappa_test, best_row$AccuracyNull_test,
  paste(critical_classes, collapse = ", "),
  best_row$Sensitivity_critical_mean_test,
  baseline_row$Sensitivity_critical_mean_test,
  acc_floor, kappa_floor, nrow(eligible), nrow(results),
  best_row$AccuracyNull_test
))
```
