---
title: "01_read_data"
---

link to presentation

https://docs.google.com/presentation/d/1j092IUsKge5XvswE18HGZPmz4NGRUtOXNunP9SOr3i0/edit?usp=sharing

```{r setup, include=FALSE}
# NOTE: глобальні налаштування knitr — показуємо код (echo), ховаємо повідомлення/попередження.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r shell}
# NOTE: що тут трапляється — команди steinbock CLI (запускаються в терміналі
# ДО будь-якого R), які з вихідних IMC-знімків генерують усі вхідні теки для аналізу.
# steinbock preprocess imc images --hpf 50
# steinbock segment deepcell --minmax
# steinbock measure intensities
# steinbock measure regionprops
# steinbock measure neighbors --type centroids --dmax 15

# NOTE: звідки беруться дані (всі теки лежать у base_path = Panel_2_10/):
#   raw/          — вихідні IMC-знімки (.mcd/.txt) з мас-цитометра (вхід для steinbock)
#   img/          — багатоканальні TIFF, з `steinbock preprocess imc images --hpf 50` з raw/
#   masks/        — маски сегментації клітин, з `steinbock segment deepcell --minmax`
#   intensities/  — середні інтенсивності маркерів на клітину, з `steinbock measure intensities`
#   regionprops/  — морфометрія клітин (площа, форма…), з `steinbock measure regionprops`
#   neighbors/    — граф сусідства клітин, з `steinbock measure neighbors --type centroids --dmax 15`
#   panel.csv, images.csv — панель з 44 маркерів і метадані знімків (надаються вручну)
#   gates/*.rds   — ручне гейтування (мітки типів клітин), експорт з cytomapperShiny - Q: do we need to add more gates?
#   *_compensation/ — TIFF-зрізи одиничних барвників для оцінки spillover (CATALYST)
#   ROut/         — вихід R: збережений spe.rds, модель тощо
```


```{r load-libraries, message=FALSE}
# NOTE: підключаємо всі бібліотеки Bioconductor/CRAN, потрібні для пайплайну
# (робота з SPE, компенсація CATALYST, гейти cytomapper, навчання XGBoost, графіки).
library(SingleCellExperiment)
library(caret)
library(xgboost)
library(imcRtools)
library(cytomapper)
library(dplyr)
library(dittoSeq)
library(ggplot2)
library(scater)
library(scuttle)
library(CATALYST)
library(patchwork)
library(stringr)
library(tidyverse)
library(scran)
library(viridis)
library(BiocParallel)
library(doParallel)
library(BiocNeighbors)
library(cowplot)
library(batchelor)
library(ComplexHeatmap)
library(circlize)
library(ggridges)
```


```{r read-spe}
# =====================================================================
# STEP 01 — Read images (steinbock outputs)
# =====================================================================
# NOTE: читаємо вихід steinbock і будуємо головний об'єкт SPE (SpatialExperiment):
# рядки = маркери IMC, колонки = клітини. Також задаємо out_dir для збереження результатів.
# NOTE: дані з intensities/, regionprops/, neighbors/, panel.csv, images.csv (вихід steinbock).
# NOTE: base_path має вказувати на реальну теку Panel_2_10 з даними steinbock.
#       Зараз стоїть ~/Desktop/...; за потреби заміни на фактичний шлях до даних.

base_path <- "/Users/svitlanamoiseyenko/Repos/genomic/xgboost-imc-panel2-cell-classifier/XGBoost_classifier/Panel_2_10"

raw_exprs_cofactor <- 5

out_dir <- file.path(base_path, "ROut")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

spe <- read_steinbock(base_path, return_as = "spe")

rownames(rowData(spe))
```

# spillover

```{r read-compensated-tiff}
# =====================================================================
# STEP 02 — Compensate (spillover correction)
# =====================================================================
# NOTE:
# Spillover в IMC (Imaging Mass Cytometry) — це коли сигнал одного металу "протікає" в інший канал.
# читаємо TIFF-зрізи одиничних барвників (compensation slide) у окремий об'єкт sceTIFF —
# вони потрібні, щоб оцінити spillover-матрицю між каналами металів.
# NOTE: дані з *_compensation/ (підтеки img/, та images.csv, panel.csv цієї теки).
comp_path <- list.files(base_path, pattern = "compensation$", full.names = TRUE)

img_path <- file.path(comp_path, "img")
image_csv <- file.path(comp_path, "images.csv")
panel_csv <- file.path(comp_path, "panel.csv")

sceTIFF <- readSCEfromTIFF(
  x = img_path,
  image_df_path = image_csv,
  panel_df_path = panel_csv
)

sceTIFF$Panel <- basename(base_path)
sceTIFF
```
```{r}
#TODO: visualise ....
```

# comp assay

```{r process-compensated-sce}
# NOTE: CATALYST-обробка compensation-зрізів: asinh-трансформація, дебаркодинг пікселів
# (assignPrelim/estCutoffs/applyCutoffs), фільтр пікселів і обчислення spillover-матриці.
compensation_pixel_exprs_cofactor <- 5
compensated_exprs_cofactor <- 1

process_comp_sce <- function(x, minevents = 40) {
  assay(x, "exprs") <- asinh(counts(x) / compensation_pixel_exprs_cofactor)

  bc_key <- as.numeric(unique(x$sample_mass))
  bc_key <- bc_key[order(bc_key)]

  x <- assignPrelim(x, bc_key = bc_key)
  x <- estCutoffs(x)
  x <- applyCutoffs(x)

  x <- filterPixels(
    x,
    minevents = minevents,
    correct_pixels = TRUE
  )

  x <- computeSpillmat(x)

  x
}

sceTIFF <- process_comp_sce(sceTIFF)
sceTIFF
```

```{r add-Ar}
# NOTE: правимо першу назву каналу в spillover-матриці на "Ar80Di" і додаємо аргон (Ar=80)
# до списку ізотопів, щоб канал коректно розпізнавався далі.
spillover_matrix <- metadata(sceTIFF)$spillover_matrix
colnames(spillover_matrix)[1] <- "Ar80Di"
metadata(sceTIFF)$spillover_matrix <- spillover_matrix

isotope_list <- CATALYST::isotope_list
isotope_list$Ar <- 80
```

```{r, fig.height=5, fig.width=5}
# NOTE: візуалізуємо spillover-матрицю (теплова карта перетікання сигналу між каналами).
# NOTE:Як читати графік?
# Рядки = канал-джерело сигналу.

# Стовпці = канал, куди сигнал "протікає".
# Чим червоніше — тим сильніший spillover.
# Число в клітинці = відсоток spillover.

plotSpillmat(sceTIFF, isotope_list = isotope_list)
```

```{r}
# NOTE: застосовуємо spillover-компенсацію до головного SPE через compCytof():
# перезаписуємо assay-и counts/exprs компенсованими значеннями і зберігаємо spe.rds у ROut/.
sm <- metadata(sceTIFF)$spillover_matrix
colnames(sm)[1] <- "Ar80Di"

# write.csv(sm, file.path(out_dir, "sm.csv"))

rowData(spe)$channel_name <- paste0(rowData(spe)$channel, "Di")

spe <- compCytof(
  spe,
  sm,
  transform = TRUE,
  cofactor = compensated_exprs_cofactor,
  isotope_list = isotope_list,
  overwrite = FALSE
)

# Keep an uncompensated copy for before/after spillover QC.
# counts is still the raw steinbock signal at this point; build a matching
# asinh exprs assay on the same cofactor so the two scatters are comparable.
spe_uncomp <- spe
assay(spe_uncomp, "exprs") <- asinh(assay(spe_uncomp, "counts") / compensated_exprs_cofactor)

assay(spe, "counts") <- assay(spe, "compcounts")
assay(spe, "exprs") <- assay(spe, "compexprs")

assay(spe, "compcounts") <- NULL
assay(spe, "compexprs") <- NULL
# already with comp assay
spe
# TODO: check  ROut -> Rout
# saveRDS(spe, file.path(base_path, "ROut", "spe.rds"))
saveRDS(spe, file.path(base_path, "Rout", "spe.rds"))
```



```{r, fig.height=5, fig.width=5}
# ADDED: but need to think more
# NOTE: QC компенсації — пара каналів з найсильнішим spillover, до vs після компенсації.
# Restrict to channels present in spe, find the strongest spillover pair, map to markers.
valid_ch <- intersect(rownames(sm), rowData(spe)$channel_name)
sm_sub   <- sm[valid_ch, valid_ch]
diag(sm_sub) <- 0

ij      <- which(sm_sub == max(sm_sub, na.rm = TRUE), arr.ind = TRUE)[1, ]
src_ch  <- rownames(sm_sub)[ij[1]]   # source channel (bleeds out)
tgt_ch  <- colnames(sm_sub)[ij[2]]   # target channel (receives bleed)

ch2marker  <- setNames(rownames(spe), rowData(spe)$channel_name)
src_marker <- ch2marker[[src_ch]]
tgt_marker <- ch2marker[[tgt_ch]]

p_before <- dittoSeq::dittoScatterPlot(
  spe_uncomp, x.var = src_marker, y.var = tgt_marker,
  assay.x = "exprs", assay.y = "exprs"
) + ggtitle(paste0("Before: ", src_marker, " vs ", tgt_marker))

p_after <- dittoSeq::dittoScatterPlot(
  spe, x.var = src_marker, y.var = tgt_marker,
  assay.x = "exprs", assay.y = "exprs"
) + ggtitle("After compensation")

p_before + p_after   # patchwork (already loaded)
```


```{r, fig.height=5, fig.width=5}
# NOTE: візуалізуємо spillover-матрицю (теплова карта перетікання сигналу між каналами). - AFTER
# TODO: check if visualisation is correct taken
# NOTE: seems we don't have some significant  difference after
 

plotSpillmat(sceTIFF, isotope_list = isotope_list)
```


Now we rewrote the assays in our spe to compensated



## already gated (labelled) images

USL - the group of colorectal cancer (CRC) tissue samples/patients that came from University Saint-Luc (Brussels)



```{r}
# =====================================================================
# STEP 04 — Label (manual gates)
# =====================================================================
# TODO: probably we can generate more gates
# NOTE: читаємо ручні гейти (.rds), нормалізуємо назви міток (прибираємо суфікси _NN,
# Cd4->CD4, MDSC_->MDSC, Cd8->CD8) і матчимо до клітин SPE через cell_id = sample_id_ObjectNumber.
# NOTE: дані з gates/*.rds (ручне гейтування, експорт з cytomapperShiny).
# spe <- readRDS(file.path(base_path, "ROut", "spe.rds"))
spe <- readRDS(file.path(base_path, "Rout", "spe.rds"))

gates <- list.files(
  file.path(base_path, "gates"),
  pattern = "\\.rds$",
  full.names = TRUE,
  recursive = TRUE
)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)

  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    cytomapper_CellLabel = as.character(x$cytomapper_CellLabel),
    gate_file = basename(path),
    row.names = NULL
  )
})

gate_labels$cytomapper_CellLabel <- sub(
  "_[0-9]+[a-zA-Z]?$",
  "",
  gate_labels$cytomapper_CellLabel
)

gate_labels$cytomapper_CellLabel <- sub(
  " \\([0-9]+\\)$",
  "",
  gate_labels$cytomapper_CellLabel
)

gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "Cd4"] <- "CD4"
gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "MDSC_"] <- "MDSC"
gate_labels$cytomapper_CellLabel[gate_labels$cytomapper_CellLabel == "Cd8"] <- "CD8"


gate_labels$cell_id <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)

sum(gate_labels$cell_id %in% colnames(spe))
nrow(gate_labels)

```


# img, masks

```{r}
# =====================================================================
# STEP 03 — Segment (cells + masks)
# =====================================================================
# NOTE: відбираємо лише гейтовані клітини, що є в SPE, визначаємо 28 гейтованих знімків
# і завантажуємо для них багатоканальні зображення та маски сегментації.
# NOTE: дані з img/ (зображення) і masks/ (маски сегментації).
matched_gate_labels <- gate_labels[
  gate_labels$cell_id %in% colnames(spe),
]

images_28 <- sort(unique(matched_gate_labels$sample_id))

images_28_pattern <- paste0(
  "^",
  images_28,
  "[.]tiff$",
  collapse = "|"
)

images_28_obj <- loadImages(
  file.path(base_path, "img"),
  pattern = images_28_pattern
)

images_28_obj <- images_28_obj[images_28]
channelNames(images_28_obj) <- rownames(spe)

masks_28 <- loadImages(
  file.path(base_path, "masks"),
  pattern = images_28_pattern,
  as.is = TRUE
)

masks_28 <- masks_28[names(images_28_obj)]

S4Vectors::mcols(images_28_obj) <- S4Vectors::mcols(masks_28) <-
  S4Vectors::DataFrame(sample_id = names(images_28_obj))

spe_28 <- spe[, spe$sample_id %in% names(images_28_obj)]

images_28
```

# Img QC

```{r heatmap, fig.height=10, fig.width=10}
# NOTE: QC на рівні знімків — рахуємо середню експресію маркерів по кожному знімку
# і малюємо теплову карту, щоб виявити порожні/аномальні зображення.
image_mean <- aggregateAcrossCells(spe_28,
                                   ids = spe_28$sample_id, 
                                   statistics="mean",
                                   use.assay.type = "counts")

assay(image_mean, "exprs") <- asinh(assay(image_mean, "counts"))

genes_to_plot <- rownames(image_mean)[
  rowSums(assay(image_mean, "exprs") != 0, na.rm = TRUE) > 0
]
if (length(genes_to_plot) == 0) {
  stop("No selected markers have non-zero image-level expression values.")
}

dittoHeatmap(
  image_mean, 
  genes = genes_to_plot,
  assay = "exprs", 
  cluster_cols = FALSE, 
  scale = "none",
  heatmap.colors = colorRampPalette(c(
    "white",
    "#B8CAE2", "#9EB7D8", "#4575B4", "#325584", "#264164", "#182A40"
  ))(100), 
  show_colnames = TRUE
)

```
delete 1 blank image from analysis

```{r}
# NOTE: data clean
# NOTE: прибираємо 1 порожній знімок з аналізу (28 -> 27 знімків) у SPE, зображеннях, масках і гейтах.
spe_27 <- spe_28[
  ,
  spe_28$sample_id != "10123883-USL-VAR-TIS-01-IMC-02_005"
]

images_27 <- setdiff(
  names(images_28_obj),
  "10123883-USL-VAR-TIS-01-IMC-02_005"
)

images_27_obj <- images_28_obj[images_27]
masks_27 <- masks_28[images_27]

gate_labels <- gate_labels[
  gate_labels$sample_id != "10123883-USL-VAR-TIS-01-IMC-02_005",
]
```


## gating via cymoviewer

```{r, message=FALSE}
# NOTE: перевіряємо, що id зображень, масок і SPE повністю збігаються, і записуємо
# узгоджені mcols (sample_id) для зображень та масок — підготовка до cytomapperShiny.
image_ids <- names(images_27_obj)
mask_ids <- names(masks_27)
spe_ids <- unique(as.character(spe_27$sample_id))

stopifnot(
  identical(image_ids, mask_ids),
  setequal(image_ids, spe_ids),
  !anyDuplicated(image_ids)
)

cur_mcols <- S4Vectors::DataFrame(
  image = image_ids,
  sample_id = image_ids,
  name = image_ids,
  row.names = image_ids
)

mcols(images_27_obj) <- cur_mcols
mcols(masks_27)      <- cur_mcols


identical(mcols(images_27_obj), mcols(masks_27))
identical(mcols(images_27_obj)$sample_id, names(masks_27))
setequal(
  mcols(images_27_obj)$sample_id,
  unique(as.character(spe_27$sample_id))
)
```

```{r}
# NOTE: інтерактивне гейтування/перегляд клітин у cytomapperShiny (лише в інтерактивній сесії RStudio).
if (interactive()) {
  cytomapperShiny(object = spe_27, mask = masks_27, image = images_27_obj,
                  cell_id = "ObjectNumber", img_id = "sample_id")
}
```


## consolidating labels

## cells assigned to the same class multiple times

```{r}
# NOTE: будуємо матрицю клітина x мітка і шукаємо клітини, яким одну й ту саму мітку
# присвоїли кілька разів (same-label duplicates).
cur_tab <- unclass(table( #make a mat no table class
  gate_labels$cell_id,
  gate_labels$cytomapper_CellLabel
))

cur_tab_raw <- cur_tab

same_label_duplicate_ids <- rownames(cur_tab_raw)[rowSums(cur_tab_raw > 1) > 0]

if (length(same_label_duplicate_ids) > 0) {
  cur_tab_raw[same_label_duplicate_ids, , drop = FALSE]
} else {
  print("No same-label duplicated gates found.")
}
```

## same label assigment

```{r}
# NOTE: зводимо лічильники до 0/1 і знаходимо справжні doublets — клітини з кількома РІЗНИМИ мітками.
cur_tab <- cur_tab_raw
cur_tab[cur_tab > 1] <- 1

# True multi-label doublets
multi_label_doublet_ids <- rownames(cur_tab)[rowSums(cur_tab) > 1]

if (length(multi_label_doublet_ids) > 0) {
  head(cur_tab[multi_label_doublet_ids, , drop = FALSE],5)
} else {
  message("No multi-label doublets found.")
}

```

# Consolidate labels

```{r}
# NOTE: консолідуємо мітки: 1 мітка -> залишаємо; "Tumor + інше" -> залишаємо НЕ-Tumor мітку;
# решту справжніх doublets -> видаляємо. Наприкінці зводимо статистику консолідації.
#assign all cell_ids as doublets
cur_labels <- rep("doublets", nrow(cur_tab))
names(cur_labels) <- rownames(cur_tab) #cell IDs as names,

single_index <- rowSums(cur_tab) == 1 #finds cells that have exactly one label

# cells with exactly 1 label, this finds which column = 1
cur_labels[single_index] <- colnames(cur_tab)[
  apply(cur_tab[single_index, , drop = FALSE], 1, which.max)
]

#cells with exactly 2 labels, one = Tumor
double_tumor_index <- rowSums(cur_tab) == 2 & cur_tab[, "Tumor"] == 1

#cells = Tumor + one other label, "-" the Tumor, keeps the other label
if (sum(double_tumor_index) > 0) {
  no_tumor <- cur_tab[, colnames(cur_tab) != "Tumor", drop = FALSE]

  cur_labels[double_tumor_index] <- colnames(no_tumor)[
    apply(no_tumor[double_tumor_index, , drop = FALSE], 1, which.max)
  ]
} else {
  message("No Tumor + other double labels found.")
}

#unresolved non-Tumor doublets
removed_doublet_ids <- names(cur_labels)[cur_labels == "doublets"]

#"-" anything that still calls doublets
if (length(removed_doublet_ids) > 0) {
  cur_labels <- cur_labels[cur_labels != "doublets"]
} else {
  message("No non-tumor doublets to remove.")
}

consolidation_stats <- data.frame(
  category = c(
    "total_unique_gated_cell_ids",
    "same_label_duplicates_collapsed",
    "single_label_kept",
    "tumor_plus_other_renamed_to_other",
    "non_tumor_doublets_removed"
  ),
  n = c(
    nrow(cur_tab),
    length(same_label_duplicate_ids),
    sum(single_index),
    sum(double_tumor_index),
    length(removed_doublet_ids)
  ),
  row.names = NULL
)

consolidation_stats
```

```{r}
# NOTE: підрахунок клітин по кожній консолідованій мітці.
table(cur_labels)
```


```{r}
# NOTE: записуємо фінальні мітки у colData(spe_27)$cell_labels; клітини без гейту = "unlabelled".
sce_labels <- rep("unlabelled", ncol(spe_27))
names(sce_labels) <- colnames(spe_27)

matched_ids <- intersect(names(cur_labels), colnames(spe_27))
unmatched_ids <- setdiff(names(cur_labels), colnames(spe_27))

message("Matched gated cells: ", length(matched_ids))
message("Unmatched gated cells: ", length(unmatched_ids))

#replaces "unlabelled" with the final labels for matching gated cells
sce_labels[matched_ids] <- cur_labels[matched_ids]

#label vector still has exactly one entry per spe_27 column, in the same order
stopifnot(length(sce_labels) == ncol(spe_27))
stopifnot(identical(names(sce_labels), colnames(spe_27)))

# final labels in colData(sce)
spe_27$cell_labels <- sce_labels

table(spe_27$cell_labels)

```

## visualization

```{r}
# NOTE: задаємо фіксовану палітру кольорів для кожного типу клітин і кладемо її в metadata(spe_27)
# для подальших візуалізацій.
cell_types <- c(
  B = "red3",
  BnT = "indianred4",
  CD4 = "palevioletred1",
  CD8 = "darkorchid",
  Treg = "maroon4",
  NK = "yellow",
  Igg = "darkred",

  DC = "green4",
  Mac = "darkolivegreen",
  MacCD209 = "olivedrab1",
  MacCD204 = "olivedrab3",
  mregDC = "green",

  Neutrophil = "blue1",
  PMN_MDSC = "deepskyblue",
  MDSC = "aquamarine4",

  Vasculature = "brown2",
  vCAF = "tomato",
  Tumor = "sienna4",
  SMA = "goldenrod2",
  Fibro = "chocolate1",

  unlabelled = "gray",
  undefined = "black"
)

metadata(spe_27)$colour_vectors$cell_types <- cell_types
```

## QC

```{r}
# NOTE: QC по мітках — беремо лише гейтовані клітини, рахуємо середню експресію
# по (sample_id, cell_labels) і готуємо маркери (без DNA/Histone) для теплової карти.
lab_sce <- spe_27[, !is.na(spe_27$cell_labels) & spe_27$cell_labels != "unlabelled"]

# Aggregate mean expression per sample and cell label
agr_sce <- scater::aggregateAcrossCells(
  lab_sce,
  ids = colData(lab_sce)[, c("sample_id", "cell_labels")],
  statistics = "mean"
)

assay(agr_sce, "exprs") <- asinh(counts(agr_sce))

# Clean column names
colnames(agr_sce) <- paste0(agr_sce$sample_id, "_", agr_sce$cell_labels)

# Markers used for visualization
cur_markers <- rownames(spe_27)[
  !grepl("DNA|Histone", rownames(spe_27), ignore.case = TRUE)
]

# Keep colors only for labels present in aggregated object
cell_type_colors <- metadata(spe_27)$colour_vectors$cell_types
cell_type_colors <- cell_type_colors[
  names(cell_type_colors) %in% unique(agr_sce$cell_labels)
]
```


```{r define-color-vector, fig.height=12, fig.width=10}
# NOTE: теплова карта середньої експресії маркерів, згрупована за типами клітин (перевірка,
# що мітки біологічно осмислені).
dittoHeatmap(
  agr_sce[cur_markers, ],
  assay = "exprs",
  annot.by = c("sample_id", "cell_labels"),
  cluster_rows = FALSE,
  scale = "none",
  heatmap.colors = viridis::viridis(100),
  order.by = "cell_labels",
  annotation_colors = list(
    cell_labels = cell_type_colors
  )
)
```

# train classifier

## classifier rules

```{r}
# NOTE: біологічні правила очистки перед навчанням: викидаємо Igg з IgG<1 та NK з CD7<0.5,
# а mregDC з високим епітеліальним сигналом (panCK>0.5 або Ecad>0.8) перемічаємо на Tumor.
lab_sce <- spe_27[, !is.na(spe_27$cell_labels) & spe_27$cell_labels != "unlabelled"]
unlab_sce <- spe_27[, is.na(spe_27$cell_labels) | spe_27$cell_labels == "unlabelled"]

# - Igg < 1
exprs_mat <- assay(lab_sce, "exprs")
low_igg <- lab_sce$cell_labels == "Igg" & exprs_mat["IgG", ] < 1
lab_sce <- lab_sce[, !low_igg]

# - NKs withCD7 expression < 0.5
exprs_mat <- assay(lab_sce, "exprs")
low_cd7_nk <- lab_sce$cell_labels == "NK" & exprs_mat["CD7", ] < 0.5
lab_sce <- lab_sce[, !low_cd7_nk]

# Relabel mregDC cells with high epithelial/tumor marker expression to Tumor
exprs_mat <- assay(lab_sce, "exprs")
mregdc_to_tumor <- lab_sce$cell_labels == "mregDC" &
  (
    exprs_mat["panCK", ] > 0.5 |
      exprs_mat["Ecad", ] > 0.8
  )

lab_sce$cell_labels[mregdc_to_tumor] <- "Tumor"

table(lab_sce$cell_labels)
```
## save lab_sce

```{r}
# NOTE: таблиця кількості клітин кожного типу по кожному знімку (вхід для оптимізатора train/test поділу).
dat <- as.data.frame(unclass(table(lab_sce$sample_id,lab_sce$cell_labels)))
dat
```


```{r}
# NOTE: скільки клітин загалом labelled vs unlabelled.
table(
  label_status = ifelse(
    spe_27$cell_labels == "unlabelled",
    "unlabelled",
    "labelled"
  )
)
```

## find a good test/train split

## split data into test & train sets

```{r}
# NOTE: визначаємо функцію split_test_rows() — оптимізатор train/test поділу НА РІВНІ ЗНІМКІВ,
# який цілиться у задані частки тест-набору окремо для частих і рідкісних типів клітин.
# Stratified train/test row split for multi-label count data.
#
# Each row is a sample with non-negative counts of each cell type in the
# columns. The split is at the row level, but it targets *cell-level* totals:
# achieved_test_frac[l] := test_count[l] / total[l] should be close to
# `default_test_frac` for common labels and `rare_test_frac` for the labels
# in `rare_labels` (default: "MDSC"). Additional labels whose total count is
# below `rare_threshold` are treated as rare too, if that argument is given.
#
# Objective (minimized):
#   sum_l  w_l * (test_count_l / total_l - target_frac_l)^2
# with w_l = `rare_weight` for rare labels, 1 otherwise.
#
# Algorithm: random init (biased toward test for rows containing rare labels),
# then steepest-descent single-row flips evaluated for all rows in one
# vectorized matrix multiply per step. Repeated from `n_restarts` initial
# seeds; the best basin wins. Converges in milliseconds for a few hundred
# rows; scales as O(restarts * flips * n * L).

split_test_rows <- function(counts,
                            default_test_frac = 0.20,
                            rare_test_frac    = 0.40,
                            rare_labels       = c("MDSC"),
                            rare_threshold    = NULL,
                            rare_weight       = 1,
                            n_restarts        = 100,
                            seed              = 42) {
  set.seed(seed)

  if (is.data.frame(counts)) {
    if (!is.numeric(counts[[1]])) {
      rn     <- as.character(counts[[1]])
      counts <- as.matrix(counts[, -1, drop = FALSE])
      rownames(counts) <- rn
    } else {
      counts <- as.matrix(counts)
    }
  }
  storage.mode(counts) <- "double"

  labels <- colnames(counts)
  totals <- colSums(counts)
  n      <- nrow(counts)

  if (!is.null(rare_threshold))
    rare_labels <- union(rare_labels, labels[totals < rare_threshold])
  rare_labels <- intersect(rare_labels, labels)
  is_rare     <- labels %in% rare_labels

  target_frac        <- ifelse(is_rare, rare_test_frac, default_test_frac)
  names(target_frac) <- labels

  # Zero out empty columns so they don't contribute a constant target_frac^2
  # penalty that no flip can fix.
  active     <- totals > 0
  w          <- ifelse(is_rare, rare_weight, 1) * active
  inv_totals <- ifelse(active, 1 / pmax(totals, 1), 0)

  # rate[i,l] = counts[i,l] / total[l]; precomputed once.
  rate      <- counts * rep(inv_totals, each = n)
  rate_sq_w <- as.numeric(rate^2 %*% w)   # length n, constant across iterations

  # Bias rows containing rare-label cells toward starting in test.
  has_rare  <- if (any(is_rare)) rowSums(counts[, is_rare, drop = FALSE]) > 0
               else rep(FALSE, n)
  init_prob <- ifelse(has_rare, rare_test_frac, default_test_frac)

  best_score   <- Inf
  best_in_test <- logical(n)

  for (restart in seq_len(n_restarts)) {
    in_test  <- runif(n) < init_prob
    test_sum <- colSums(counts[in_test, , drop = FALSE])

    repeat {
      cur_err   <- test_sum * inv_totals - target_frac
      cur_score <- sum(w * cur_err * cur_err)

      sgn        <- 1 - 2 * in_test                            # +1 = will add to test, -1 = will leave test
      linear     <- as.numeric(rate %*% (w * cur_err))
      flip_delta <- 2 * sgn * linear + rate_sq_w               # change in objective for each candidate flip

      best_i <- which.min(flip_delta)
      if (flip_delta[best_i] >= -1e-12) break                  # local minimum reached

      test_sum        <- test_sum + sgn[best_i] * counts[best_i, ]
      in_test[best_i] <- !in_test[best_i]
    }

    if (cur_score < best_score) {
      best_score   <- cur_score
      best_in_test <- in_test
    }
  }

  test_sum <- colSums(counts[best_in_test, , drop = FALSE])
  achieved <- ifelse(active, test_sum * inv_totals, NA_real_)

  summary_df <- data.frame(
    label              = labels,
    total              = totals,
    test_count         = test_sum,
    train_count        = totals - test_sum,
    target_test_frac   = target_frac,
    achieved_test_frac = round(achieved, 3),
    is_rare            = is_rare,
    row.names          = NULL
  )

  list(
    test_idx       = which(best_in_test),
    train_idx      = which(!best_in_test),
    test_rownames  = rownames(counts)[best_in_test],
    train_rownames = rownames(counts)[!best_in_test],
    summary        = summary_df,
    score          = best_score
  )
}
```


## looking for a seed with the best distribution between train/test

```{r}
# NOTE: запускаємо оптимізатор поділу з підвищеною вагою (rare_weight=20) для рідкісних типів
# (B, BnT, MDSC, Igg, mregDC, NK), 1000 рестартів; res$summary показує досягнуті частки тесту.
res <- split_test_rows(
  dat,
  default_test_frac = 0.20,
  rare_test_frac = 0.20,
  rare_labels = c("B", "BnT", "MDSC", "Igg", "mregDC", "NK"),
  rare_weight = 20,
  n_restarts = 1000,
  seed = 42
)

res$summary
```


```{r}
# NOTE: друкуємо підсумок поділу (скільки знімків у тесті, значення цільової функції).
# УВАГА: цей чанк використовує df і dt, яких не визначено вище — впаде, якщо їх не задати.
cat(sprintf("Test rows: %d / %d (%.1f%%) | objective = %.5f | %.3fs\n",
            length(res$test_idx), nrow(df),
            100 * length(res$test_idx) / nrow(df),
            res$score, dt))
print(res$summary)
```

```{r}
# NOTE: фактично ділимо клітини на train_sce/test_sce за списком знімків з оптимізатора.
test_sce <- lab_sce[, lab_sce$sample_id %in% res$test_rownames]
train_sce <- lab_sce[, lab_sce$sample_id %in% res$train_rownames]
```


```{r}
# NOTE: зводимо таблицю кількості клітин кожного типу окремо у train і test (перевірка балансу поділу).
all_labels <- sort(unique(c(
  as.character(train_sce$cell_labels),
  as.character(test_sce$cell_labels)
)))

all_labels <- all_labels[
  !is.na(all_labels) &
    all_labels != "unlabelled"
]

train_tab <- table(
  factor(as.character(train_sce$cell_labels), levels = all_labels)
)

test_tab <- table(
  factor(as.character(test_sce$cell_labels), levels = all_labels)
)

split_stats <- data.frame(
  cell_label = all_labels,
  train_n = as.integer(train_tab),
  test_n = as.integer(test_tab),
  row.names = NULL
)

split_stats
```

## rare cell types distribition between samples

```{r}
# NOTE: окремо перевіряємо баланс рідкісних типів між train/test — і кількість клітин, і кількість знімків.
rare_celltypes <- c("BnT", "Igg", "MDSC", "mregDC", "vCAF")

rare_balance <- bind_rows(
  as.data.frame(colData(train_sce)) |> 
    mutate(split = "train"),
  as.data.frame(colData(test_sce)) |> 
    mutate(split = "test")
) |> 
  dplyr::filter(cell_labels %in% rare_celltypes) |> 
  group_by(split, cell_labels) |> 
  summarise(
    total_cells = n(),
    n_samples = n_distinct(sample_id),
    .groups = "drop"
  ) |> 
  arrange(split, cell_labels)

rare_balance
```

## prepare data for classifier

1. split the labelled data into training + test data at the image level by `sample_id`. 
All cells from the same image are assigned training or test set

The split was generated with 5 grouped folds using `groupKFold`, 
and `Fold5` was used as the held-out test set. 

Within the training set another 5-fold grouped cross-validation for XGBoost parameter tuning was applied

```{r prepare-data, message = FALSE, eval=TRUE}
# =====================================================================
# STEP 05 — Vectorize (cell × marker matrix)
# =====================================================================
# NOTE: готуємо дані для XGBoost — будуємо матриці ознак X (клітини в рядках, маркери в стовпцях,
# без DNA/Histone, тому транспонуємо), цілі y (0-based), label_mapping (число<->тип),
# інверсно-частотні ваги класів (з пониженням undefined та Igg) і об'єкти xgb.DMatrix.

# Create unique cell identifiers so each cell can be traced back to its image
# and object number after training, prediction, or error analysis.
train_sce$cell_id <- paste0(train_sce$sample_id, "_", train_sce$ObjectNumber)
test_sce$cell_id <- paste0(test_sce$sample_id, "_", test_sce$ObjectNumber)

colnames(train_sce) <- train_sce$cell_id
colnames(test_sce) <- test_sce$cell_id

marker_keep <- !grepl("DNA|Histone", rownames(train_sce))

# XGBoost expects cells as rows and features as columns.
# The assay matrix has markers as rows and cells as columns, so we transpose it.
X <- t(assay(train_sce, "exprs")[marker_keep, ])
X_test_all <- t(assay(test_sce, "exprs")[marker_keep, ])

stopifnot(identical(colnames(X), colnames(X_test_all)))
stopifnot(!anyDuplicated(colnames(X)))

# XGBoost requires numeric class labels starting at 0.
# The factor levels from the training set define the fixed mapping between
# numeric class IDs and biological cell labels.
y_true <- factor(train_sce$cell_labels)
label_mapping <- levels(y_true)

# Convert training labels from factors to 0-based numeric labels.
y <- as.integer(y_true) - 1

# Number of classes required by the multiclass XGBoost objective.
num_classes <- length(label_mapping)

stopifnot(!anyNA(y))
stopifnot(min(y) == 0)
stopifnot(max(y) == num_classes - 1)

# Keep only labelled test cells for evaluation.
# Unlabelled cells are excluded from Xy_test because XGBoost cannot accept NA labels.
test_eval_keep <- !is.na(test_sce$cell_labels) &
  test_sce$cell_labels != "unlabelled" &
  test_sce$cell_labels %in% label_mapping

X_test_eval <- t(assay(test_sce, "exprs")[marker_keep, test_eval_keep])

y_test_true <- factor(
  test_sce$cell_labels[test_eval_keep],
  levels = label_mapping
)

y_test <- as.integer(y_test_true) - 1

stopifnot(!anyNA(y_test))
stopifnot(min(y_test) >= 0)
stopifnot(max(y_test) <= num_classes - 1)

# Inverse-frequency class weights so rare cell types contribute more
# to the loss function during training.
w_class <- 1 / (table(y_true) / length(y_true))

# Down-weight undefined and IgG classes so they do not dominate the classifier
# despite being noisy or less biologically informative labels.
if ("undefined" %in% names(w_class)) {
  w_class["undefined"] <- w_class["undefined"] / 10
}

if ("Igg" %in% names(w_class)) {
  w_class["Igg"] <- w_class["Igg"] / 30
}

# Convert class-level weights into one weight per training cell.
# Rescale weights to mean 1 for numerical stability.
w <- as.numeric(w_class[y_true])
w <- w / mean(w)

# Training data: features, labels, and cell weights.
Xy <- xgboost::xgb.DMatrix(
  data = X,
  label = y,
  weight = w
)

# Labelled held-out test data, for evaluation only.
Xy_test <- xgboost::xgb.DMatrix(
  data = X_test_eval,
  label = y_test
)

# All held-out/test cells, including unlabelled cells, for prediction/classification.
Xy_test_all <- xgboost::xgb.DMatrix(
  data = X_test_all
)

test_eval_cell_ids <- colnames(test_sce)[test_eval_keep]
test_all_cell_ids <- colnames(test_sce)

table(y_test_true)
```

## check the rules

Here it is clearly seen the upward rules for the specific markers work

# TEMP / QC plots
```{r}
# NOTE: QC-графіки, що правила очистки спрацювали — точкові діаграми для Igg (IgG vs DNA1)
# та mregDC (Ecad vs panCK) з пороговими лініями.
coldat <- as.data.frame(t(assay(train_sce, "exprs")))
coldat$label <- train_sce$cell_labels

ggplot(coldat |> dplyr::filter(label == "Igg"),
       aes(x = IgG, y = DNA1, col = label)) +
  geom_point() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  theme_bw()

ggplot(coldat |> dplyr::filter(label == "mregDC"),
       aes(x = Ecad, y = panCK, col = label)) +
  geom_point() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0.8, linetype = "dashed") +
  theme_bw()

```

## params for classofier, cross validation

All heavy calculations are here

```{r train-model-cv, message = FALSE, eval = TRUE}
# =====================================================================
# STEP 06 — Train (grouped XGBoost)
# =====================================================================
# NOTE: НАЙВАЖЧІ обчислення — grid-search крос-валідація XGBoost. Створюємо згруповані за
# знімками фолди (groupKFold, щоб клітини одного знімка не текли між train/valid), перебираємо
# сітку гіперпараметрів (12 комбінацій) через xgb.cv з early stopping і зберігаємо найкращий
# nrounds та mlogloss кожної моделі.

# Make sure cells from the same image are not split across CV train/test folds.
# groupKFold returns training indices, while xgb.cv expects validation indices,
# so we convert each fold to the held-out indices.
cv_train_index <- caret::groupKFold(
  group = factor(train_sce$sample_id),
  k = 5
)

all_indices <- seq_len(ncol(train_sce))

fold_list <- lapply(
  cv_train_index,
  function(train_idx) setdiff(all_indices, train_idx)
)

# XGBoost builds an ensemble of decision trees sequentially
# Each new tree tries to correct the errors made by the previous trees

param_grid <- expand.grid( #creates all combinations of the parameter values 3 * 2 * 1 * 1 * 2 = 12 models
  
  # Maximum depth of each tree.
  # Smaller values make simpler trees and reduce overfitting
  # Larger values allow more complex interactions between markers
  max_depth = c(3, 4, 5),

  # Learning rate controls how much each new tree contributes to the final model
  # < eta usually needs more trees but can generalize better
  eta = c(0.05, 0.1),

  # Fraction of cells sampled for each tree
  # Subsampling reduces overfitting by making trees less correlated
  subsample = c(0.8), #for each tree, XGBoost randomly uses 80% of the training cells and leaves out the other 20%

  # fraction of markers/features sampled for each tree, reduces overfitting and makes the model more robust.
  colsample_bytree = c(0.8), #40 * 0.8 = 32 markers

  # Minimum amount of weighted data required in a child node.
  # Higher values make the model more conservative, especially for rare classes
  min_child_weight = c(1, 5)
)

# Max N of boosting rounds, 1 round adds one tree per class for multiclass classification
# Early stopping below can stop before this limit if validation loss stops improving
nround <- 300

# Store the full CV models and a compact summary of their performance.
models <- list()
cv_summary <- list()

# Loop over all parameter combinations
for (i in seq_len(nrow(param_grid))) {
  
  # Parameters for the current XGBoost model
  params <- list(
    booster = "gbtree",

    # Multiclass classification with predicted probabilities for each cell type
    objective = "multi:softprob",

    # Optimize and monitor multiclass log-loss.
    #  <  mlogloss means predicted probabilities are closer to true labels.
    eval_metric = "mlogloss", #CD8 probability = 0.95 - low loss, good

    # Number of cell-type classes.
    num_class = num_classes,

    max_depth = param_grid$max_depth[i],
    eta = param_grid$eta[i],
    subsample = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i],
    min_child_weight = param_grid$min_child_weight[i],

    # Histogram-based tree construction, faster for larger datasets.
    tree_method = "hist",
    nthread = max(1, parallel::detectCores() - 1)
  )
  
  message(
    "Running: max_depth=", params$max_depth,
    "; eta=", params$eta,
    "; subsample=", params$subsample,
    "; colsample_bytree=", params$colsample_bytree,
    "; min_child_weight=", params$min_child_weight
  )
  
  # Cross-validation on image-level folds.
  # Each fold holds out complete images, so cells from the same image are not
  # split between training and validation.
  cv_results <- xgboost::xgb.cv(
    params = params,
    data = Xy,
    folds = fold_list, #@Round 1: train Fold2 + Fold3 + Fold4 + Fold5, test Fold1, etc.
    nrounds = nround, #300 ceiling, 107 - best

    # Stop if validation log-loss does not improve for 20 rounds.
    # This helps prevent overfitting and saves computation time.
    early_stopping_rounds = 20, #can go more after the best (108), if not will keep 108

    verbose = F, #if T - there is a message about the ML process

    # During grid search we only need test loss, not cell-level predictions
    prediction = F
  )
  
  models[[i]] <- cv_results
  
  # Select the boosting round with the lowest validation log-loss.
  best_iteration <- which.min(cv_results$evaluation_log$test_mlogloss_mean)
  best_mlogloss <- cv_results$evaluation_log$test_mlogloss_mean[best_iteration]

  # Store parameter values and their best CV score.
  cv_summary[[i]] <- data.frame(
    param_grid[i, ],
    best_iteration = best_iteration,
    best_mlogloss = best_mlogloss
  )
}

# Combine all model summaries into one table for selecting the best parameters.
cv_summary <- bind_rows(cv_summary)
```


```{r show-cv-summary}
# NOTE: таблиця всіх комбінацій гіперпараметрів, відсортована за найкращим mlogloss (кращі зверху).
cv_summary |>
  dplyr::arrange(best_mlogloss)
```


## cross-validation on the best param

```{r select-best-model, message = FALSE, eval = TRUE}
# NOTE: обираємо комбінацію гіперпараметрів з найменшим CV-mlogloss і перезапускаємо CV лише для неї
# з prediction=TRUE, щоб отримати out-of-fold прогнози для оцінки точності по типах клітин.
# parameter combination with the lowest cross-validation log-loss
best_params <- cv_summary |> 
  arrange(best_mlogloss) |> 
  dplyr::slice(1)

best_params

# > best_params
#   max_depth eta subsample colsample_bytree
# 1         4 0.1       0.8              0.8
#   min_child_weight best_iteration best_mlogloss
# 1                5            108      0.292158

# Use the best number of boosting rounds found during CV.
# This is the round with the lowest validation log-loss before early stopping.
best_nrounds <- as.integer(best_params$best_iteration)

# Define the final set of XGBoost parameters selected by grid search.
params_best <- list(
  booster = "gbtree",

  # Multiclass classification with class probabilities as output.
  objective = "multi:softprob",

  # Evaluation metric used during tuning.
  eval_metric = "mlogloss",

  # Number of cell-label classes.
  num_class = num_classes,

  # Best tree and regularization parameters from CV.
  max_depth = as.integer(best_params$max_depth),
  eta = best_params$eta,
  subsample = best_params$subsample,
  colsample_bytree = best_params$colsample_bytree,
  min_child_weight = best_params$min_child_weight,

  # Faster tree construction for larger datasets.
  tree_method = "hist",
  nthread = max(1, parallel::detectCores() - 1)
)

# Rerun cross-validation only for the selected best model.
# Here prediction = TRUE stores out-of-fold predictions for each training cell.
# These predictions are used to estimate per-cell-type performance.
best_cv <- xgboost::xgb.cv(
  params = params_best,
  data = Xy,
  folds = fold_list,
  nrounds = best_nrounds,
  verbose = FALSE,
  prediction = TRUE
)

```


## Check CV accuracy per cell type for the best model

```{r cv-accuracy-per-celltype, fig.height=8, fig.width=12}
# NOTE: з out-of-fold ймовірностей беремо передбачений клас для кожної клітини (max.col),
# повертаємо назви типів через label_mapping і будуємо confusion matrix проти істинних міток.
# Extract out-of-fold predicted class probabilities from the best CV model.
# Rows are cells, columns are cell-label classes.
cv_prob <- best_cv$cv_predict

if (is.list(cv_prob)) {
  cv_prob <- do.call(rbind, cv_prob)
}

if (is.null(dim(cv_prob))) {
  cv_prob <- matrix(
    cv_prob,
    ncol = num_classes,
    byrow = TRUE
  )
}

cv_prob <- as.matrix(cv_prob)
storage.mode(cv_prob) <- "double"

# Check that the prediction matrix has one row per training cell
# and one column per cell-label class.
stopifnot(nrow(cv_prob) == length(y))
stopifnot(ncol(cv_prob) == num_classes)

# Convert predicted probabilities to predicted cell labels.
# max.col() finds the class with the highest probability for each cell.
# label_mapping converts numeric class positions back to cell-type names.
cv_pred <- factor(
  label_mapping[max.col(cv_prob)],
  levels = label_mapping
)

# Compare predicted labels against the true training labels.
# Because predictions are out-of-fold, each cell was predicted by a model
# that was not trained on that cell/image.
cm_cv <- caret::confusionMatrix(
  data = cv_pred,
  reference = y_true
)

# cm_cv$table
#              Reference
# Prediction        B   BnT   CD4   CD8    DC Fibro   Igg   Mac MacCD204 MacCD209
#   B             587     2     0     0     2     0     1     0        0        0
#   BnT             0   230     1     0     0     0     0     0        0        0
#   CD4             5    12  1796    21     2     0     3     0        0        2
#   CD8             0     1     9  2464     3     0     0     0        2        1
#   DC              1     0     0     0   346     0     0     0       11        2
#   Fibro           0     0     1     1     3   738    55     0        0        2
#   Igg             0     0     0     0     0    21    66     0        0        0
#   Mac             0     0     0     0     0     0     0   509       66       27
#   MacCD204        0     0     1     0    11     3     1    36     2711        1
#   MacCD209        0     1     2     0     1     1    13    50        1      328
#   MDSC            0     0     0     0     0     0     0     0        0        0
#   mregDC          0     0     1     0     0     1     0     1        2        2
#   Neutrophil      0     0     0     0     0     0     0     0        0        0
#   NK              0     0     1     0     1     4     0     0        0        0
#   PMN_MDSC        0     0     1     0     0     0     0     0        0        0
#   SMA             0     0     0     0     0    12     0     0        0        0
#   Treg            0     1     9     1     0     0     1     0        1        1
#   Tumor           0     0     1     0     0    80     0     0        0        0
#   undefined       0     0    20     1     0     2     1     0        0        0
#   Vasculature     0     0     0     1     0     0     0     0        0        0
#   vCAF            0     0     0     0     0     0     0     0        0        0
```

check where fibroblasts come from

```{r}
# NOTE: діагностика — звідки беруться помилки (напр. Tumor, передбачені як Fibro): дивимось,
# з яких знімків походять такі клітини.
train_sce$predicted <- cv_pred
table(train_sce$predicted,train_sce$cell_labels)
table(train_sce$cell_labels)
table(test_sce$cell_labels)
cur_sce <- train_sce[,train_sce$cell_labels == "Tumor" & train_sce$predicted == "Fibro"]
table(cur_sce$sample_id)
table(cur_sce$predicted,cur_sce$cell_labels)
```


```{r cv-accuracy-per-celltype, fig.height=8, fig.width=12}
# NOTE: загальні метрики CV-моделі (Accuracy, Kappa, no-information rate, p-value).
cm_cv$overall
# > cm_cv$overall
#       Accuracy          Kappa  AccuracyLower  AccuracyUpper   AccuracyNull AccuracyPValue  McnemarPValue 
#      0.9508750      0.8964330      0.9491264      0.9525804      0.7359367      0.0000000            NaN 

#95.1% of out-of-fold predictions are correct.
#Kappa corrects for agreement expected by chance/class imbalance, very good
# AccuracyNull the accuracy you would get by always predicting the most common class, probably Tumor

#AccuracyPValue = 0
#The model is significantly better than the null/majority-class classifier
#McnemarPValue = NaN: non-important
```



```{r cv-accuracy-per-celltype, fig.height=8, fig.width=12}
# NOTE: метрики окремо по кожному типу клітин — Sensitivity (TPR), Specificity, Pos Pred Value (precision).
cm_cv$byClass
# > cm_cv$byClass
#                    Sensitivity Specificity Pos Pred Value
# Class: B             0.9898820   0.9998836      0.9882155
# Class: BnT           0.9311741   0.9999669      0.9913793
# Class: CD4           0.9744981   0.9989474      0.9666308
# Class: CD8           0.9899558   0.9996567      0.9919485
# Class: DC            0.9376694   0.9995859      0.9326146

# sensitivity (true positive rate). Of all true cells of that class, how many were found?
# specificity (fpr) of all cells that are not that class, how many were correctly not predicted as that class?
#Pos Pred Value (precision). Of all cells predicted as that class, how many were truly that class?
#B, CD4, CD8 are very strong.
```


## FPR vs TPR on the best model param

```{r accuracy-per-cell-type, fig.height=8, fig.width=12}
# NOTE: графік TPR (Sensitivity) vs FPR (1-Specificity) по типах клітин; розмір точки = частота виявлення.
cv_perf <- data.frame(cm_cv$byClass) |>
  mutate(
    class = sub("Class: ", "", rownames(cm_cv$byClass)),
    fpr = 1 - Specificity
  )

ggplot(cv_perf) +
  geom_point(
    aes(
      x = fpr,
      y = Sensitivity,
      size = Detection.Rate,
      fill = class
    ),
    shape = 21,
    color = "black",
    alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    aes(
      x = fpr,
      y = Sensitivity,
      label = class
    ),
    size = 3,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  scale_fill_manual(
    values = metadata(spe_27)$colour_vectors$cell_types,
    guide = "none"
  ) +
  scale_size(
    range = c(1, 15),
    guide = guide_legend(title.position = "top")
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_text(hjust = 0.5)
  ) +
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")
```


## check best model 
extract evaluation dataframes

```{r extract-evaluation-dataframes}
# NOTE: збираємо evaluation_log усіх CV-моделей у один «довгий» data.frame (train/test mlogloss
# по ітераціях + параметри) для подальших таблиць і графіків.
#models <- readRDS(file.path(base_path, "ROut_rds_all", "models.rds"))

eval_data <- purrr::imap_dfr(models, function(model, model_id) {
  
  df <- model$evaluation_log
  
  best_iteration <- which.min(df$test_mlogloss_mean)
  best_mlogloss <- df$test_mlogloss_mean[best_iteration]
  
  df |> 
    mutate(
      model_id = model_id,
      max_depth = model$params$max_depth,
      eta = model$params$eta,
      subsample = model$params$subsample,
      colsample_bytree = model$params$colsample_bytree,
      min_child_weight = model$params$min_child_weight,
      best_iteration = best_iteration,
      best_mlogloss = best_mlogloss,
      is_best_iteration = iter == best_iteration
    ) |> 
    pivot_longer(
      cols = c(
        train_mlogloss_mean,
        train_mlogloss_std,
        test_mlogloss_mean,
        test_mlogloss_std
      ),
      names_to = c("Type", ".value"),
      names_pattern = "(train|test)_mlogloss_(mean|std)"
    ) |> 
    dplyr::rename(
      mlogloss_mean = mean,
      mlogloss_std = std
    ) |> 
    mutate(
      Type = stringr::str_to_title(Type),
      max_depth = as.factor(max_depth),
      eta = as.factor(eta),
      min_child_weight = as.factor(min_child_weight),
      params = paste(
        "depth", max_depth,
        "eta", eta,
        "sub", subsample,
        "col", colsample_bytree,
        "mcw", min_child_weight,
        sep = "_"
      )
    )
})

eval_data
```


```{r best-model-summary}
# NOTE: підсумкова таблиця моделей за тестовим mlogloss на їхній найкращій ітерації (кращі зверху).
best_model_summary <- eval_data |>
  dplyr::filter(Type == "Test", is_best_iteration) |>
  arrange(best_mlogloss) |>
  select(
    model_id,
    max_depth,
    eta,
    subsample,
    colsample_bytree,
    min_child_weight,
    best_iteration,
    best_mlogloss,
    params
  )

best_model_summary
```


## visualize model evaluation

```{r model-evaluation}
# NOTE: готуємо дані для теплової карти — мінімальний mlogloss по кожній комбінації гіперпараметрів
# окремо для Train і Test.
plot_heatmap_data <- eval_data |>
  dplyr::group_by(
    Type,
    eta,
    subsample,
    colsample_bytree,
    min_child_weight,
    max_depth
  ) |> 
  dplyr::summarise(
    mlogloss = min(mlogloss_mean),
    .groups = "drop"
  ) |> 
  dplyr::mutate(
    eta = as.factor(eta),
    subsample = as.factor(subsample),
    colsample_bytree = as.factor(colsample_bytree),
    min_child_weight = as.factor(min_child_weight),
    max_depth = as.factor(max_depth)
  )
```


```{r model-evaluation-2}
# NOTE: функція make_heatmap() будує теплову карту mlogloss (рядки=комбінації параметрів,
# стовпці=max_depth) з анотацією гіперпараметрів збоку; створюємо карти для Train і Test.
make_heatmap <- function(plot_data, cur_type, show_param_annotation = TRUE) {
  
  df <- plot_data |> 
    dplyr::filter(Type == cur_type) |> 
    arrange(
      eta,
      subsample,
      colsample_bytree,
      min_child_weight,
      max_depth
    ) |> 
    pivot_wider(
      names_from = max_depth,
      values_from = mlogloss
    )
  
  depth_cols <- sort(unique(as.character(plot_data$max_depth)))
  
  mat <- df |> 
    select(dplyr::all_of(depth_cols)) |> 
    as.matrix()
  
  cur_col_fun <- if (cur_type == "Train") {
  circlize::colorRamp2(
    c(
      min(mat, na.rm = TRUE),
      median(mat, na.rm = TRUE),
      max(mat, na.rm = TRUE)
    ),
    c("#2166ac", "white", "#b2182b")
  )
} else {
  circlize::colorRamp2(
    c(
      min(mat, na.rm = TRUE),
      median(mat, na.rm = TRUE),
      max(mat, na.rm = TRUE)
    ),
    c("darkgreen", "white", "red")
  )
}
  
  anno_colors <- list(
  "learning rate" = c(
    "0.05" = "#1b9e77",
    "0.1"  = "#d95f02"
  ),
  "cell subsampl" = c(
    "0.8" = "#7570b3"
  ),
  "marker subsam" = c(
    "0.8" = "#66a61e"
  ),
  "min child w" = c(
    "1" = "#e7298a",
    "5" = "#e6ab02"
  )
)
  
  row_anno <- if (show_param_annotation) {
    ComplexHeatmap::rowAnnotation(
      "learning rate" = df$eta,
      "cell subsampl" = df$subsample,
      "marker subsam" = df$colsample_bytree,
      "min child w" = df$min_child_weight,
      col = anno_colors
    )
  } else {
    NULL
  }
  
  ComplexHeatmap::Heatmap(
    mat,
    name = paste0(cur_type, "\nmlogloss"),
    column_title = paste0(cur_type, " multiclass log-loss"),
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    col = cur_col_fun,
    left_annotation = row_anno,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 10),
    show_row_names = FALSE,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid::grid.text(
        sprintf("%.3f", mat[i, j]),
        x,
        y,
        gp = grid::gpar(fontsize = 9)
      )
    }
  )
}

hm_train <- make_heatmap(plot_heatmap_data, "Train", show_param_annotation = TRUE)
hm_test  <- make_heatmap(plot_heatmap_data, "Test", show_param_annotation = FALSE)

```


```{r model-evaluation-heatmap, fig.height=7.5, fig.width=12}
# NOTE: показуємо теплові карти Train і Test поруч для порівняння (де перенавчання).
hm_train + hm_test
```


The best cross-validation performance was obtained with 
eta = 0.1, subsample = 0.8, colsample_bytree = 0.8, 
min_child_weight = 5, and max_depth = 3. 

The test mlogloss indicates that max_depth = 3 gives the best balance between model complexity and generalization, 
while deeper trees tend to reduce train loss but increase test loss, suggesting overfitting. 

The higher min_child_weight value performs better, 
indicating that a more conservative tree-splitting strategy is preferable for these data


```{r model-evaluation-curves, fig.height=10, fig.width=12}
# NOTE: визначаємо єдину найкращу модель (один рядок) за тестовим mlogloss.
best_model_summary <- eval_data |>
  dplyr::filter(Type == "Test", is_best_iteration) |>
  arrange(best_mlogloss) |>
  dplyr::slice(1)

best_model_summary
```


```{r model-evaluation-curves, fig.height=10, fig.width=12}
# NOTE: дістаємо назву параметрів, найкращу ітерацію і loss найкращої моделі для підпису на графіку.
best_param_name <- best_model_summary$params
best_iter <- best_model_summary$best_iteration
best_loss <- best_model_summary$best_mlogloss
```


```{r model-evaluation-curves, fig.height=10, fig.width=14}
# NOTE: криві mlogloss по ітераціях бустингу для всіх моделей (фасет Train/Test x eta),
# з позначкою обраної найкращої ітерації — видно перенавчання при глибших деревах.
ggplot(
  eval_data,
  aes(
    x = iter,
    y = mlogloss_mean,
    color = params,
    group = params
  )
) +
  geom_line(alpha = 0.75, linewidth = 0.6) +
  geom_vline(
    data = best_model_summary,
    aes(xintercept = best_iteration),
    inherit.aes = FALSE,
    linetype = "dashed",
    color = "black"
  ) +
  geom_point(
    data = best_model_summary,
    aes(
      x = best_iteration,
      y = best_mlogloss
    ),
    inherit.aes = FALSE,
    size = 3,
    color = "black"
  ) +
  facet_grid(rows = vars(Type), cols = vars(eta), scales = "free_y") +
  scale_color_brewer(palette = "Set3") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    legend.title = element_blank()
  ) +
  labs(
    title = "XGBoost Multiclass Log-Loss Across Boosting Iterations",
    subtitle = paste0(
      "Best CV model: ",
      best_param_name,
      "; iteration = ", best_iter,
      "; test mlogloss = ", round(best_loss, 3)
    ),
    x = "Boosting iteration",
    y = "Mean multiclass log-loss"
  )
```


Columns separate the learning rate values: eta = 0.05 on the left and eta = 0.1 on the right. 
In both training and test panels, log-loss decreases as more boosting iterations are added. 
Training loss continues to approach zero, especially for more flexible models, while test loss decreases initially and then plateaus. 
This gap between train and test loss reflects expected overfitting as boosting continues

The black dashed vertical line marks the selected best iteration (105)

The selected XGBoost model showed strong grouped cross-validation performance, with an overall accuracy of 95.1% and Cohen's kappa of 0.896
learly above the no-information rate of 73.6%. 
Most major cell types were classified with high sensitivity and precision, including B, BnT, CD4, CD8, DC, MacCD204, SMA, Treg, Tumor, undefined, and Vasculature

The main weaker classes were Fibro, Igg, Neutrophil, and PMN_MDSC. 
Fibro and PMN_MDSC showed high sensitivity but low precision, indicating that the model over-predicts these labels. 
Igg and Neutrophil showed low sensitivity, meaning that a substantial fraction of true cells from these classes are missed. 
These classes should therefore be interpreted with caution in downstream predictions.


# final classifier

```{r train-final-model, message = FALSE, eval = TRUE}
# NOTE: навчаємо ФІНАЛЬНУ модель xgb.train на ВСІХ train-клітинах з найкращими параметрами та nrounds
# (CV використовувалась лише для вибору параметрів). Саме ця модель іде на тест і прогнози.
# Train the final XGBoost model on all training cells using the best CV parameters.
# CV is only used to select parameters; this model is the one used for held-out
# test evaluation and later prediction on unlabelled cells.
final_model <- xgboost::xgb.train(
  params = params_best,
  data = Xy,
  nrounds = best_nrounds,
  verbose = TRUE
)
```

save final model via saveRDS

```{r train-final-model, message = FALSE, eval = TRUE}
# NOTE: (закоментовано) збереження фінальної моделі у .rds; розкоментуй, щоб зберегти на диск.
# saveRDS(
#   final_model,
#   file = file.path(base_path, "classifier_xgboost.rds")
# )
```



This uses:

max_depth = 3
eta = 0.1
subsample = 0.8
colsample_bytree = 0.8
min_child_weight = 5
nrounds = 108



```{r variable-importance, fig.height=8, fig.width=8}
# NOTE: рахуємо важливість маркерів (xgb.importance за метрикою Gain) і беремо топ-20 для графіка.
# final_model <- readRDS(
#   file.path(base_path, "classifier_xgboost.rds")
# )

# Extract marker importance from the trained XGBoost model.
# Gain reflects how much each marker contributed to reducing the loss.
#xgb.importance(model = final_model,feature_names =c("SMA"))

var_imp <- xgboost::xgb.importance(
  feature_names = colnames(X),
  model = final_model
)

var_imp |> 
  slice_max(Gain, n = 20)

var_imp_top <- var_imp |> 
  dplyr::slice_max(Gain, n = 20) |> 
  dplyr::mutate(
    Feature = forcats::fct_reorder(Feature, Gain)
  )
```


```{r variable-importance, fig.height=8, fig.width=8}
# NOTE: стовпчиковий графік топ-20 найважливіших маркерів за Gain.
ggplot(var_imp_top, aes(x = Gain, y = Feature)) +
  geom_col(fill = "steelblue") +
  theme_classic() +
  labs(
    title = "Top XGBoost Marker Importances",
    x = "Gain",
    y = "Marker"
  )
```



Variable importance was assessed using XGBoost gain, which measures how much each marker contributes to reducing the model loss across tree splits. 
The most informative markers were CD3, CD204, CD209, CD20, Lamp3, CD31, FOXP3, CD66b, and CD146. 
These markers correspond well to the expected biology of the annotated cell types, including T cells, B cells, macrophage/DC populations, vasculature, neutrophil-like cells, and stromal compartments.

This is a useful sanity check: the model is not relying on strange markers; it is using biologically meaningful features.

# prepare test data

```{r prepare-test-data}
# NOTE: готуємо held-out тестові дані — той самий набір маркерів, що і в навчанні, ті самі
# рівні label_mapping; будуємо X_test і Xy_test (xgb.DMatrix) лише для гейтованих тест-клітин.
test_sce$cell_id <- paste0(
  test_sce$sample_id,
  "_",
  test_sce$ObjectNumber
)

colnames(test_sce) <- test_sce$cell_id

# Labelled cells only for evaluation
test_eval_keep <- !is.na(test_sce$cell_labels) &
  test_sce$cell_labels != "unlabelled" &
  test_sce$cell_labels %in% label_mapping

# Use the same marker set as training
X_test <- t(assay(test_sce, "exprs")[marker_keep, test_eval_keep])

# Check that train and test feature columns match
stopifnot(identical(colnames(X), colnames(X_test)))

# Convert test labels using the training label mapping
y_test_true <- factor(
  test_sce$cell_labels[test_eval_keep],
  levels = label_mapping
)

y_test <- as.integer(y_test_true) - 1

stopifnot(!anyNA(y_test))

# Store labelled test data in XGBoost format
Xy_test <- xgboost::xgb.DMatrix(
  data = X_test,
  label = y_test
)

Xy_test
```


## visualize predicted results

```{r test-performance, fig.height=8, fig.width=12}
# NOTE: прогноз фінальної моделі на тесті; xgb віддає плаский вектор ймовірностей, тому пробуємо
# обидва способи reshape у матрицю (клітини x класи) і перевіряємо, де ймовірності сумуються в 1.
raw_test_pred <- predict(final_model, newdata = Xy_test)

# Try both possible reshaping directions
test_prob_byrow <- matrix(
  raw_test_pred,
  ncol = num_classes,
  byrow = TRUE
)

test_prob_bycol <- matrix(
  raw_test_pred,
  ncol = num_classes,
  byrow = FALSE
)

# Check which one has probabilities summing to 1 per cell
mean(abs(rowSums(test_prob_byrow) - 1))
mean(abs(rowSums(test_prob_bycol) - 1))
```


```{r test-performance, fig.height=8, fig.width=12}
# NOTE: обираємо правильно зорієнтовану матрицю ймовірностей, перевіряємо розміри та підписуємо
# рядки (cell_id) і стовпці (типи клітин з label_mapping).
if (
  mean(abs(rowSums(test_prob_byrow) - 1)) <
  mean(abs(rowSums(test_prob_bycol) - 1))
) {
  test_prob <- test_prob_byrow
} else {
  test_prob <- test_prob_bycol
}

stopifnot(nrow(test_prob) == sum(test_eval_keep))
stopifnot(nrow(test_prob) == nrow(X_test))
stopifnot(ncol(test_prob) == num_classes)
stopifnot(mean(abs(rowSums(test_prob) - 1)) < 1e-6)

rownames(test_prob) <- colnames(test_sce)[test_eval_keep]
colnames(test_prob) <- label_mapping
```

```{r}
# NOTE: з ймовірностей беремо передбачений клас (max.col), зіставляємо з істинними мітками тесту
# і будуємо confusion matrix на held-out наборі.
test_pred <- factor(
  label_mapping[max.col(test_prob, ties.method = "first")],
  levels = label_mapping
)

test_idx <- match(
  rownames(test_prob),
  colnames(test_sce)
)

stopifnot(!anyNA(test_idx))

test_true <- factor(
  test_sce$cell_labels[test_idx],
  levels = label_mapping
)

stopifnot(
  length(test_pred) == length(test_true),
  !anyNA(test_true)
)

cm_test <- caret::confusionMatrix(
  data = test_pred,
  reference = test_true
)
```


```{r}
# NOTE: confusion matrix на тесті (рядки=передбачено, стовпці=істина).
cm_test$table
```


```{r}
# NOTE: загальні метрики моделі на held-out тесті (Accuracy, Kappa тощо).
cm_test$overall
```



```{r test-performance, fig.height=8, fig.width=12}
# NOTE: фінальний графік продуктивності на тесті — TPR (Sensitivity) vs FPR (1-Specificity)
# по кожному типу клітин; розмір точки = частота виявлення.
# Per-cell-type performance plot.
test_perf <- data.frame(cm_test$byClass) |> 
  dplyr::mutate(
    class = sub("Class: ", "", rownames(cm_test$byClass)),
    fpr = 1 - Specificity
  )

ggplot(test_perf) +
  geom_point(
    aes(
      x = fpr,
      y = Sensitivity,
      size = Detection.Rate,
      fill = class
    ),
    shape = 21,
    color = "black",
    alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    aes(
      x = fpr,
      y = Sensitivity,
      label = class
    ),
    size = 3,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  scale_fill_manual(
    values = metadata(spe_27)$colour_vectors$cell_types,
    guide = "none"
  ) +
  scale_size(
    range = c(1, 15),
    guide = guide_legend(title.position = "top")
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_text(hjust = 0.5)
  ) +
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")
```








