---
title: "Random Forest & Consensus Gating Performance Report"
output:
  html_document:
    toc: true       
    toc_float: true      
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 7)

```

## 1. Summary

This report evaluates the classification performance of the **Random Forest (RF)** model, optimized via inverse-frequency class weights (`classwt`), and assesses its alignment with the baseline **XGBoost** model through a **Consensus Gating Pipeline**.

---

## 2. Model Evaluation Metrics

### 2.1 Per-Class Sensitivity and Specificity Deep Dive

This tabset breaks down how well the Random Forest model managed to distinguish each specific cell type, providing crucial insight into our rare cell optimization.

::: {.tabset .nav-pills}

#### Class Performance Table

```{r class-metrics-table}
library(knitr)
library(kableExtra)

rf_by_class <- as.data.frame(cm_rf$byClass[, c("Sensitivity", "Specificity", "Pos Pred Value", "Balanced Accuracy")])
rownames(rf_by_class) <- sub("Class: ", "", rownames(rf_by_class))


rf_by_class |> 
  kable(caption = "Random Forest Performance Statistics by Biological Class") |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

#### Sensitivity vs. Specificity Trade-off Plot

```{r rf-performance-scatter}
cv_perf_rf <- as.data.frame(cm_rf$byClass) |> 
  dplyr::mutate(
    class = sub("Class: ", "", rownames(cm_rf$byClass)),
    fpr = 1 - Specificity
  )

ggplot(cv_perf_rf, aes(x = fpr, y = Sensitivity, fill = class)) +
  geom_point(shape = 21, color = "black", alpha = 0.85, size = 4) + 
  ggrepel::geom_text_repel(aes(label = class), size = 3.5, max.overlaps = Inf) +
  scale_fill_manual(values = metadata(spe_27)$colour_vectors$cell_types, guide = "none") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Random Forest Classifier: Sensitivity vs. False Positive Rate",
    subtitle = "",
    x = "False Positive Rate (1 - Specificity)",
    y = "Sensitivity (True Positive Rate)"
  )

```

:::

---

## 3. Model Consensus & Disagreement Diagnostics

By running a cross-validation consensus matrix, we evaluate the structural agreement between **XGBoost** and **Random Forest**.

### 3.1 Consensus Metric

```{r consensus-percentage, echo=FALSE}
cat(sprintf("Overall Model Agreement Percentage = %.2f%%\n", agree_percent))

```

### 3.2 Matrix of Disagreements

The cross-tabulation below reveals exactly where the two models clash on border cases (e.g., `Fibro` vs. `Tumor`).

```{r disagreement-crosstab}
disagreement_matrix <- table(XGBoost = consensus_df$XGBoost, RandomForest = consensus_df$RandomForest)

disagreement_matrix |> 
  kable(caption = "Cross-Table of Classifier Decisions (XGBoost Rows vs. Random Forest Columns)") |> 
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 11)

```

---

## 4. Biomarker Expression Profiling (Conflict Plots)

To understand *why* models disagree, we plot the conflicting cell arrays across real metal-channel marker intensities.

::: {.tabset .nav-pills}

#### Tumor vs. Fibro (panCK / SMA)

```{r plot-tumor-fibro}
# Scatter plot panCK vs SMA for conflicting cells
ggplot(coldat_sub, aes(x = panCK, y = SMA, color = conflict)) +
  geom_point(alpha = 0.7, size = 1.8) +
  theme_bw(base_size = 12) +
  labs(
    title = "Epithelial-Mesenchymal Transition Boundaries: panCK vs. SMA",
    x = "panCK (Tumor Marker Expression)",
    y = "SMA (Fibroblast/Stromal Expression)",
    color = "Model Conflict Status"
  )

```

#### T-Cell Subsets (FOXP3 / CD3)

```{r plot-treg-cd4}
# Scatter plot FOXP3 vs CD3
ggplot(coldat_treg, aes(x = FOXP3, y = CD3, color = conflict)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "darkgrey") +
  theme_bw(base_size = 12) +
  labs(
    title = "Regulatory T-Cell Sorting: FOXP3 vs. CD3 Profiles",
    x = "FOXP3 Expression (Treg Selection)",
    y = "CD3 Expression (Pan-T Cell Lineage)",
    color = "Model Conflict Status"
  )

```

#### Lineage Sorting (CD20 / CD8a)

```{r plot-bnt-cd8}
ggplot(
  coldat |> filter(XGBoost %in% c("BnT", "CD8") | RandomForest %in% c("BnT", "CD8")),
  aes(x = CD20, y = CD8a, color = paste0("XGB:", XGBoost, " RF:", RandomForest))
) +
  geom_point(alpha = 0.7, size = 1.8) +
  theme_bw(base_size = 12) +
  labs(
    title = "B-Cell vs. Cytotoxic T-Cell Boundaries: CD20 vs. CD8a",
    x = "CD20 (B-Cell Marker Expression)",
    y = "CD8a (Cytotoxic T-Cell Expression)",
    color = "Classification Discrepancy"
  )

```

:::

---

## 5. Random Forest Feature Importance

The chart below shows which phenotypic markers contributed most heavily to the Random Forest splits (Mean Decrease in Accuracy/Gini index).

```{r rf-importance}
# Extract and plot variable importance from randomForest
rf_imp_df <- as.data.frame(importance(rf_model))
rf_imp_df$Marker <- rownames(rf_imp_df)
rf_imp_df <- rf_imp_df |> dplyr::arrange(desc(MeanDecreaseGini)) |> head(20)
rf_imp_df$Marker <- forcats::fct_reorder(rf_imp_df$Marker, rf_imp_df$MeanDecreaseGini)

ggplot(rf_imp_df, aes(x = MeanDecreaseGini, y = Marker)) +
  geom_col(fill = "#008080") +
  theme_classic(base_size = 12) +
  labs(
    title = "Top 20 Variable Importances (Gini Index)",
    x = "Mean Decrease in Gini Impurity",
    y = "Protein Target (Marker Name)"
  )

```


