---
title: "XGBoost rare-cell tuning — split-parameter report (v2, factorial)"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyr)
base_path <- "/Users/svitlanamoiseyenko/Repos/genomic/xgboost-imc-panel2-cell-classifier/XGBoost_classifier/Panel_2_10"
obj  <- readRDS(file.path(base_path, "ROut", "report_cms_split_v2.rds"))
res  <- obj$results
cm_store    <- obj$cm_store
split_store <- obj$split_store
split_grid  <- obj$split_grid
best_xgb    <- obj$best_xgb
best_s_id   <- obj$best_s_id

critical <- c("MDSC", "BnT", "Igg", "mregDC", "PMN_MDSC", "Neutrophil")
# stable display order: by index V1..V27
res <- res[order(suppressWarnings(as.integer(sub("V", "", res$model_id)))), ]
r3 <- function(x) round(as.numeric(x), 4)
# short factor handles for the three swept knobs
res$fracC <- res$tier_test_frac_critical
res$wC    <- res$tier_weight_critical
res$floor <- res$min_test_count_critical
```

# Summary

A **3×3×3 factorial** over the three deterministic split knobs, with the **model held fixed at M9**
(the guardrail-best XGBoost model: `max_depth=3, eta=0.05, min_child_weight=10, gamma=1, lambda=1,
max_delta_step=1, subsample=0.8, colsample_bytree=0.6`):

- critical **test-fraction** ∈ {0.20, 0.30, 0.40}
- critical **weight** ∈ {1, 40, 100}
- critical **floor** (`min_test_count`) ∈ {0, 8, 15}

= **27 split scenarios (V1–V27)**. Medium/non-critical fractions (0.25 / 0.20), weights (10 / 1) and
floors (5 / 0) are held fixed, and the seed is fixed at 42 — v1 (`02_split_params`) already showed
the optimiser is **seed-stable** (its three seed replicates landed on the same split; residual recall
spread ≈ 0.007 was model-training stochasticity). For each scenario the model is re-tuned for
`nrounds` via image-grouped `xgb.cv` and evaluated on **that split's own held-out images**.

> **Read this like v1, not the model report.** Every scenario has a *different* held-out test set, so
> a single high number can just mean "easier images landed in test." The factorial's payoff is **§5
> marginal effects**: averaging over the other two knobs cancels much of that noise and isolates what
> each knob actually does. Treat the noise floor as ≈ 0.007 (from v1) — gaps smaller than that are
> not real. We rank on **critical-class recall** (`AccuracyNull ≈ 0.84`, so raw accuracy hides rare
> cells).

# 1. The split knobs being varied (3 levels each)

- **`tier_test_frac` (critical: 0.20 / 0.30 / 0.40)** — share of each critical type's cells *targeted*
  for test. Higher → more rare cells in test (steadier recall estimate) but fewer left to train on.
- **`tier_weight` (critical: 1 / 40 / 100)** — how hard the optimiser hits the critical target. `1`
  removes tier priority (critical treated like any class); `100` forces it almost regardless of cost
  to other tiers.
- **`min_test_count` (critical: 0 / 8 / 15)** — hard floor of critical cells per type in test. `0` =
  no guarantee; `15` may be infeasible for the rarest types (Igg n=13, NK n=12) and pull whole images
  in to compensate.

# 2. The factorial grid

```{r grid-table}
knitr::kable(split_grid[, c("frac_crit", "w_crit", "min_crit", "seed", "note")],
             caption = "27-cell factorial (critical frac × weight × floor); model (M9) + cleaning held constant")
```

```{r fixed-model}
knitr::kable(as.data.frame(best_xgb),
             caption = "Fixed model across all 27 splits: M9 'conservative' XGBoost params")
```

# 3. Scenario results (sorted by critical-class recall)

Test cells/images vary per scenario (shown), so single-row gaps mix split strategy with test-set
composition — read §5 for the de-noised picture.

```{r scenario-table}
tab <- res |>
  dplyr::arrange(dplyr::desc(Sensitivity_critical_mean_test)) |>
  dplyr::transmute(
    model_id, fracC, wC, floor,
    Accuracy_test  = r3(Accuracy_test),
    Kappa_test     = r3(Kappa_test),
    Sens_crit_test = r3(Sensitivity_critical_mean_test),
    F1_crit_test   = r3(F1_critical_mean_test),
    n_test_cells, n_test_images, nrounds_best)
knitr::kable(tab, caption = "Held-out test metrics per split scenario (best = top)")
```

# 4. Per-class critical sensitivity (recall) matrix

```{r sens-matrix}
sens_cols <- paste0("Sens_", critical, "_test")
m <- res[, c("model_id", "fracC", "wC", "floor", sens_cols)]
colnames(m) <- c("id", "frac", "w", "floor", critical)
m[critical] <- lapply(m[critical], r3)
m <- m[order(-rowMeans(sapply(m[critical], as.numeric), na.rm = TRUE)), ]
knitr::kable(m, row.names = FALSE,
             caption = "Per-class test recall for the 6 critical cell types (higher = better)")
```

# 5. Marginal effects — what each knob actually does

Each table averages a metric over the **other two** knobs (9 scenarios per level). This is the part
the factorial is for: it cancels most of the test-set-composition noise and shows the knob's own
effect. Compare differences against the ≈ 0.007 noise floor.

```{r marginal-effects}
marg <- function(by) {
  res |>
    dplyr::group_by(.level = .data[[by]]) |>
    dplyr::summarise(
      n_scen        = dplyr::n(),
      Acc           = r3(mean(Accuracy_test)),
      Kappa         = r3(mean(Kappa_test)),
      Sens_crit     = r3(mean(Sensitivity_critical_mean_test)),
      F1_crit       = r3(mean(F1_critical_mean_test)),
      n_test_cells  = round(mean(n_test_cells)),
      n_test_images = round(mean(n_test_images), 1),
      .groups = "drop") |>
    dplyr::rename(!!by := .level)
}
knitr::kable(marg("fracC"), caption = "Marginal effect of critical TEST-FRACTION (averaged over weight × floor)")
knitr::kable(marg("wC"),    caption = "Marginal effect of critical WEIGHT (averaged over fraction × floor)")
knitr::kable(marg("floor"), caption = "Marginal effect of critical FLOOR (averaged over fraction × weight)")
```

Critical-recall as a **fraction × weight** grid (averaged over the floor levels) — look for whether
the fraction effect depends on the weight (interaction) or is roughly additive:

```{r interaction-grid}
inter <- res |>
  dplyr::group_by(fracC, wC) |>
  dplyr::summarise(Sens_crit = r3(mean(Sensitivity_critical_mean_test)), .groups = "drop") |>
  tidyr::pivot_wider(names_from = wC, values_from = Sens_crit, names_prefix = "w=")
knitr::kable(inter, caption = "Mean critical recall by critical fraction (rows) × weight (cols), over floor levels")
```

# 6. Per-iteration detail

```{r iterations, results='asis'}
analyse_row <- function(r) {
  sv <- as.numeric(r[paste0("Sens_", critical, "_test")]); names(sv) <- critical
  missed <- names(sv)[!is.na(sv) & sv < 0.5]
  msgs <- c()
  if (length(missed)) msgs <- c(msgs, paste0("missed (recall<0.5): ", paste(missed, collapse = ", ")))
  if (r$Accuracy_test - r$AccuracyNull_test < 0.05)
    msgs <- c(msgs, "accuracy barely above the majority-Tumor null — leaning on the dominant class")
  if (r$Precision_critical_mean_test < r$Sensitivity_critical_mean_test - 0.1)
    msgs <- c(msgs, "critical classes over-predicted (precision < recall)")
  if (!length(msgs)) msgs <- "no critical class below recall 0.5; balanced precision/recall"
  paste(msgs, collapse = "; ")
}

ord <- order(-res$Sensitivity_critical_mean_test)
for (i in ord) {
  r <- res[i, ]
  cat(sprintf("\n\n## %s — frac=%.2f, weight=%s, floor=%s\n\n",
              r$model_id, r$fracC, r$wC, r$floor))

  cat("**Split params (test-fraction targets / weights / floor)**\n\n")
  print(knitr::kable(data.frame(
    tier      = c("critical", "medium", "non_critical"),
    test_frac = c(r$tier_test_frac_critical, r$tier_test_frac_medium, r$tier_test_frac_noncrit),
    weight    = c(r$tier_weight_critical, r$tier_weight_medium, r$tier_weight_noncrit),
    min_test  = c(r$min_test_count_critical, 5, 0), row.names = NULL)))
  cat(sprintf("\n\nseed = %s | Train %d cells / Test %d cells across %d test images.\n\n",
              r$seed, r$n_train_cells, r$n_test_cells, r$n_test_images))

  sp <- split_store[[r$model_id]]
  if (!is.null(sp)) {
    cat("\n**Achieved split** (per-class counts and achieved test fraction)\n\n")
    print(knitr::kable(sp))
  }

  cat("\n\n**Test metrics**\n\n")
  print(knitr::kable(data.frame(
    Accuracy = r3(r$Accuracy_test), Kappa = r3(r$Kappa_test),
    Lower = r3(r$AccuracyLower_test), Upper = r3(r$AccuracyUpper_test),
    Null = r3(r$AccuracyNull_test), Sens_crit = r3(r$Sensitivity_critical_mean_test),
    F1_crit = r3(r$F1_critical_mean_test), nrounds_best = r$nrounds_best, row.names = NULL)))

  cm <- cm_store[[r$model_id]]
  if (!is.null(cm)) {
    cat("\n\n**Confusion matrix** (rows = Prediction, cols = Reference)\n\n")
    print(knitr::kable(as.data.frame.matrix(cm)))
  }
  cat("\n\n**Analysis:** ", analyse_row(r), "\n\n")
}
```

# 7. Conclusion — best split & which knob matters

```{r conclusion, results='asis'}
acc_floor <- 0.93; kappa_floor <- 0.85
elig <- res[res$Accuracy_test >= acc_floor & res$Kappa_test >= kappa_floor, ]
elig <- elig[order(-elig$Sensitivity_critical_mean_test, -elig$F1_critical_mean_test,
                   -elig$Kappa_test), ]
best <- res[res$model_id == best_s_id, ][1, ]

# marginal spreads (effect size of each knob on mean critical recall)
spread <- function(by) {
  a <- tapply(res$Sensitivity_critical_mean_test, res[[by]], mean)
  max(a) - min(a)
}
sp_frac <- spread("fracC"); sp_w <- spread("wC"); sp_floor <- spread("floor")
ranked_knob <- sort(c(`test-fraction` = sp_frac, weight = sp_w, floor = sp_floor), decreasing = TRUE)

cat(sprintf(
"**Guardrail:** Accuracy_test ≥ %.2f and Kappa_test ≥ %.2f → %d/%d scenarios eligible.\n\n",
acc_floor, kappa_floor, nrow(elig), nrow(res)))

cat(sprintf(
"**Selected split: %s** (critical frac = %.2f, weight = %s, floor = %s)\n\n- Held-out test: Accuracy = %.3f, Kappa = %.3f (null = %.3f)\n- Critical-class recall = %.3f, critical F1 = %.3f\n- Test set: %d cells / %d images; model re-tuned to nrounds = %s\n\n",
best$model_id, best$fracC, best$wC, best$floor,
best$Accuracy_test, best$Kappa_test, best$AccuracyNull_test,
best$Sensitivity_critical_mean_test, best$F1_critical_mean_test,
best$n_test_cells, best$n_test_images, best$nrounds_best))

cat("**Knob influence on mean critical recall (marginal spread, vs ≈0.007 noise floor):**\n\n")
for (nm in names(ranked_knob))
  cat(sprintf("- %s: %.3f%s\n", nm, ranked_knob[[nm]],
              if (ranked_knob[[nm]] < 0.01) "  (within noise — negligible)" else ""))
```

**Takeaways**

- The model is identical in all 27 rows — this measures the **evaluation split**, not the classifier.
  The "best split" is the most trustworthy rare-cell estimate, not a better model.
- Trust **§5 marginals over §3 single rows**: any knob whose marginal spread is under ≈0.007 (the v1
  noise floor) is doing nothing measurable to rare-class recall — it's just moving which images test.
- The **floor** mainly governs *feasibility*: watch the achieved-split tables (§6) for a critical
  class left under its floor (recall then unestimable) or a high floor (15) dragging extra images into
  test and shrinking the training set.
- A high **test-fraction** can inflate test recall while starving training of rare cells — confirm any
  apparent win against the production split (v1 S1: frac 0.30 / weight 40 / floor 8).
```
