---
title: "Label-permutation test — 0.994 classifier"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 4)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r load}
# Outputs produced by permutation_build.R (run that Rscript first).
base_path <- "/Users/svitlanamoiseyenko/Repos/genomic/xgboost-imc-panel2-cell-classifier/XGBoost_classifier/Panel_2_10"
rds_path  <- file.path(base_path, "ROut", "permutation_null.rds")
stopifnot(file.exists(rds_path))

perm <- readRDS(rds_path)
null_df  <- perm$null
observed <- perm$observed
pvalues  <- perm$pvalues
cfg      <- perm$config
```

## What this is

The held-out **test accuracy (~0.994)** is inflated by Tumor dominance: the no-information rate
(NIR) is ~0.87, so a "predict Tumor for everything" model already scores ~0.87. To check whether the
classifier learns **real marker → cell-type structure** (and not just the class imbalance), we
shuffled the `cell_labels`, refit the *same* estimator, and repeated it
**`r cfg$N_PERM`** times (`null_type = "`r cfg$null_type`"`) to build a **null distribution** for
each metric. Everything except the labels was held fixed: the train/test image split, the CV folds,
and the tuned params (`max_depth=`r cfg$params_best$max_depth`, eta=`r cfg$params_best$eta`,
min_child_weight=`r cfg$params_best$min_child_weight`, nrounds=`r cfg$best_nrounds`).

`r cfg$n_train_cells` train cells / `r cfg$n_test_cells` test cells, `r cfg$num_classes` classes.

## p-values

One-sided empirical p-value `p = (1 + #{null >= observed}) / (N + 1)`. Smallest achievable here is
`r round(1/(cfg$N_PERM + 1), 4)`.

```{r pval-table}
knitr::kable(pvalues, digits = 4)
```

## Null distributions vs observed

Read these top-to-bottom: **accuracy's null sits near the NIR (~0.87)** — that floor is *why raw
accuracy is misleading* — while **Kappa and macro-F1 nulls sit near 0**. The observed values
(dashed red line) landing far in the upper tail is the actual proof the model learned biology.

```{r hist-fn}
plot_null <- function(metric, title) {
  if (!metric %in% names(null_df)) return(invisible(NULL))
  obs <- observed[[metric]]
  ggplot(data.frame(value = null_df[[metric]]), aes(value)) +
    geom_histogram(bins = 30, fill = "grey75", color = "white") +
    geom_vline(xintercept = obs, color = "red3", linetype = "dashed", linewidth = 1) +
    annotate("text", x = obs, y = Inf, label = sprintf("observed = %.3f", obs),
             vjust = 1.5, hjust = 1.05, color = "red3", size = 3.5) +
    labs(title = title, x = metric, y = "permutations") +
    theme_bw()
}
```

### Held-out test

```{r hist-test, fig.height=3.4}
plot_null("Accuracy_test", "Test accuracy — null vs observed (null ~ NIR)")
plot_null("Kappa_test",    "Test Kappa — null vs observed (null ~ 0)")
plot_null("F1_macro_test", "Test macro-F1 — null vs observed (null ~ 0)")
```

### Cross-validation (out-of-fold)

```{r hist-cv, fig.height=3.4, eval=isTRUE(cfg$DO_CV)}
plot_null("Accuracy_cv", "CV accuracy — null vs observed")
plot_null("Kappa_cv",    "CV Kappa — null vs observed")
plot_null("F1_macro_cv", "CV macro-F1 — null vs observed")
```

## Sanity checks

```{r sanity}
sane <- data.frame(
  check = c(
    "null test accuracy ~ NIR (good)",
    "null test Kappa ~ 0 (good; far from 0 => leak)",
    "null test macro-F1 ~ 0 (good)"
  ),
  null_mean = c(
    round(mean(null_df$Accuracy_test, na.rm = TRUE), 3),
    round(mean(null_df$Kappa_test,    na.rm = TRUE), 3),
    round(mean(null_df$F1_macro_test, na.rm = TRUE), 3)
  ),
  reference = c(round(observed$AccuracyNull_test, 3), 0, 0)
)
knitr::kable(sane)
```

## Conclusion

```{r conclusion, results='asis'}
acc_p   <- pvalues$p_value[pvalues$metric == "Accuracy_test"]
kappa_p <- pvalues$p_value[pvalues$metric == "Kappa_test"]
f1_p    <- pvalues$p_value[pvalues$metric == "F1_macro_test"]
cat(sprintf(
"- Observed test accuracy %.3f vs null mean %.3f (p = %.4f).\n",
  observed$Accuracy_test, mean(null_df$Accuracy_test, na.rm = TRUE), acc_p))
cat(sprintf(
"- Observed test Kappa %.3f vs null mean %.3f (p = %.4f).\n",
  observed$Kappa_test, mean(null_df$Kappa_test, na.rm = TRUE), kappa_p))
cat(sprintf(
"- Observed test macro-F1 %.3f vs null mean %.3f (p = %.4f).\n",
  observed$F1_macro_test, mean(null_df$F1_macro_test, na.rm = TRUE), f1_p))
cat("\nThe model's performance is **not** explained by the Tumor imbalance: under shuffled labels ",
    "the best it can do is predict Tumor (accuracy ~ NIR, Kappa/macro-F1 ~ 0). Kappa and macro-F1, ",
    "not raw accuracy, are the metrics that demonstrate the real learned signal.\n", sep = "")
```
